PRRT2 as an auxiliary regulator of Nav channel slow inactivation

Version 1

Abstract

During sustained activity, voltage-gated sodium (Nav) channels enter a slow-inactivated state to limit cellular hyperexcitability. Disruption of this regulatory process has been implicated in skeletal, cardiac and neurological disorders. While the kinetics of this process are well characterized, its endogenous modulators remain unclear. Here, we identify Proline-Rich Transmembrane Protein 2 (PRRT2) as a native regulator of Nav channel slow inactivation. We show that PRRT2 facilitates the entry of Nav channels into slow-inactivated state and delays their recovery, a regulatory effect conserved from zebrafish to humans. PRRT2 forms molecular complexes with Nav channels both in vitro and in vivo. In the mouse cortex, PRRT2 deficiency impairs the slow inactivation of Nav channels in neuronal axons, leading to reduced cortical resilience in response to hyperexcitable challenges. Together, these findings establish PRRT2 as a physiological modulator of Nav channel slow inactivation and reveal a mechanism that supports cortical resilience to pathological perturbations.

Introduction

Voltage-gated sodium (Nav) channels are critical for the initiation and propagation of electrical signals in excitable cells, contributing to cellular excitability in various tissues(Hodgkin & Huxley, 1952; Ulbricht, 1977). In mammals, including humans, Nav channels consist of nine = subunit isoforms (Nav1.1-Nav1.9), which are differentially expressed and distributed across different tissues to regulate specific physiological functions(Catterall, Goldin, & Waxman, 2005). Mutations in Nav channels that lead to either loss- or gain-of-function of these chan-nels are associated with a wide spectrum of diseases, including neurological disorders(Escayg & Goldin, 2010; Hedrich, Lauxmann, & Lerche, 2019; Meisler, Hill, & Yu, 2021), pain syndromes(Dib-Hajj, Geha, & Waxman, 2017; Waxman, 2013), muscle illnesses(Cannon, 2018; Mantegazza, Cestele, & Catterall, 2021) and cardiac diseases(Remme, 2013). Consequently, the Nav channels are considered promising therapeutic targets for treating these conditions(Noreng, Li, & Payandeh, 2021; Wisedchaisri & Gamal El-Din, 2022).

Voltage-dependent conformational changes enable Nav channels to transition between resting, open and inactivated states(Catterall, Wisedchaisri, & Zheng, 2020). Eukaryotic Nav channels undergo two major forms of inactivation (fast and slow inactivation), which are distinguished by their kinetics of onset and recovery(Goldin, 2003). Fast inactivation is triggered within milliseconds by brief depolarization and is rapidly reversed upon hyperpolarization, typically within tens of milliseconds (Armstrong & Bezanilla, 1977; Bezanilla & Armstrong, 1977). In contrast, slow inactivation develops during prolonged depolarizations and the slow inactivated channels require substantially longer periods to recover, ranging from hundreds of milliseconds to minutes(Rudy, 1978). Previous studies have demonstrated that slow inactivation regulates Nav channel availability during sustained or repetitive depolarization, thus modulating the cellular excitability(Goldin, 2003; Silva, 2014; Vilin & Ruben, 2001). Impairment of this regulatory process due to the mutations in Nav channels has been associated with disorders such as periodic paralysis(Hayward, Sandoval, & Cannon, 1999), cardiac arrhythmias(Richmond, Featherstone, Hartmann, & Ruben, 1998; Vilin, Makita, George, & Ruben, 1999) and epilepsy(Ghovanloo et al., 2023).

Beyond the intrinsic hierarchical gating mechanisms, auxiliary regulators further modulate Nav channel inactivation dynamics(Goldin, 2003). For example, β subunits of sodium channels (SCN1B) regulate inactivation of Nav channels in the heterologous expression system, although their effects vary across different Nav isoforms and cell types(He & Soderlund, 2014; Vilin et al., 1999; Webb, Wu, & Cannon, 2009). Fibroblast growth factor homologous factors (FHFs) interact with the carboxy terminus of Nav channels(Gade, Marx, & Pitt, 2020; Goetz et al., 2009), promoting intermediated or long-term inactivation, with recovery times spanning hundreds of milliseconds(Dover, Solinas, D’Angelo, & Goldfarb, 2010; Venkatesan, Liu, & Goldfarb, 2014). Although these auxiliary proteins have been identified as regulators of fast or intermediate inactivation, native modulators that specifically targeting slow inactivation of Nav channels remain largely unknown.

Recent studies on proline-rich transmembrane protein 2 (PRRT2) provide new insights into this issue. The PRRT2, a member of the Dispanins B subfamily (DspB)(Sallman Almen, Bringeland, Fredriksson, & Schioth, 2012), has been identified as a causative gene for paroxysmal kinesigenic dyskinesia disease(W. J. Chen et al., 2011; Lee et al., 2012; J. L. Wang et al., 2011). In vitro studies have shown that PRRT2 regulates Nav channel surface expression and delays recovery of these channels from inactivated states induced by brief depolarization(Fruscione et al., 2018; Lu et al., 2021; Valente et al., 2023). However, these studies did not clearly distinguish between the effects of PRRT2 on fast versus slow inactivation, nor did they provide direct in vivo evidence for its role in regulating Nav channel inactivation. Our original observations that PRRT2 expression increased the accumulation of inactivated Nav channels during repetitive stimulation(Lu et al., 2021) suggest that PRRT2 may be a previously unrecognized auxiliary regulator of Nav channel slow inactivation. In this study, we combined whole-cell voltage-clamp recordings, co-immunoprecipitation, and in vivo approaches to investigate the role of PRRT2 in regulating Nav channel slow inactivation, and to examine how deficits in PRRT2-mediated modulation of this process affects neuronal function in awake mice.

Results

PRRT2 regulates the slow inactivation of Nav1.2 channels

Fast and slow inactivation of Nav channels proceed through distinct gating pathways(Goldin, 2003). Although previous studies have implicated PRRT2 in the regulation of Nav channel inactivation(Fruscione et al., 2018; Lu et al., 2021), it remains unclear whether PRRT2 primarily influences fast inactivation, slow inactivation or both. To address this, we express mouse PRRT2 in HEK293T cells stably expressing Nav1.2 (Figure 1A) and applied classic electrophysiological paradigms to distinguish between the two inactivation processes.

PRRT2 regulates slow inactivation of Nav1.2 channels.

(A) Representative immunoblots showing PRRT2 protein expression in Nav1.2-stably expressing HEK293T cells transfected with constructs encoding either EGFP or mouse PRRT2 (mPRRT2). MW, molecular weight. (B) Schematic of whole-cell recording (left) and representative traces of sodium currents evoked by a brief depolarization (right). The arrow indicate sodium current decay due to fast inactivation. Scale bar, 1 nA by 2 ms. (C) Decay time constants of sodium currents in response to 20-ms depolarization (EGFP: n = 13 cells, mPRRT2: n = 13 cells). (D) Recovery from fast inactivation of Nav1.2 channels (EGFP: n = 10 cells, mPRRT2: n = 7 cells). Insert: Protocol for examining fast inactivation recovery. (E) Schematic diagram of entry into and recover from slow inactivation of voltage-gated sodium channels. Slow inactivation modulates the availability of the sodium channels. (F) Entry into slow-inactivated state of Nav1.2 channels induced by progressive longer depolarization (EGFP: n = 6 cells, mPRRT2: n = 6 cells). (G) The effects of PRRT2 on the recovery of Nav1.2 channels from slow-inactivated state induced by 5-s depolarization (EGFP: n = 6 cells, PRRT2: n = 6 cells). Data were presented as mean ± s.e.m. Two-tailed, unpaired Student’s t-test was used in C, and two-way ANOVAs were used in D, F, and G. ****P < 0.0001. n.s.,not significant.

PRRT2 promotes steady-state slow inactivation.

(A)Protocol of steady-state slow inactivation test (left) and representative traces of the sodium currents evoked by test pulses (right). The sodium currents gradually decreased as the voltage in conditioning steps increased. Scale bar, 0.5 nA by 2 ms. (B and C) The effects of PRRT2 on voltage-dependent steady-state slow inactivation of Nav1.2 channels (B) (EGFP: n = 11 cells, mPRRT2: n = 13 cells). mPRRT2, mouse PRRT2. Data were presented as mean ± s.e.m. Two-way ANOVAs were used in (B) to determine the statistical significance of main effects of group. ****P < 0.0001.

Fast inactivation of Nav channels is rapidly triggered by brief depolarization and manifests as a current decay following a transient pore opening (Figure 1B), consistent with previous reports(Misra, Kahlig, & George, 2008; Thompson et al., 2023). To evaluate the kinetics of fast inactivation, we measured the time constant of current decay and found that PRRT2 expression did not significantly alter the rate at which Nav channels entered the fast-inactivated state (Figure 1C). Moreover, the majority of Nav1.2 channels recovered from fast inactivation within ∼10 ms at a hyperpolarized potential of −120 mV, and PRRT2 expression had a minimal effect on this recovery (Figure 1D).

In contrast to fast inactivation, Nav channel slow inactivation develops and recovers over much longer timescales (Figure 1E), typically ranging from hundreds of milliseconds to seconds, or even to minutes(Silva, 2014). Because slow-inactivated channels do not recover during brief hyperpolarization, they can be functionally distinguished from fast-inactivated channels using a classic two-pulse protocol. In this approach, a brief hyperpolarization step (e.g., −120 mV, 10 ms) is inserted between a conditioning pulse and a test pulse to recover fastinactivated channels. Channels that have entered the slow-inactivated state remain unavailable during the test pulse, resulting in a reduction in peak sodium current (Figure 1E)(Featherstone, Richmond, & Ruben, 1996).

To assess the development of slow inactivation, we applied depolarization pulses of variable duration, followed by a fixed 10-ms hyperpolarization before the test pulse (Figure 1F). We found that a 5-second depolarization at 0 mV was sufficient to drive most Nav1.2 channels into the slow-inactivated state (Figure 1F), consistent with previous studies(Ganguly, Thompson, & George, 2021). PRRT2 significantly promoted the entry of Nav1.2 channels into the slow-inactivated state during sustained depolarization (Figure 1F).

To examine the role of PRRT2 in channel recovery from slow inactivation, we used a protocol in which a fixed 5-second depolarization at 0 mV was followed by varying durations of hyperpolarization at −120 mV to allow channel repriming (Figure 1G). PRRT2 expression markedly slowed the recovery of Nav channels from the slow-inactivated state (Figure 1G).

To further investigate whether the regulation of PRRT2 in slow inactivation is voltage dependent, we performed a steady-state slow inactivation protocol. In this assay, 10-s depolarizing steps from −110 to −10 mV were applied to drive Nav channels into steady state, followed by a fixed hyperpolarization and a test pulse to assess channel availability (Figure 1-figure supplement 1A). We found that PRRT2 promotes the transition of Nav channels into the slowinactivated state across a broad range of membrane potentials. PRRT2 expression induced a pronounced hyperpolarizing shift in the inactivation curve, with a V0.5 of −76.27 mV compared to −48.05 mV in controls (Figure 1-figure supplement 1B). Together, these findings indicate that PRRT2 is a potent regulator of Nav1.2 channel slow inactivation in vitro.

Functional auxiliary factors in the regulation of Nav channel slow inactivation

Several auxiliary factors of Nav channels have been identified, including the SCN1B and FHF2A (also known as FGF13a), both of which are implicated in modulating Nav channel inactivation(Dover et al., 2010; Isom et al., 1992). In this study, we compared the effects of mouse SCN1B and FGF13a with those of PRRT2 on the development and recovery of Nav channel slow inactivation (Figure 2A).

Functional auxiliary factors in the regulation of Nav channel slow inactivation.

(A) Diagram of Nav channel with potent auxiliary factors, encompassing SCN1B, FHF2A (also known as FGF13a) and PRRT2. (B) The effects of auxiliary factors on the entry of Nav1.2 channels into slow inactivation (EGFP: n = 12 cells, mPRRT2: n = 14 cells, SCN1B: n = 9 cells, FGF13a: n = 12 cells). (C) The effects of auxiliary factors on the recovery of Nav1.2 channels from slow-inactivated state induced by 5-s depolarization (EGFP: n = 12 cells, mPRRT2: n = 12 cells, SCN1B: n = 12 cells, FGF13a: n = 12 cells). (D) Schematic representation of the putative topology of mouse PRRT2 and three truncates. Numbers indicate the positions of the amino acid. REL, re-entrant loop. TM, transmembrane domain. N, amino terminus. C, carboxyl terminus. aa, amino acid. (E) The effects of PRRT2 truncates on the entry of Nav1.2 channels into slow inactivation (EGFP: n = 5 cells, mPRRT2(1–266): n = 14 cells, mPRRT2(222–346): n = 12 cells, mPRRT2(256–346): n = 12 cells). (F) The effects of PRRT2 truncates on the recovery of Nav1.2 channels from slow-inactivated state induced by 5-s depolarization (EGFP: n = 5 cells, mPRRT2(1–266): n = 10 cells, mPRRT2(222–346): n = 10 cells, mPRRT2(256–346): n = 12 cells). Data were presented as mean ± s.e.m. The main effect of group was assessed using two-way ANOVAs (B, C, E and F). **P < 0.01, ****P < 0.0001, n.s., not significant.

We found that SCN1B had negligible effects on both the entry into and recovery from the slow-inactivated state of Nav1.2 compared to controls (Figures 2B and C). In contrast, FGF13a markedly accelerated the onset of inactivation (Figure 2B), consistent with previous reports(Dover et al., 2010; Venkatesan et al., 2014). Notably, FGF13a had minimal effect on the recovery of Nav channels from the slow-inactivated state induced by a 5-second depolarization (Figure 2C). These distinct effects of FGF13a on inactivation suggest that FGF13a does not modulate classical slow inactivation, but rather promotes a form of rapid-onset and long-term inactivation(Dover et al., 2010; Venkatesan et al., 2014). By comparison, PRRT2 uniquely regulated slow inactivation by both facilitating entry into and delaying recovery from the slow-inactivated state of Nav channels (Figures 2B and C).

The rodent Prrt2 gene encodes a 346-amino-acid protein containing a proline-rich domain in the N-terminus, a re-entrant loop, and a transmembrane domain in the C-terminus (Figure 2D). To identify the functional domain(s) responsible for regulating slow inactivation, we generated a series of protein-truncating variants (Figure 2D).

We first examined PRRT2(222–346), which lacks the N-terminal proline-rich domain (Figure 2D). This variant retained activity similar to the full-length PRRT2, enhancing slow inactivation development and delaying recovery (Figures 2E and F). A shorter construct, PRRT2(256– 346), which retains the intact C-terminus, also preserved regulatory function in modulating slow inactivation compared to the EGFP control (Figures 2E and F). To further test whether the C-terminal region is necessary for the effect of PRRT2 on Nav channel slow inactivation, we examined PRRT2(1–266), a truncated variant lacking the membrane-associated C-terminal domains (Figure 2D). Unlike variants containing the C-terminal region, PRRT2(1–266) showed no detectable effect on either the development or recovery of Nav channel slow inactivation compared to the control (Figures 2E and F).

Together, these results indicate that the C-terminal region of PRRT2, including the re-entrant loop and transmembrane domain, is both necessary and sufficient to mediate its regulatory effect on slow inactivation of Nav1.2 channels in vitro.

Evolutionarily conserved effects of PRRT2 on Nav channel slow inactivation

The PRRT2 and PRRT2-like genes emerged in vertebrates and have been conserved throughout evolution(Sallman Almen et al., 2012). The alignment of amino acid sequences from zebrafish, mouse, and human PRRT2 revealed 77.46% identity between human and mouse PRRT2, and 43.60% identity between zebrafish and mouse PRRT2 (Figures 3A and 3B). Notably, the majority of conserved residues are located within the C-terminal region of the PRRT2 proteins (Figure 3A). Given the essential role of the C-terminus of mouse PRRT2 in regulating Nav channel slow inactivation, we examined whether PRRT2 proteins from different species exhibit similar effects.

Evolutionarily conserved effects of PRRT2 on Nav channel slow inactivation.

(A) Sequence alignment of PRRT2 protein in human, mouse and zebrafish. Conserved amino acids are highlighted and the membrane-associated domains in carboxyl terminus of PRRT2 are underlined. REL, re-entrant loop. TM, transmembrane domain. (B) Sequence identity of PRRT2 protein in human (hPRRT2), mouse (mPRRT2) and zebrafish (zfPRRT2). Identity is calculated by using protein BLAST tool from NCBI. (C) The effects of PRRT2 from different species on the entry of Nav1.2 channels into slow inactivation (EGFP: n = 16 cells, mPRRT2: n = 19 cells, hPRRT2: n = 20 cells, zfPRRT2: n = 11 cells). (D) The effects of PRRT2 from different species on the recovery of Nav1.2 channels from slow-inactivated state induced by 5-s depolarization (EGFP: n = 16 cells, mPRRT2: n = 19 cells, hPRRT2: n = 19 cells, zfPRRT2: n = 11 cells). (E) Diagram for the chimeric construct of PRRT2 originated from human and zebrafish. (F) The effect of chimeric PRRT2 on the entry of Nav1.2 channels into slow inactivation (EGFP: n = 11 cells, hPRRT2: n = 6 cells, zfPRRT2: n = 10 cells, Chimera: n = 11 cells). (G) The effect of Chimeric PRRT2 on the recovery of Nav1.2 channels from slow-inactivated state induced by 5-s depolarization (EGFP: n = 11 cells, hPRRT2: n = 6 cells, zfPRRT2: n = 10 cells, Chimera: n = 9 cells). Data were presented as mean ± s.e.m. The main effect of group was assessed using two-way ANOVAs (C, D, F and G). **P < 0.01, ****P < 0.0001, n.s., not significant.

Paralogs of PRRT2 regulate Nav channel slow inactivation.

(A) Sequence alignment of mouse TRARG1, TMEM233 and PRRT2 proteins. Conserved amino acids are highlighted and the transmembrane domains in carboxyl terminus of these proteins are underlined. REL, re-entrant loop. TM, transmembrane domain. (B) Schematic showing the phylogenetic relationship between members of Dispanin subfamily B (DspB). (C) The effects of members of DspB on the entry of Nav1.2 channels into slow inactivation (EGFP: n = 13 cells, TRARG1: n = 17 cells, TMEM233: n = 13 cells, PRRT2: n = 13 cells). (D) The effects of members of DspB on the recovery of Nav1.2 channels from slowinactivated states induced by 5-s depolarization (EGFP: n = 10 cells, TRARG1: n = 14 cells, TMEM233: n = 12 cells, PRRT2: n = 12 cells). Data were presented as mean ± s.e.m. The main effect of group was assessed using two-way ANOVAs (C and D). **P < 0.01, ****P < 0.0001, n.s., not significant.

By comparing the effects of zebrafish, mouse, and human PRRT2 on Nav channel slow inactivation, we found that all three PRRT2 orthologs similarly promoted the entry of Nav1.2 channels into the slow-inactivated state (Figure 3C). Surprisingly, in recovery assays, zebraf-ish PRRT2 was more effective at delaying the recovery process compared to its mammalian counterparts (Figure 3D). To identify the determinant domain of zebrafish PRRT2 for this enhanced effect, we generated a chimeric PRRT2 construct by fusing the C-terminal region of zebrafish PRRT2 to the N-terminal region of human PRRT2 (Figure 3E). The chimeric PRRT2 exhibited comparable effects on the rate of slow inactivation development to those observed with both human and zebrafish PRRT2 (Figure 3F). Notably, the chimeric variant displayed enhanced modulation of recovery, resembling zebrafish PRRT2 and significantly differing from human PRRT2 (Figure 3G). These findings indicate that C-terminal region of zebrafish PRRT2 confers its greater ability to modulate recovery of Nav1.2 from slow-inactivated state. We next extended our analysis to other DspB family members, including Trafficking regulator of GLUT4-1 (trarg1) and Transmembrane protein 233 (tmem233), both of which encode proteins with C-terminal regions similar to that of PRRT2 (Figure 3-figure supplement 3A and B). Given the critical role of the PRRT2 C-terminus in modulating slow inactivation, we hypothesized that TRARG1 and TMEM233 might exert similar regulatory effects. Under heterologous expression conditions, both TRARG1 and TMEM233 facilitated the entry of Nav1.2 channels into the slow-inactivated state and delayed their recovery (Figure 3-figure supplement 3C and D). These effects were comparable to those observed with mouse PRRT2, although TMEM233 exhibited slightly weaker modulation than PRRT2 and TRARG1 (Figure 3-figure supplement 3C and D).

Together, these results suggest that the role of PRRT2 in regulating Nav channel slow inactivation is evolutionarily conserved.

Uniform influence of PRRT2 on slow inactivation across human Nav isoforms

In humans, the nine Nav channel isoforms (Nav1.1-1.9) share high sequence homology and similar structural features(Goldin, 2001). To investigate whether PRRT2 modulates slow inactivation across different Nav channel isoforms, we selected Nav1.1, Nav1.4, and Nav1.5, which are expressed in distinct tissues, and assessed the effects of human PRRT2 on their slow inactivation properties (Figures 4A-H).

Uniform influence of PRRT2 on slow inactivation across human Nav isoforms.

(A) Protocol for examining the entry of Nav channels into slow-inactivated state. (B-D) The effects of human PRRT2 (hPRRT2) on the entry of Nav1.1, Nav1.4 and Nav1.5 channels into slow inactivation. In (B), analyses for Nav1.1 (EGFP: n = 10 cells, hPRRT2: n = 12 cells); In (C), analyses for Nav1.4 (EGFP: n = 11 cells, hPRRT2: n = 11 cells); In (D), analyses for Nav1.5 (EGFP: n = 17 cells, hPRRT2: n = 17 cells). (E) Protocol for examining the recovery of Nav channels from slow-inactivated state. (F-H) The effects of human PRRT2 on the recovery of Nav1.1, Nav1.4 and Nav1.5 channels from slow-inactivated state induced by 5-s depolarization. In (F), analyses for Nav1.1 (EGFP: n = 10 cells, hPRRT2: n = 12 cells); In (G), analyses for Nav1.4 (EGFP: n = 11 cells, hPRRT2: n = 11 cells); In (H), analyses for Nav1.5 (EGFP: n = 16 cells, hPRRT2: n = 15 cells). Data were presented as mean ± s.e.m. The main effect of group was assessed using two-way ANOVAs (B-D and F-H). ****P < 0.0001.

Absence of effect of PRRT2 on Kv1.4 channel inactivation.

(A) Illustration showing the whole-cell voltage-clamp recording in Kv1.4 stably expressing HEK293T cells (upper) and representative traces of potassium currents evoked by a depolarization at 40 mV (bottom). Scale bar, 1 nA by 50 ms. (B) The effects of human PRRT2 on the entry of Kv1.4 channels into inactivated state (EGFP: n = 9 cells, hPRRT2: n = 8 cells). (C) The effects of human PRRT2 on the recovery of Kv1.4 channels from inactivated states induced by 5-s depolarization (EGFP: n = 8 cells; hPRRT2: n = 7 cells). Data were presented as mean ± s.e.m. The main effect of group was assessed using two-way ANOVAs (B and C). n.s., not significant.

Both Nav1.1 and Nav1.4 exhibited similar kinetics in the development and recovery of slow inactivation (Figures 4B, C, F and G). In contrast, Nav1.5 showed greater resistance to entering the slow-inactivated state (Figures 4D and H), consistent with previous findings(Richmond et al., 1998). Despite these isoform-specific differences, the heterologous expression of PRRT2 consistently promoted entry of Nav channels into the slow-inactivated state (Fig. 4A-D) and delayed recovery from it (Figures 4E-H) across all three Nav isoforms tested. These results suggest that PRRT2 modulates slow inactivation through a common mechanism shared by multiple isoforms of Nav channels.

To assess whether the regulatory effect of PRRT2 on slow inactivation is specific to Nav channels, we further examined its influence on the A-type voltage-dependent potassium channel, Kv1.4 (Figure 4-figure supplement 4A). Kv1.4 exhibits two types of inactivation: N-type and C-type, which are functionally analogous to the fast and slow inactivation observed in mammalian Nav channels(Hoshi, Zagotta, & Aldrich, 1990). Unlike Nav channels, where fast and slow inactivation can be temporally separated, C-type inactivation in Kv1.4 cannot be clearly isolated from N-type inactivation due to the relatively slow recovery kinetics of the latter(Bett, Dinga-Madou, Zhou, Bondarenko, & Rasmusson, 2011). Given this limitation, we employed a modified assay that did not distinguish between N- and C-type inactivation. Using this protocol, we found that PRRT2 has negligible effects on both the entry into and recovery from the inactivated state of Kv1.4 (Figure 4-figure supplement 4B and C). Since recovery from inactivation in Kv1.4 is primarily governed by C-type inactivation(Bett et al., 2011), this lack of effect suggests that PRRT2 does not modulate slow inactivation in Kv1.4.

Together, these findings indicate that PRRT2 selectively regulates slow inactivation in Nav channels but does not influence inactivation in Kv1.4, supporting the channel-type specificity of PRRT2 function.

Intermolecular interaction between PRRT2 and Nav channels in vitro

PRRT2 may regulate slow inactivation by interacting with Nav channels. To examine the potential physical association between human PRRT2 and Nav1.2, we co-expressed PRRT2-HA and Flag-Nav1.2 in HEK293T cells and performed co-immunoprecipitation (co-IP) assays (Figures 5A and B). As controls, cells were transfected with one of the two interacting proteins (either PRRT2-HA or Flag-Nav1.2) together with the empty vectors carrying the corresponding tag of the other protein (Figures 5A and B). Immunoprecipitation using an anti-Flag antibody to capture Flag-Nav1.2 successfully co-precipitated PRRT2 in the co-transfection group, but not in the control groups where either protein was expressed (Figure 5A). Conversely, immunoprecipitation with anti-HA antibody showed that Nav1.2 was specifically co-precipitated with PRRT2-HA in the co-transfection group but not in controls (Figure 5B). These results confirm that PRRT2 and Nav1.2 formed a protein complex in vitro.

Intermolecular interaction between PRRT2 and Nav channels in vitro.

(A and B) Schematic diagram showing the co-immunoprecipitation for detecting human PRRT2-Nav1.2 interaction. Flag-tagged Nav1.2 (+), HA-tagged PRRT2 (+) and empty (−) vectors were transfected in HEK293T cells as indicated. The cell lysates were immunoprecipitated with anti-Flag (A) or anti-HA (B) magnetic beads, the captured proteins were analyzed by SDS-PAGE and immunoblotting. (C) Diagram for HA-tagged truncation of human PRRT2 (PRRT2(1–268)), in which the carboxyl terminus of PRRT2 was deleted. REL, re-entrant loop. TM, transmembrane domain. (D and E) Co-immunoprecipitation assay for potential interaction between Nav1.2 and PRRT2(1–268). The proteins immunoprecipitated by anti-Flag (D) or anti-HA (E) magnetic beads were analyzed by immunoblotting. Note that HA-tagged PRRT2(1–268) was detected by anti-HA antibody in (D and E). (F) Diagram for HA-tagged truncation of human PRRT2 (PRRT2(250–340)), in which the amino terminus of PRRT2 was deleted. (G and H) Co-immunoprecipitation assay for potent interaction between Nav1.2 and PRRT2(250–340). The proteins immunoprecipitated by anti-Flag (G) or anti-HA (H) were analyzed by immunoblotting. Note that HA-tagged PRRT2(250–340) was detected by anti-HA antibody in (G and H). (I and J) Co-immunoprecipitation assay for potential interaction between Nav1.1 and PRRT2. The proteins captured by anti-Flag (I) or anti-HA (J) magnetic beads were analyzed by immunoblotting. In (A, B, D, E, G, H, I and J), blots shown are representative of at least three independent co-immunoprecipitation experiments with similar results. Red rectangles indicate the immunoblotting of bait and potential prey proteins in co-immunoprecipitation. IB, immunoblot. IP, immunoprecipitation. MW, molecular weight.

To identify the Nav1.2-binding domain within human PRRT2, we generated truncated constructs encoding either the N-terminus (residues 1–268 amino acid) or C-terminus (residues 250–340 amino acid) of PRRT2. Using the same co-IP approach, we found that PRRT2(1– 268), which contains the intracellular proline-rich domain, failed to interact with Nav1.2 (Figures 5C-E). In contrast, the PRRT2(250–340), which includes the re-entry loop and transmembrane domain, was able to bind Nav1.2 (Figures 5F-H). These results demonstrate that C-terminal region of PRRT2 is both necessary and sufficient for interacting with Nav1.2 channels.

Given that PRRT2 modulates slow inactivation across multiple Nav isoforms, we hypothesized that it could interact with other Nav isoforms, such as Nav1.1. To test this, we coexpressed PRRT2-HA and Flag-Nav1.1 in HEK293T cells and performed co-IP assays. Under these conditions, PRRT2 was found to interact with Nav1.1 (Figures 5I and J), similar to the interaction observed with Nav1.2 (Figures 5A and B).

These findings contrast with previous reports, which reported that PRRT2 preferentially interacted with Nav1.2 over Nav1.1 in vitro(Franchi et al., 2023; Fruscione et al., 2018; Valente et al., 2023). This discrepancy may stem from differences in the experimental conditions used to assess protein-protein interactions. For example, in our study, Nav1.1 and PRRT2 were co-expressed in HEK293T cells, and cells were solubilized using the detergent n-Dodecyl β-D-maltoside (DDM). Variations in experimental conditions could influence the stability of PRRT2-Nav channel complex and the efficiency of Co-IP assay.

Collectively, our findings suggest that PRRT2 interacts directly with Nav channels through its membrane-associated C-terminal domains, forming molecular complexes in vitro.

PRRT2 forms a molecular complex with Nav1.2 in the mouse brain

To extend our findings from the heterologous expression system to an in vivo context, we first investigated whether PRRT2 forms a protein complex with Nav channels in brain tissue. Due to the limited efficacy of available antibodies for immunoprecipitating native PRRT2 or Nav1.2 from brain lysates, we employed a tag-based capture strategy, which relies on the appropriate insertion of a commonly used epitope tag into the target protein.

We generated Prrt2-V5 knock-in mice via CRISPR/Cas9 technology(Yang, Wang, & Jaenisch, 2014), in which a V5 epitope tag was inserted at the C-terminus of PRRT2 (Figures 6A and B). Notably, compared to wild-type mice, PRRT2 protein levels in lysates were markedly reduced in Prrt2-V5 knock-in mice, likely due to the insertion of V5 tag, whereas Nav1.2 expression remained unchanged (Figures 6C-E). Despite the reduced PRRT2 expression, coimmunoprecipitation assays were effective. Using anti-V5 nanobody-conjugated magnetic beads, PRRT2 was effectively immunoprecipitated from membrane lysates of brain tissue (Figure 6C). A notable amount of Nav1.2 were co-immunoprecipitated in the Prrt2-V5 knockin group, whereas only trace levels were detected in wild-type controls (Figures 6C and F). In contrast, ATP1B2, another membrane protein, was not co-immunoprecipitated in the Prrt2-V5 group (Figures 6C and F), indicating that the interaction between PRRT2 and Nav1.2 is specific. Together, these results provide in vivo evidence that PRRT2 forms a molecular complex with Nav1.2 channels in the mouse brain.

PRRT2 forms a molecular complex with Nav1.2 in the mouse brain.

(A) Schematic showing the generation of Prrt2-V5 knock-in (V5-KI) mice using Cas9 technology. V5-tag sequence was inserted precisely at the end of Prrt2 gene. PAM, protospaceradjacent motif (NGG). ssDNA, single-strand DNA. (B) Sanger sequencing around sgRNA targeting site in Prrt2-V5 knock-in mouse. The V5-tag, stop codon and homologous arms were denoted. (C) Co-immunoprecipitation assay for potential interaction between Nav1.2 and PRRT2-V5 in the brain tissue. The proteins immunoprecipitated by anti-V5 nanobody were analyzed by immunoblotting. Blots shown are representative of four independent co-immunoprecipitation experiments. IB, immunoblot. IP, immunoprecipitation. MW, molecular weight. WT, wild-type. KI, knock-in. (D-F) Quantification of the density of protein bands for PRRT2 (D) and Nav1.2 (E) in lysate, and for Nav1.2 and ATP1B2 in Co-immunoprecipitation experiments (F) (n = 4 mice for each group). Data were presented as mean ± s.e.m. In (D-F), two-tailed, paired Student’s t-test was used for determining statistical significance. **P < 0.01, ****P < 0.0001, n.s., not significant.

PRRT2 deficiency impairs the regulation of Nav channel slow inactivation in cortical neuronal axons

Given the dense distribution of both PRRT2 and Nav channels in the cerebral and cerebellar cortices(Yamano, Miyazaki, & Nukina, 2022), we next investigated whether PRRT2 regulates Nav channel slow inactivation in cortical neurons. Cortical slices were prepared from wild-type and Prrt2-mutant mice, the latter lacking PRRT2 expression(Tan et al., 2018). Unlike the relatively simple morphology of HEK293T cells, pyramidal neurons have complex arborizations, which present space-clamp limitations during whole-cell voltage-clamp recordings(Bar-Yehuda & Korngreen, 2008). To mitigate these issues, we followed established protocols to isolate axonal blebs from cortical pyramidal neurons(Hu & Shu, 2012). These spherical, membrane-bound blebs form at the cutting surface of cortical slices and are readily visualized (Figure 7A). Sodium currents recorded from axon blebs were voltage-dependent and sensitive to tetrodotoxin (TTX), confirming that they were primarily mediated by Nav channels(Hu & Shu, 2012).

PRRT2 deficiency impairs the regulation of Nav channel slow inactivation and neuronal resilience in mice.

(A) Schematic showing the isolated axonal bleb recording in cortical neuron (left) and the representative brightfield image for the axonal bleb in brain slice (right). The arrow indicates an axonal bled. Scale bar, 20 μm. (B) Sodium current density recorded in isolated axonal blebs from wild-type (WT) and Prrt2-mutant mice (WT: n = 18 blebs from 7 mice, Prrt2-mutant: n = 20 blebs from 8 mice). (C) Protocol for assessing Nav channel slow-inactivation (upper) and the representative traces of the sodium currents evoked by conditioning depolarization pulse and test pulses (bottom). Scale bar, 0.5 nA by 5 ms. (D) Fraction of available Nav channels after 5-s depolarization (WT: n = 18 blebs from 7 mice, Prrt2-mutant: n = 19 blebs from 8 mice). (E) Schematic illustration of the cortical electrostimulation and EEG recording in mice. Ref, reference electrode. Gnd, ground electrode. Stim, Stimulation. (F) Illustration showing the experimental schedule and the stimulation parameters. Stimulation was delivered once daily with stepwise increases in current intensity. (G) Representative traces and power of EEG signals in wild-type and Prrt2-mutant mice before and after 2-s cortical stimulation. After-discharges were indicated by the arrow and red line. Scale bar: 0.2 mV. Electrical stimulation (120 µA) was applied during 60–62 s period, where the stimulus-induced artifacts were removed prior to analysis. (H) Threshold of electrical stimulation to induce after-discharges in WT and Prrt2-mutant mice (WT: n = 16 mice, Prrt2-mutant: n = 8 mice). (I) Percentage of after-discharge occurrence in WT and Prrt2-mutant mice after electrical stimulation (WT: n = 16, Prrt2-mutant: n = 8). Data were presented as mean ± s.e.m. Two-tailed, unpaired Student’s t-test was used in (B and D) and two-tailed, unpaired Mann-Whitney test was used in (H). In (I), the main effect of group was assessed using two-way ANOVAs. ****P < 0.0001, n.s., not significant.

Protein levels of Nav1.2 are unchanged in brain tissue from Prrt2-mutant mice.

(A and B) Immunoblots (A) and quantifications (B) of Nav1.2 proteins in brain tissue of wildtype (WT) and Prrt2-mutant mice (n = 3 mice). Data were presented as mean ± s.e.m. Two-tailed, paired Student’s t-test was used in (B) to determine the statistical significance. n.s., not significant.

PRRT2 deficiency did not affect Nav1.2 protein expression in forebrain tissues (Figure 7-figure supplement 7A and B), nor did it alter the sodium current density in cortical axonal blebs (Figure 7B). Because long-term voltage-clamp recordings from isolated axonal bleb are technically challenging, we employed a simplified protocol to assess Nav channel slow inactivation, using a holding potential of −70 mV to approximate the resting membrane potential of cortical neurons. In this protocol, a single 5-s depolarization to 0 mV was applied, and the fraction of available Nav channels was determined after a 10-ms hyperpolarization (Figure 7C).

Approximately 20% of Nav channels in wild-type blebs remained available, compared to ∼40% Prrt2-mutant blebs (Figures 7C and D), indicating that during sustained depolarization, fewer Nav channels in the Prrt2-mutant group entered the slow-inactivated state. Notably, because the proportion of channels entering slow inactivation differed substantially between wild-type and Prrt2-mutant groups, direct comparison of recovery kinetics was not feasible and was therefore not performed. Nevertheless, these findings provide direct evidence that endogenous PRRT2 is required for effective regulation of Nav channel slow inactivation in cortical neurons.

PRRT2 deficiency compromises cortical resilience in awake mice

Slow inactivation of Nav channels is essential for limiting neuronal excitability during episodes of aberrant hyperactivity in the neuronal networks(Zang, Marder, & Marom, 2023). Given the ability of PRRT2 to regulate this process, we hypothesized that PRRT2 might contribute to cortical resilience in awake mice, a property reflecting the resistance of the cerebral cortex to pathological perturbations.

To assess this, we employed a stimulation-response paradigm in which stimulation electrodes were implanted into the left hemisphere of the cerebral cortex, targeting both the sensory and visual cortices, while EEG recording electrodes were placed in the contralateral hemisphere that receives projections from the stimulation region (Figure 7E). We applied pulsed electrostimulations (1-ms width at 60 Hz for 2 s) to the cortex of mice once daily with stepwise increases in intensity, and the minimal current required to evoke after-discharges (ADs) in the contralateral cortex was determined in each animal (Figure 7F). The current threshold required to induce after-discharges serves as a functional indicator of cortical resilience to aberrant excitatory inputs(Blume, Jones, & Pathak, 2004).

Prrt2-mutant mice exhibited significantly lower stimulation thresholds for evoking after-discharges compared to wild-type controls (Figures 7G and H). Specifically, stimulation at an average intensity of 85 µA reliably induced after-discharges in Prrt2-mutant mice, whereas a significantly higher intensity (190 μA) was required to elicit similar responses in wild-type mice (Figures 7H and I), suggesting that PRRT2 is essential for maintaining cortical resilience in response to excitatory perturbations.

Discussion

In this study, we identified PRRT2 as a crucial auxiliary factor of Nav channel slow inactivation. We demonstrate that PRRT2 exerts a dual action by both facilitating the transition of Nav channels into a slow-inactivated state and delaying their recovery. Importantly, this regulatory action is evolutionarily conserved, and requires the integrity of the PRRT2’s C-terminal region. Using both Prrt2-V5 knock-in and Prrt2-mutant mouse models, we provide compelling evidence that PRRT2 forms a molecular complex with Nav channels in vivo and plays an essential role in modulating their slow inactivation, thereby contributing to the maintenance of neuronal resilience to perturbations.

In eukaryotic Nav channels, ion permeability is regulated through at least four distinct and coordinated gating mechanisms: voltage-gated activation and deactivation, fast inactivation, and slow inactivation(Catterall, 2023). Slow inactivation, an evolutionarily conserved intrinsic property of Nav channels, is mediated by conformational rearrangements and has distinct molecular determinants from those governing fast inactivation(Payandeh, 2018; Silva, 2014; Vilin & Ruben, 2001).

Previous studies have suggested that slow inactivation involves conformational changes at multiple structural regions of the Nav channels. The external pore mouth(Balser et al., 1996; Xiong et al., 2006), the selectivity filter(Chatterjee et al., 2018; Todt, Dudley, Kyle, French, & Fozzard, 1999), the S6 segments(Chancey, Shockett, & O’Reilly, 2007; Y. Chen, Yu, Surmeier, Scheuer, & Catterall, 2006; S. Y. Wang & Wang, 1997), and the regions within S4, S4-S5 linker and S5 segments(Bendahhou, Cummins, Kula, Fu, & Ptacek, 2002; Cummins & Sigworth, 1996; Hayward et al., 1999; Mitrovic, George, & Horn, 2000; D. W. Wang, Makita, Kitabatake, Balser, & George, 2000), have been implicated in modulating slow inactivation of Nav channels. The present study did not directly verify whether these structural regions mediated the effect of PRRT2 on Nav channel slow inactivation. However, insights from molecular modeling of protein complex provide valuable clues. By using the recently developed AlphaFold3(Abramson et al., 2024), we modeled a putative interaction between PRRT2 and the Nav channel. The predicted complex suggests that the C-terminal region of PRRT2 interacts with the S3 segment of domain IV (DIV-S3) of Nav isoforms. This raises the possibility that residues within DIV-S3 may serve as critical determinants for the regulatory effects of PRRT2 on Nav channel slow inactivation.

The role of PRRT2 in the regulation of Nav channel slow inactivation appears to be unique compared to known regulators such as SCN1B and FHFs. The modulatory effects of the SCN1B on Nav channel inactivation were initially observed in Xenopus oocytes(Wallner, Weigl, Meera, & Lotan, 1993), where it was found to influence sodium currents. Subsequent studies proposed that the SCN1B impedes slow inactivation, a hypothesis later confirmed in mammalian cells co-expressing Nav1.4 and SCN1B(Webb et al., 2009). However, its regulatory effects on slow inactivation were varied across Nav isoforms(He & Soderlund, 2014; Makita, Bennett, & George, 1996; Vilin et al., 1999; Xu et al., 2007). In this study, we observed little effect of SCN1B on the onset and recovery of slow inactivation in Nav1.2, consistent with previous findings(Xu et al., 2007). In contrast to SCN1B, PRRT2 exerts a more consistent effect across Nav isoforms, enhancing slow inactivation regardless of the specific Nav subtype.

FHF proteins (FHF1-4, also known as FGF11-14) constitute a subfamily of FGFs that lack a secretion signal and function distinctly from traditional FGFs(Olsen et al., 2003). Previous studies demonstrated that FHFs mediate a rapid-onset, long-term inactivation, which differs mechanically from traditional slow inactivation(Dover et al., 2010). Unlike the FHF2A-mediated long-term inactivation, which can be triggered by a brief 2-ms depolarization, PRRT2 requires a longer depolarization to affect the onset of slow inactivation of Nav channels. Additionally, PRRT2 significantly delays the Nav channel recovery from slow inactivation induced by 5-second depolarization, whereas FHF2A has little effect on this process. These findings suggest that PRRT2 and FHF2A modulate Nav channel inactivation through different mechanisms.

In mammals, Nav channel isoforms (Nav1.1–1.9) are distributed across the nervous system, skeletal muscle, heart and secretory glands(Cusdin, Clare, & Jackson, 2008). Although we have shown that PRRT2 exerts a consistent effect on slow inactivation across various Nav channel isoforms in vitro, the regulation of Nav channel slow inactivation is likely to be more complex under physiological conditions. PRRT2 is predominantly expressed in the nervous system(W. J. Chen et al., 2011), with enrichment in excitatory neurons(Tan et al., 2018), making it unlikely to directly modulate Nav1.4 and Nav1.5 channels, which are primarily expressed in skeletal muscle and cardiac tissue, respectively(Cusdin et al., 2008). A similar rationale applies to Nav1.1, which is largely confined to inhibitory interneurons(Yu et al., 2006), where PRRT2 expression is minimal or absent.

The partial mismatch between PRRT2 and Nav channel expression in vivo raises an important question that how Nav channel slow inactivation is regulated in cells that do not express PRRT2. One possibility is the expression of alternative regulators, such as TRARG1 or TMEM232, both of which belong to the DspB family and are expressed in tissues distinct from PRRT2(Ehrlich, Lacey, & Ehrlich, 2020; Oort, Warden, Baumann, Knotts, & Adams, 2007; Santana-Varela et al., 2021; Shibata et al., 2007). These proteins may serve analogous functions in modulating Nav channel slow inactivation. Another possibility is that certain PRRT2-negative cells have a reduced requirement for Nav channel slow inactivation modulation due to their intrinsic rhythmic firing properties. For instance, in the cerebellar cortex, the PRRT2 is absent in Purkinje cells, which exhibit repetitive firing(Lu et al., 2021). In such rhythmic firing cells, strong regulation of Nav channel slow inactivation may be undesirable, as excessive accumulation of inactivated Nav channels could disturb the regular firing patterns of these cells.

Given the critical role of PRRT2 in regulating Nav channel slow inactivation, its deficiency is expected to alter neuronal responses to intense and prolonged activity, thereby contributing to dysfunction in the central nervous system. Evidence from both animal models and clinical studies supports this hypothesis. In the present study, we found that PRRT2 deficiency impairs slow inactivation of Nav channels in cortical neurons, leading to reduced resistance to pathological perturbations, as evidenced by an increased susceptibility to after-discharge induction in awake mice. Previous studies have also reported that loss-of-function mutations or knockout of PRRT2 increase intrinsic excitability in cerebellar granule cells and promote spreading depolarization in the cerebellar cortex(Binda, Valente, Marte, Baldelli, & Benfenati, 2021; Lu et al., 2021). In humans, PRRT2 mutations are associated with infantile convulsions(Lee et al., 2012), paroxysmal dyskinesia(W. J. Chen et al., 2011; J. L. Wang et al., 2011), and hemiplegic migraine(Riant et al., 2022), all of which are linked to dysfunction in neuronal network excitability. Based on these lines of evidence, we conclude that PRRT2 deficiency in excitatory neurons impairs Nav channel slow inactivation, thereby compromising neuronal resilience to aberrant excitability and ultimately contributing to these episodic neurological disorders.

Together, the identification of PRRT2 as an auxiliary factor regulating Nav channel slow inactivation has significant implications for understanding the fine-tuning mechanisms of Nav channel function. Additionally, these findings may inform the development of novel therapeutic strategies for treating neurological disorders associated with excitability dysregulation.

Limitations of the study

While our study successfully identifies the functional domain of PRRT2, the specific interaction site on the Nav channel that binds PRRT2 and mediates the regulation of slow inactivation remains unknown. Further investigations are required to map the PRRT2 binding region on Nav channels and to determine how mutations at these sites influence PRRT2-dependent modulation of slow inactivation. Cryo-electron microscopy (cryo-EM) is a promising approach for resolving the structural details of the Nav-PRRT2 complex, particularly given recent advancements in determining high-resolution structures of various Nav channel isoforms and their complexes with auxiliary factors(Jiang et al., 2020; Pan et al., 2019; Pan et al., 2021). A detailed structural characterization of the Nav-PRRT2 interaction would provide critical insights into the molecular mechanisms governing Nav channel slow inactivation.

Materials and methods

Molecular cloning

Mammalian expression constructs were generated using the pCAGIG backbone (Addgene, 11159), which contains an internal ribosome entry site (IRES) followed by EGFP(Matsuda & Cepko, 2004). Coding sequences for full-length or truncated protein-of-interests (POIs) were inserted in the pCAGIG vector using a seamless cloning method (Beyotime, D7010). The POIs including human PRRT2 (hPRRT2), hPRRT2(1–268), hPRRT2(250–340), mouse PRRT2 (mPRRT2), mPRRT2(222–346), mPRRT2(256–346), mPRRT2(1–266), zebrafish PRRT2 (zfPRRT2), mouse SCN1B, mouse FHF2A, mouse TRARG1 and mouse TMEM233. A PRRT2 chimeric construct was generated by fusing the N-terminal fragment of human PRRT2 (hPRRT2(1–222)) with the C-terminal region of zebrafish PRRT2 (zPRRT2(110–226)), and cloning the resulting sequence into pCAGIG vector. Plasmids encoding human PRRT2(W. J. Chen et al., 2011) were kindly provided by Dr. Zhi-Ying Wu’s Laboratory (Zhejiang University). Plasmids of pCAG-Flag-Nav1.2(Pan et al., 2019) and pCAG-Flag-Nav1.1(Pan et al., 2021) were provided by Huai-Zong Shen’s Laboratory (Westlake University). All constructs generated in this study were verified by Sanger sequencing (BioSune, Boshang Biology Technology, Shanghai).

Cell culture and heterologous expression

HEK293T cell lines stably expressing human Nav1.2 or Kv1.4 respectively and CHO cell lines stably expressing human Nav1.1, Nav1.4 or Nav1.5 were obtained from the laboratory of ICE Bioscience Inc. Wild-type HEK293T cells (SCSP-502) used for transient transfection were obtained from Cell Bank/Stem Cell Bank of Chinese Academy of Sciences. All cells were cultured in Dulbecco’s modified Eagle medium (DMEM; Gibco, 11995065) supplemented with 10% fetal bovine serum (FBS; Gibco,10099141) and penicillin (50 U/mL)-streptomycin (50 μg/mL) (Gibco,15140163), and maintained at 37 °C in a humidified incubator with 5% CO2. The HEK293T cell line used in this study is of female origin and tested negative for mycoplasma contamination.

For heterologous protein expression in whole-cell voltage-clamp recording, 2 μg of plasmid DNA was transfected into HEK293T or CHO cells using Lipofectamine 2000 (ThermoFisher, 11668-019) according to the manufacturer’s instructions. For co-immunoprecipitation assays, 2–3 μg of total plasmid DNA (encoding two proteins of interest) was used for co-transfection. Four hours after the transfection, the medium was replaced with fresh growth medium, and cells were maintained for an additional 20 hours before subsequent experiments.

Animals

Prrt2-mutant mice, carrying a premature stop codon in exon 2 of the Prrt2 gene(Tan et al., 2018), were used in electrophysiological recording experiments.

Prrt2-V5 knock-in mice, used for co-immunoprecipitation (Co-IP) experiments, were generated in this study via CRISPR/Cas9 technology(Yang et al., 2014). Briefly, single-guide RNA (sgRNA: 5’-TCTCCCACAGTGTATAAGTG-3’) targeting the location nearby the stop codon of Prrt2 gene allele, together with a single-stranded DNA donor template encoding a V5 tag flanked by 53-bp homology arms (ssDNA: 5’-CTTCCCTGTC TGTCTTTCCC TCTCCTCTCC CACAGTGTAT AAGGGCAAGC CCATCCCCAA CCCCCTGCTG GGCCTGGACA GCACCTGAGG GGCTCTGCCC TGCATTCCAA GACTTTTCTT CCTGTTGG-3’) and Cas9 mRNA, were microinjected into zygotes from fertilized C57BL/6J mice. The injected zygotes were implanted into pseudopregnant female mice. Offspring were screened for correct insertion of the V5 tag immediately upstream of the Prrt2 stop codon by PCR, and the insertion was subsequently verified by Sanger sequencing. Positive founder (F0) mice were crossed with C57BL/6J mice to establish the Prrt2-V5 knock-in line.

Both Prrt2-mutant and Prrt2-V5 lines were backcrossed onto the C57BL/6J background for more than 10 generations. C57BL/6J mice used for breeding were obtained from the Shanghai Laboratory Animal Center (SLAC), Chinese Academy of Science.

Mice were housed in a 12-h light-dark cycle (light on at 7:00 a.m.) with ad libitum access to food and water. All procedures involving animals were approved by the Animal Care and Use Committee of the Center for Excellence in Brain Science and Intelligence Technology, Chinese Academy of Sciences (Approval number, NA-009-2022). Both male and female mice were used in experiments.

Whole-cell voltage clamp recording for sodium currents in cell lines

Sodium currents were recorded from HEK293T or CHO cell lines stably expressing individual human Nav channel isoforms (ICE Bioscience Inc.). Cells were transfected with plasmids as described in the main text. Twenty-four hours after transfection, cells were trypsinized and replated at lower density onto poly-D-lysine-coated coverslips. EGFP-positive cells were selected for recording 24 hours later.

Whole-cell voltage-clamp recordings were conducted at room temperature (23-24 °C). The bath solution contained (in mM): 50 NaCl, 90 choline chloride, 3.5 KCl, 1 MgCl2, 2 CaCl2, 10 D-Glucose, 1.25 NaH2PO4, and 10 HEPES (pH adjusted to 7.4 with NaOH). The internal pipette solution contained (in mM): 50 CsCl, 60 CsF, 10 NaCl, 20 EGTA, 10 HEPES (pH adjusted to 7.2 with CsOH). The osmolarity of the external and internal solutions was adjusted to 300 and 290 mOsmol/kg, respectively.

Pipette electrodes were pulled from borosilicate glass capillary (Sutter Instruments, BF150-86-10) using a horizontal puller (P-97, Sutter Instruments). The pipette resistance was approximately 2 MΩ when filled with internal pipette solution. After whole-cell configuration was achieved, cells were allowed to equilibrate for 5 minutes prior to recording. Series resistance was compensated by 80% to minimize voltage errors. Currents were recorded with an amplifier (HEKA, EPC-10), sampled at 10 kHz, and acquired using Patchmaster (HEKA, v2×92).

Fast inactivation assay for Nav channels

To assess the fast inactivation kinetics of Nav channels, sodium currents were evoked by 20ms depolarizing pulse to 0 mV from a holding potential of −120 mV. The time constant (tau) of current decay was obtained by mono-exponential fitting of the current trace.

Recovery from fast inactivation was measured using a two-pulse protocol, in which a 10-ms depolarizing pulse to 0 mV (P1) was followed by a variable-duration recovery period at −120 mV and a second 10-ms test pulse at 0 mV (P2). Recovery was quantified as the normalized peak current of P2 relative to P1.

Slow inactivation assay for Nav channels

To evaluate the development of slow inactivation of Nav channels, a series of conditioning prepulses of increasing duration at 0 mV was applied, followed by a brief 10-ms recovery at −120 mV and a 10-ms test pulse at 0 mV. Inter-sweep intervals were set at 15 s. Available fractions of Nav channels were assessed by normalizing the peak currents in the test pulse to that of the conditioning pulse.

For slow inactivation recovery, a 5-s conditioning pulse at 0 mV was followed by variable recovery durations at −120 mV, then a 10-ms test pulse at 0 mV. Inter-sweep interval was 15 s. In experiments assessing the effect of zebrafish PRRT2 on slow inactivation recovery, a modified protocol was used to avoid cumulative inactivation. Within a single sweep, a 5-s depolarization was followed by a series of 10-ms test pulses to 0 mV, each separated by progressively longer recovery periods at −120 mV(Webb et al., 2009).

Steady-state slow inactivation assay for Nav channels

To measure steady-state slow inactivation, 10-s conditioning pulses ranging from −110 to −10 mV (in 10 mV increments) were applied, followed by a 10-ms hyperpolarizing step to −120 mV and subsequently a 5-ms test pulse to 0 mV. The peak test current was normalized to the maximum sodium current recorded across all conditions in the protocol and reported as I/Imax.

Whole-cell voltage clamp recording for potassium currents in cell lines

HEK-293 cells stably expressing Kv1.4 channels (ICE Bioscience Inc.) were used to examine the effects of PRRT2 on Kv1.4 channel inactivation. Plasmid transfection and cell preparation were conducted as described in the sodium current recordings section.

For potassium current recording, the external bath solution contained (in mM): 140 NaCl, 3.5 KCl, 1 MgCl2, 2 CaCl2, 10 D-Glucose, 1.25 NaH2PO4, 10 HEPES (pH adjusted to 7.4 with NaOH). The internal pipette solution contained (in mM): 20 KCl, 115 K-aspartate, 1 MgCl2, 5 EGTA, 2 Na2-ATP, 10 HEPES (pH adjusted to 7.2 with KOH).

Inactivation assay for Kv1.4 channels

Unlike Nav channels, the ‘fast’ (N-type) and ‘slow’ (C-type) inactivation components of Kv1.4 channels cannot be easily separated by brief hyperpolarization due to the slow recovery kinetics of N-type inactivation. As such, both components were assessed together in this study. To assess the onset of Kv1.4 inactivation, cells were held at −80 mV and subjected to a series of depolarizing prepulses to +40 mV with increasing durations, followed by a 20-ms test pulse at +40 mV. Inter-sweep interval was 15 s. The fraction of available Kv1.4 channels during the test pulse was calculated by normalizing the peak test current to the peak current during the prepulse step.

To evaluate recovery kinetics, a 5-s depolarizing pulse to +40 mV was applied to induce inactivation, followed by a series of test pulses (40 mV, 20 ms), each separated by progressively increasing recovery periods at −80 mV.

Slice preparation

Mice (postnatal day 14-20) were anesthetized with 4% isoflurane. The brain tissues were rapidly extracted and immersed in ice-cold sucrose-based cutting solution containing (in mM): 2.5 KCl, 1.25 NaH2PO4, 26 NaHCO3, 10 MgSO4, 0.5 CaCl2, 10 D-Glucose, 205 sucrose, 1 sodium pyruvate, 1.3 sodium L-ascorbate. Coronal cortical slices (250 μm) were prepared using a vibratome (Leica, VT1200S).

Slices were then transferred immediately to an incubation beaker containing aerated artificial cerebrospinal fluid (ACSF) composed of (in mM): 126 NaCl, 2.5 KCl, 2 MgSO4, 2 CaCl2, 26 NaHCO3, 1.25 NaH2PO4 and 25 dextrose (pH 7.4, 315 mOsm). Slices were incubated at 35 °C for 30 min, then maintained in the same solution at room temperature until use. The oxygenated ACSF was equilibrated with 95% O2 and 5% CO2.

Electrophysiological recording in isolated axonal blebs

Slices were transferred to a recording chamber and stabilized with a platinum anchor grid during recording. Resealed axonal ends (“blebs”) in the cerebral cortex were visualized using an upright infrared differential interference contrast (IR-DIC) microscope. Individual blebs were mechanically isolated by sweeping a cutting pipette beneath it, and subsequently patched with glass electrodes (resistance ∼7 MΩ).

For sodium current recordings, pipettes were filled with a potassium gluconate-based internal solution containing (in mM): 130 K-gluconate, 1 MgCl2, 11 EGTA, 1 CaCl2, 1 KCl, 2 Mg-ATP, 0.3 Na-GTP, 10 HEPES (pH adjusted to 7.2, ∼290 mOsm), supplemented with 2 mM TEA-Cl. The external ACSF solution was supplemented with 5 mM 4-Aminopydine (4-AP) and 200 μM CdCl2 to block K+ and Ca2+ currents, respectively.

Whole-cell voltage-clamp recordings were performed using a MultiClamp amplifier (Molecular Devices, 700B). Currents were filtered at 4 kHz (low-pass) and digitized at 20 kHz via a Digidata interface (Molecular Devices, 1440A). Series resistance was compensated by 80% to minimize voltage errors. Leak currents were subtracted by using an online P/4 procedure. The holding potential was set to −70 mV, approximating the resting membrane potential of cortical neurons. To assess sodium current density, a hyperpolarization prepulse to −110 mV (200 ms) was used to remove inactivation, followed by a 20-ms depolarization to 0 mV. Sodium current density (nA/pF) was calculated by normalizing peak current to bleb capacitance. To evaluate Nav channel slow inactivation a single 5-s depolarizing pulse (0 mV) was applied to induce slow inactivation followed by a test pulses (0 mV, 20 ms) separated by 10-ms recovery intervals at −70 mV. The fraction of available Nav channels was calculated by normalizing the peak current amplitude of each test pulse to the peak current measured at the onset of the 5-s depolarization.

Co-immunoprecipitation in transfected HEK293T cells

In co-immunoprecipitation experiments, one protein-of-interest was tagged with an HA epitope and the other with a Flag epitope. The pCAGIG vector bearing either a HA or Flag tag alone was used as a control in paired constructs. HEK293T cells transfected with a pair of constructs were harvested 24 hours post-transfection for protein expression. Cells were lysed in ice-cold lysis buffer (20 mM Tris-HCl, 150 mM NaCl, 0.5% n-Dodecyl β-D-maltoside (DDM, Anatrace, D310), and protease inhibitor cocktail (Selleckchem, B14002)) with gentle agitation at 4 °C for 2 hours. Lysates were centrifuged at 20,000 x g for 20 minutes at 4 °C to remove insoluble debris. The supernatant was transferred to a fresh tube and kept on ice. Protein concentration of supernatant was determined using a colorimetric bicinchoninic acid (BCA) assay (Beyotime, P0012S). For each immunoprecipitation (IP), 400 μg of proteins was incubated with 30 μL of anti-HA nanobody magnetic beads (AlfaLifeBio, KTSM1335) or anti-Flag magnetic beads (Bimake, B26102) at 4 °C with rotation for 1 hour. Beads were then separated with DynaMag spin and washed three times with wash buffer (20 mM Tris-HCl, 150 mM NaCl, 0.05% DDM). The captured proteins were eluted with 25 μL SDS-PAGE loading buffer containing 100 mM DTT, followed by heating at 45 °C for 10 min. Samples were stored at −20 °C prior to western blot analysis.

Co-immunoprecipitation in mouse brain tissues

Brain tissues (∼100 mg per sample) were freshly dissected from wild-type or Prrt2-V5 knockin mice, and homogenized on ice using glass homogenizers in 1 mL of Tris-buffered saline (TBS, 20 mM Tris-HCl, 150 mM NaCl, protease inhibitor cocktail). The homogenates were centrifuged at 800 x g for 10 min at 4 °C to remove nuclei and cell debris. Supernatants were subjected to centrifugation at 40,000 x g for 45 minutes at 4 °C to isolate membrane fractions. The resulting membrane-enriched pellets were resuspended in TBS and solubilized in lysis buffer containing 0.3% DDM for 3 hours at 4 °C. Lysates were then cleared by centrifugation at 20,000 x g for 20 minutes at 4 °C. Protein concentration was measured via BCA assay. For each IP, ∼200 μg of proteins was incubated with 30 μL of V5-Trap magnetic beads (Chromotek, v5tma-20) at 4 °C with rotation for 1.5 hours. Beads were washed three times with wash buffer (20 mM Tris-HCl, 150 mM NaCl, 0.15% DDM), and bound proteins were eluted with 25 μL SDS-PAGE loading buffer containing 100 mM DTT, followed by heating at 45 °C for 10 min. Eluates were kept at −20 °C until use in western blotting.

Western blotting

Protein samples were separated using 4-12% PAGE Gel (Nanjing ACE Biotechnology, ET15412L) in MES-SDS running buffer at 140 V for 60 minutes. Proteins were transferred onto polyvinylidene difluoride (PVDF) membranes at 300 mA for 2 hours at 4 °C. The PVDF membranes were blocked in 5% nonfat milk in TBST (TBS with 0.1% Tween-20) for 1 hour at room temperature, followed by incubation with indicated primary antibodies overnight at 4 °C with gentle agitation.

After three washes in TBST, membranes were incubated with HRP-conjugated secondary antibodies for 2 hours at room temperature. Following another three washes, chemiluminescent substrate (TIANGEN Biotech, PA112) was applied, and signals were detected using a blot imaging system (Analytikjena, ChemStudio 815). Densitometric analysis of protein bands was performed using QuantityOne software (Biorad, Quantity One 1-D analysis, v4.6.2). The primary antibodies we used in western blotting were as follow: Anti-human PRRT2 (Atlas antibodies, HPA014447), Anti-Nav1.2 (Alomone Labs, ASC-002), Anti-Nav1.1 (Alomone Labs, ASC-001), Anti-HA-Tag (Cell signaling technology, 3724S), Anti-GAPDH-HRP (KangChen Bio-tech,KC-5G5), Anti-ATP1B2 (Abcam, ab185210), Anti-mouse PRRT2 (Wiiget Biotech, Rp3246) and Anti-β Actin-HRP (Cell signaling technology, 5125S).

Animal Surgery

Mice were anesthetized with isoflurane (RWD life science, R510-22-10), using 4% for induction and 2% for maintenance. Body temperature was maintained throughout the procedure with a heating pad. Meloxicam (5 mg/kg, subcutaneous) was administered preoperatively for analgesia.

After securing the mice in a stereotaxic frame (RWD life science, Model 68528), 0.25% bupivacaine was locally injected at the incision sites for perioperative analgesia, and ophthalmic ointment was applied to protect the cornea. A midline incision was made to expose the skull, which was cleaned with 2% hydrogen peroxide using sterile cotton swabs to remove connective tissue. The skull surface was leveled to a stereotaxic plane, and the bregma was identified and marked. Once dry, the exposed skull was coated with a light-curing self-etch adhesive (3M ESPE Single Bond Universal).

Two small craniotomies (∼0.6 mm in diameter) were made over the left sensory cortex (anterior-posterior (AP) −0.8 mm; medial-lateral (ML) 1.5 mm) and visual cortex (AP −3.5 mm, ML 1.5 mm) for the insertion of bipolar stimulation electrodes (180 μm in diameter), which were implanted to a depth of −0.6 mm from the dura. Stainless steel screws were used as EEG electrodes: one was implanted into the right visual cortex (AP −2.25 mm, ML −1.5 mm), and the second into the right cerebellar vermis to serve as the reference. An additional screw was placed in the left cerebellar hemisphere as the ground. The electrodes and a custom-made head plate were affixed to the skull using dental cement.

Post-operatively, meloxicam (5 mg/kg, subcutaneously) was administered once daily for 3 consecutive days. Mice were allowed to recover for one week prior to electrical stimulation and EEG recording.

Electrostimulation and EEG recording in the cerebral cortex

One week after surgery, mice were habituated to a custom-built head plate holder for 30 minutes per day over three consecutive days prior to testing.

For stimulation experiments, the pre-implanted electrodes were connected to a stimulus isolator (AMPI, ISO Flex). Electrical pulses (1 ms pulse width, 60 Hz frequency, 2 s duration) were delivered through the isolator and controlled via TTL signals from a pulse generator (RWD life science, R820). EEG signals (bandpass: 1-300 Hz) were amplified 1000-fold using an AC/DC differential amplifier (A-M Systems, Model 3000), digitized at 1 kHz with a Digidata (Molecular Devices, 1332A), and acquired using Clampex software (Molecular Devices, v10.7).

To assess cortical resilience in Prrt2-mutant versus wild-type mice, a stimulus-response protocol was employed using progressively increasing current intensities (0, 20, 40, 80, 120, 160, 200, 240, 280, 320 µA). Each stimulation was separated by a 24-hour interval. The stimulation threshold was defined as the lowest current intensity that induced after-discharges (lasting ≥ 2 seconds) in the EEG trace. Each EEG recording lasted for 5 minutes, with electrical stimulation delivered immediately following a 1-minute baseline period.

Statistics

All grouped data are presented as mean ± standard error of mean (s.e.m) unless otherwise stated. Statistical analyses were performed using Graphpad Prism 8 (GraphPad Software Inc, La Jolla, US). No statistical methods were used to pre-determine the sample sizes. The D’Agostino-Pearson omnibus (K2) test was used to assess normality.

Two-tailed, unpaired Student’s t-tests were applied in Figures1B, 7B and 7D. A two-tailed, paired Student’s t-test was used in Figures 6D-F and supplement 4B. Two-tailed, unpaired Mann-Whitney test was applied in Figure 7H. Two-way ANOVAs were conducted for comparisons in Figures 1D, 1F-G, 2B-C, 2E-F, 3C-D, 3F-G, 4B-D, 4F-H, 7I, supplement 1B, supplement 2C-D, and supplement 3B-C. If necessary, the Gessier-Greenhouse correction was applied to adjust for violations of sphericity.

Statistical tests and sample sizes (including number of cells, slices, or animals) are stated in the figure legends. No data were excluded from analyses. Asterisks in figures denote significance levels: *P=<=0.05, **P=<=0.01, ***P=<=0.001, ****P=<=0.0001, n.s., not significant. Co-immunoprecipitation assays were independently repeated at least three times in both in vitro and in vivo tests. All attempts at replicating experiments presented in the manuscript have obtained consistent conclusions.

Electrophysiological recordings were conducted with the experimenter blinded to group identity (transfected condition or animal genotype).

eLife Assessment

This important study identifies PRRT2 as an auxiliary regulator of Nav channel slow inactivation, proposing that PRRT2 facilitates entry into, and delays recovery from, the slow-inactivated state. The evidence provided is compelling and well executed, though the work would be bolstered by additional studies of Nav1.6, as well as structural studies to directly investigate the molecular basis of gating modulation. Overall, this study will be of interest to ion channel biophysicists and neurophysiologists, particularly those studying channelopathies.

Reviewer #1 (Public review):

Summary:

The manuscript by Lu and colleagues demonstrates convincingly that PRRT2 interacts with brain voltage-gated sodium channels to enhance slow inactivation in vitro and in vivo. The work is interesting and rigorously conducted. The relevance to normal physiology and disease pathophysiology (e.g., PRRT2-related genetic neurodevelopmental disorders) seems high. Some simple additional experiments could elevate the impact and make the study more complete.

Strengths:

Experiments are conducted rigorously, including experimenter blinding and appropriate controls. Data presentation is excellent and logical. The paper is well written for a general scientific audience.

Weaknesses:

There are a few missing experiments and one place where data are over-interpreted.

(1) An in vitro study of Nav1.6 is conspicuously absent. In addition to being a major brain Na channel, Nav1.6 is predominant in cerebellar Purkinje neurons, which the authors note lack PRRT2 expression. They speculate that the absence of PRRT2 in these neurons facilitates the high firing rate. This hypothesis would be strengthened if PRRT2 also enhanced slow inactivation of Nav1.6. If a stable Nav1.6 cell were not available, then simple transient co-transfection experiments would suffice.

(2) To further demonstrate the physiological impact of enhanced slow inactivation, the authors should consider a simple experiment in the stable cell line experiments (Figure 1) to test pulse frequency dependence of peak Na current. One would predict that PRRT2 expression will potentiate 'run down' of the channels, and this finding would be complementary to the biophysical data.

(3) The study of one K channel is limited, and the conclusion from these experiments represents an over-interpretation. I suggest removing these data unless many more K channels (ideally with measurable proxies for slow inactivation) were tested. These data do not contribute much to the story.

(4) In Figure 2, the authors should confirm that protein is indeed expressed in cells expressing each truncated PRRT2 construct. Absent expression should be ruled out as an explanation for absent enhancement of slow inactivation.

Reviewer #2 (Public review):

Summary:

As a member of DspB subfamily, PRRT2 is primarily expressed in the nervous system and has been associated with various paroxysmal neurological disorders. Previous studies have shown that PRRT2 directly interacts with Nav1.2 and Nav1.6, modulating channel properties and neuronal excitability.

In this study, Lu et al. reported that PRRT2 is a physiological regulator of Nav channel slow inactivation, promoting the development of Nav slow inactivation and impeding the recovery from slow inactivation. This effect can be replicated by the C-terminal region (256-346) of PRRT2, and is highly conserved across species from zebrafish, mouse, to human PRRT2. TRARG1 and TMEM233, the other two DspB family members, showed similar effects on Nav1.2 slow inactivation. Co-IP data confirms the interaction between Nav channels and PRRT2. Prrt2-mutant mice, which lack PRRT2 expression, require lower stimulation thresholds for evoking after-discharges when compared to WT mice.

Strengths:

(1) This study is well designed, and data support the conclusion that PRRT2 is a potent regulator of slow inactivation of Nav channels.

(2) This study reveals similar effects on Nav1.2 slow inactivation by PRRT2, TMEM233, and TRARG1, indicating a common regulation of Nav channels by DspB family members (Supplemental Figure 2). A recent study has shown that TMEM233 is essential for ExTxA (a plant toxin)-mediated inhibition on fast inactivation of Nav channels; and PRRT2 and TRARG1 could replicate this effect (Jami S, et al. Nat Commun 2023). It is possible that all three DspB members regulate Nav channel properties through the same mechanism, and exploring molecules that target PRRT2/TRARG1/TMEM233 might be a novel strategy for developing new treatments of DspB-related neurological diseases.

Weaknesses:

(1) Previously, the authors have reported that PRRT2 reduces Nav1.2 current density and alters biophysical properties of both Nav1.2 and Nav1.6 channels, including enhanced steady-state inactivation, slower recovery, and stronger use-dependent inhibition (Lu B, et al. Cell Rep 2021, Fig 3 & S5). All those changes are expected to alter neuronal excitability and should be discussed.

(2) In this study, the fast inactivation kinetics was examined by a single stimulus at 0 mV, which may not be sufficient for the conclusion. Inactivation kinetics at more voltage potentials should be added.

(3) It is a little surprising that there is no difference in Nav1.2 current density in axon-blebs between WT and Prrt2-mutant mice (Figure 7B). PRRT2 significantly shifts steady-state slow inactivation curve to hyperpolarizing direction, at -70 mV, nearly 70% of Nav1.2 channels are inactivated by slow inactivation in cells expressing PRRT2 when compared to less than 10% in cells expressing GFP (Figure supplement 1B); with a holding potential of -70 mV, I would expect that most of Nav channels are inactivated in axon-blebs from WT mice but not in axon-blebs from Prrt2-mutant mice, and therefore sodium current density should be different in Figure 7B, which was not. Any explanation?

(3) Besides Nav channels, PRRT2 has been shown to act on Cav2.1 channels as well as molecules involved in neurotransmitter release, which may also contribute to abnormal neuronal activity in Prrt2-mutant mice. These should be mentioned when discussing PRRT2's role in neuronal resilience.

Reviewer #3 (Public review):

This paper reveals that the neuronal protein PRRT2, previously known for its association with paroxysmal dyskinesia and infantile seizures, modulates the slow inactivation of voltage-gated sodium ion (Nav) channels, a gating process that limits excitability during prolonged activity. Using electrophysiology, molecular biology, and mouse models, the authors show that PRRT2 accelerates entry of Nav channels into the slow-inactivated state and slows their recovery, effectively dampening excessive excitability. The effect seems evolutionarily conserved, requires the C-terminal region of PRRT2, and is recapitulated in cortical neurons, where PRRT2 deficiency leads to hyper-responsiveness and reduced cortical resilience in vivo. These findings extend the functional repertoire of PRRT2, identifying it as a physiological brake on neuronal excitability. The work provides a mechanistic link between PRRT2 mutations and episodic neurological phenotypes.

Comments:

(1) The precise structural interface and the molecular basis of gating modulation remain inferred rather than demonstrated.

(2) The in vivo phenotype reflects a complex circuit outcome and does not isolate slow-inactivation defects per se.

(3) Expression of PRRT2 in muscle or heart is low, so the cross isoform claims are likely of limited physiological significance.

(4) The mechanistic separation between the trafficking of PRRT2 and its gating effects is not clearly resolved.

(5) Additional studies with Nav1.6 should be carried out.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

Summary:

The manuscript by Lu and colleagues demonstrates convincingly that PRRT2 interacts with brain voltage-gated sodium channels to enhance slow inactivation in vitro and in vivo. The work is interesting and rigorously conducted. The relevance to normal physiology and disease pathophysiology (e.g., PRRT2-related genetic neurodevelopmental disorders) seems high. Some simple additional experiments could elevate the impact and make the study more complete.

Strengths:

Experiments are conducted rigorously, including experimenter blinding and appropriate controls. Data presentation is excellent and logical. The paper is well written for a general scientific audience.

We thank the reviewer for these positive comments and for the thoughtful evaluation of our work.

Weaknesses:

There are a few missing experiments and one place where data are over-interpreted.

(1) An in vitro study of Nav1.6 is conspicuously absent. In addition to being a major brain Na channel, Nav1.6 is predominant in cerebellar Purkinje neurons, which the authors note lack PRRT2 expression. They speculate that the absence of PRRT2 in these neurons facilitates the high firing rate. This hypothesis would be strengthened if PRRT2 also enhanced slow inactivation of Nav1.6. If a stable Nav1.6 cell were not available, then simple transient co-transfection experiments would suffice.

We thank the reviewer for raising this point. In our previous work, PRRT2 produced broadly similar effects on Nav1.2 and Nav1.6. Therefore, in the initial version of this study, we focused primarily on Nav1.2 as a representative neuronal Nav channel isoform and placed greater emphasis on testing whether PRRT2-dependent regulation of slow inactivation extends across additional Nav isoforms.

We have now performed new heterologous expression experiments to test whether PRRT2 modulates Nav1.6 slow inactivation. Consistent with our findings for other Nav isoforms, PRRT2 significantly enhances the slow inactivation of Nav1.6. We have incorporated these data into the revised Results and Figures, please refer to Page 8, Lines 211-215; Figures 4E and J.

(2) To further demonstrate the physiological impact of enhanced slow inactivation, the authors should consider a simple experiment in the stable cell line experiments (Figure 1) to test pulse frequency dependence of peak Na current. One would predict that PRRT2 expression will potentiate 'run down' of the channels, and this finding would be complementary to the biophysical data.

We thank the reviewer for this helpful suggestion. In the revised manuscript, we performed a pulse-train protocol in the stable Nav1.2 cell line and quantified the use-dependent attenuation (“run-down”) of peak sodium current across successive depolarizations (Figure 1-figure supplement 1C). Compared with control cells, PRRT2-expressing cells exhibited a larger decline in peak current during trains, indicating greater reduction in channel availability during repetitive depolarizations (Figure 1-figure supplement 1C). This pattern is consistent with our observations above showing that PRRT2 enhances Nav channel slow inactivation. These new data have been incorporated into the revised manuscript. Please refer to Page 5, Lines 133-140; Figure 1-figure supplement 1C.

(3) The study of one K channel is limited, and the conclusion from these experiments represents an over-interpretation. I suggest removing these data unless many more K channels (ideally with measurable proxies for slow inactivation) were tested. These data do not contribute much to the story.

We agree with the reviewer’s assessment. To avoid over-interpretation and to maintain focus on PRRT2-dependent regulation of Nav channel slow inactivation, we have removed the potassium channel dataset and the associated conclusions from the revised manuscript.

(4) In Figure 2, the authors should confirm that protein is indeed expressed in cells expressing each truncated PRRT2 construct. Absent expression should be ruled out as an explanation for the enhancement of slow inactivation.

We thank the reviewer’s concern regarding expression of the truncated PRRT2 constructs in the Nav1.2 stable cell line, particularly PRRT2(1-266), which shows little effect on slow inactivation of Nav1.2 channels. In the revised manuscript, we conducted western blot to verify expression of the PRRT2(1-266)-HA construct in the Nav1.2 stable cell line. We have added these results to the revised manuscript, please refer to Page 6, Lines 171-173; Figure 2-figure supplement 1A and B.

Reviewer #2 (Public review):

Summary:

As a member of DspB subfamily, PRRT2 is primarily expressed in the nervous system and has been associated with various paroxysmal neurological disorders. Previous studies have shown that PRRT2 directly interacts with Nav1.2 and Nav1.6, modulating channel properties and neuronal excitability.

In this study, Lu et al. reported that PRRT2 is a physiological regulator of Nav channel slow inactivation, promoting the development of Nav slow inactivation and impeding the recovery from slow inactivation. This effect can be replicated by the C-terminal region (256-346) of PRRT2, and is highly conserved across species from zebrafish, mouse, to human PRRT2. TRARG1 and TMEM233, the other two DspB family members, showed similar effects on Nav1.2 slow inactivation. Co-IP data confirms the interaction between Nav channels and PRRT2. Prrt2-mutant mice, which lack PRRT2 expression, require lower stimulation thresholds for evoking after-discharges when compared to WT mice.

Strengths:

(1) This study is well designed, and data support the conclusion that PRRT2 is a potent regulator of slow inactivation of Nav channels.

(2) This study reveals similar effects on Nav1.2 slow inactivation by PRRT2, TMEM233, and TRARG1, indicating a common regulation of Nav channels by DspB family members (Supplemental Figure 2). A recent study has shown that TMEM233 is essential for ExTxA (a plant toxin)-mediated inhibition on fast inactivation of Nav channels; and PRRT2 and TRARG1 could replicate this effect (Jami S, et al. Nat Commun 2023). It is possible that all three DspB members regulate Nav channel properties through the same mechanism, and exploring molecules that target PRRT2/TRARG1/TMEM233 might be a novel strategy for developing new treatments of DspB-related neurological diseases.

We thank the reviewer for careful evaluation and insightful suggestions.

Weaknesses:

(1) Previously, the authors have reported that PRRT2 reduces Nav1.2 current density and alters biophysical properties of both Nav1.2 and Nav1.6 channels, including enhanced steady-state inactivation, slower recovery, and stronger use-dependent inhibition (Lu B, et al. Cell Rep 2021, Fig 3 & S5). All those changes are expected to alter neuronal excitability and should be discussed.

We thank the reviewer for this suggestion. Although the present study focuses on PRRT2-dependent regulation of slow inactivation, we agree that PRRT2 may influence excitability through additional Nav-dependent mechanisms, including reduced current density and shifts in the voltage dependence of channel inactivation (Fruscione et al., 2018; Lu et al., 2021; Valente et al., 2023). Notably, because PRRT2 facilitates entry of Nav channels into slow-inactivated states both from closed states and from open states during prolonged depolarization, some of these previously reported effects may partly reflect enhanced slow inactivation and the resulting reduction in Nav channel availability. We have expanded the Discussion to integrate these prior findings and to clarify that these additional PRRT2-dependent effects may converge to shape neuronal excitability. Please refer to Page 16, Lines 445-452.

(2) In this study, the fast inactivation kinetics was examined by a single stimulus at 0 mV, which may not be sufficient for the conclusion. Inactivation kinetics at more voltage potentials should be added.

We thank the reviewer for this helpful suggestion. In the revised manuscript, we expanded our analysis of Nav1.2 fast-inactivation kinetics to include a range of test potentials (-20, -10, 0, +10, +20 and +30 mV) in the presence and absence of PRRT2. These experiments showed that PRRT2 expression did not significantly affect Nav1.2 fast-inactivation kinetics under these conditions. We have incorporated these new results into the revised manuscript. Please refer to Page 4, Lines 100-103; Figure 1C.

(3) It is a little surprising that there is no difference in Nav1.2 current density in axon-blebs between WT and Prrt2-mutant mice (Figure 7B). PRRT2 significantly shifts steady-state slow inactivation curve to hyperpolarizing direction, at -70 mV, nearly 70% of Nav1.2 channels are inactivated by slow inactivation in cells expressing PRRT2 when compared to less than 10% in cells expressing GFP (Figure supplement 1B); with a holding potential of -70 mV, I would expect that most of Nav channels are inactivated in axon-blebs from WT mice but not in axon-blebs from Prrt2-mutant mice, and therefore sodium current density should be different in Figure 7B, which was not. Any explanation?

We thank the reviewer for raising this point. In our axonal bleb recordings, although the holding potential was -70 mV, sodium current density was measured after a hyperpolarizing pre-pulse to -110 mV, which was applied before the test depolarization to relieve inactivation as much as possible (as described in the Methods). Therefore, the current density measurement in Figure 7B reflects the available current after this recovery step, rather than the steady-state availability at -70 mV. The lack of a difference in Figure 7B does not contradict the PRRT2-dependent shift in steady-state slow inactivation. In the revised manuscript, we have clarified this point explicitly in the Results and figure legend to avoid confusion. Please refer to Page 10, Lines 294-295.

(4) Besides Nav channels, PRRT2 has been shown to act on Cav2.1 channels as well as molecules involved in neurotransmitter release, which may also contribute to abnormal neuronal activity in Prrt2-mutant mice. These should be mentioned when discussing PRRT2's role in neuronal resilience.

We thank the reviewer for this suggestion. In addition to the Nav-dependent mechanisms, previous studies have shown that PRRT2 also regulates synaptic vesicle cycling (Valente et al., 2016; Coleman et al., 2018; Tan et al., 2018) and presynaptic surface expression of Cav2.1 channels (Ferrante et al., 2021). These effects are also expected to influence neurotransmitter release and, consequently, neuronal and network excitability. In the revised manuscript, we have expanded the Discussion to acknowledge that these additional PRRT2-dependent mechanisms may also contribute to cortical resilience. Please refer to Page 16, Lines 452-457.

Reviewer #3 (Public review):

This paper reveals that the neuronal protein PRRT2, previously known for its association with paroxysmal dyskinesia and infantile seizures, modulates the slow inactivation of voltage-gated sodium ion (Nav) channels, a gating process that limits excitability during prolonged activity. Using electrophysiology, molecular biology, and mouse models, the authors show that PRRT2 accelerates entry of Nav channels into the slow-inactivated state and slows their recovery, effectively dampening excessive excitability. The effect seems evolutionarily conserved, requires the C-terminal region of PRRT2, and is recapitulated in cortical neurons, where PRRT2 deficiency leads to hyper-responsiveness and reduced cortical resilience in vivo. These findings extend the functional repertoire of PRRT2, identifying it as a physiological brake on neuronal excitability. The work provides a mechanistic link between PRRT2 mutations and episodic neurological phenotypes.

We thank the reviewer for this positive evaluation of our work and for the constructive comments.

Comments:

(1) The precise structural interface and the molecular basis of gating modulation remain inferred rather than demonstrated.

We thank the reviewer for this comment. To avoid over-interpretation, we have removed the AlphaFold-based interaction prediction from the revised manuscript. We have also expanded the Limitations section to emphasize that direct structural and biochemical mapping of the PRRT2-Nav channel interface—through approaches such as targeted mutagenesis, crosslinking, and structural determination—will be required to define the binding interface and establish the molecular basis of gating modulation. Please refer to Page 16, Lines 465-468.

(2) The in vivo phenotype reflects a complex circuit outcome and does not isolate slow-inactivation defects per se.

We agree with the reviewer. Impaired slow inactivation in Prrt2-mutant mice is one plausible contributor to reduced cortical resilience. PRRT2 has also been reported to regulate surface exposure of Nav and Cav2.1 channels (Ferrante et al., 2021), as well as neuronal synaptic vesicle cycling (Valente et al., 2016; Coleman et al., 2018; Tan et al., 2018). Each of these PRRT2-associated processes could influence cortical excitability in vivo. We have therefore expanded the Discussion to clarify that the cortical phenotype likely reflects the combined contribution of multiple PRRT2-dependent mechanisms, rather than an isolated defect in slow inactivation alone. Please refer to Page 16, Lines 446-458.

(3) Expression of PRRT2 in muscle or heart is low, so the cross-isoform claims are likely of limited physiological significance.

We thank the review for this comment regarding physiological relevance. In the revised manuscript, we clarify that the cross-isoform analysis was intended to assess mechanistic generality at the channel level, rather than to imply equivalent physiological relevance across tissues. The functional consequence of PRRT2 depend on the Nav isoform composition and cellular context of each tissue. We also note that the broad isoform activity of the PRRT2 should be considered in any future attempt to manipulate PRRT2 function therapeutically. Please refer to Page 14 and 15, Lines 414-416; Lines 429-430.

(4) The mechanistic separation between the trafficking effect of PRRT2 and its gating effects is not clearly resolved.

We thank the reviewer’s concern regarding the possible contribution of trafficking effects to PRRT2-dependent regulation of Nav channel slow inactivation. Previous studies in heterologous overexpression systems have shown that PRRT2 can influence Nav channel trafficking and surface expression, raising the possibility that the observed effects on slow inactivation regulation might be secondary to altered channel abundance or localization. However, slow inactivation develops on a timescale of tens of milliseconds to seconds, whereas detectable changes in Nav channel trafficking and surface abundance generally occur over much longer intervals (minutes to hours) (Freal et al., 2023; Higerd-Rusli et al., 2023). These distinct temporal profiles argue against trafficking as the primary basis for the effects of PRRT2 on Nav channel slow inactivation described here, although direct quantification of dynamic changes in Nav channel surface expression will be required to fully exclude such a contribution (Liu et al., 2022; Tyagi et al., 2025). We have incorporated this point into the Discussion section. Please refer to Pages 13, Lines 378-388.

(5) Additional studies with Nav1.6 should be carried out.

We thank the reviewer for this suggestion. We have performed experiments to directly examine the effects of PRRT2 on Nav1.6 slow inactivation and incorporated these new data into the revised Results and figures, please refer to Page 8, Lines 211-215; Figures 4E and J.

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

Suggestions for future experiments (not for this paper)

(1) Exploit the lower protein expression in V5-PRRT2 mice to examine the effects of a hypomorphic allele.

We thank the reviewer for this insightful suggestion. We note that the V5 epitope knock-in reduced PRRT2 protein expression, which may functionally resemble a hypomorphic allele. Accordingly, in addition to its utility for biochemical experiments (e.g., co-immunoprecipitation), this line could serve as a genetic tool to interrogate PRRT2 dose-dependent effects in vivo. We have added this point to the revised manuscript, please refer to Page 9, Lines 265-267.

(2) Examine disease-causing PRRT2 mutations.

We thank the reviewer for this constructive suggestion. Testing disease-associated PRRT2 variants for their ability to regulate Nav channel slow inactivation would be an important next step to strengthen the disease relevance of the mechanism proposed here. Moreover, identifying missense variants that selectively disrupt slow-inactivation regulation could help pinpoint residues that are critical for PRRT2-Nav functional coupling and thereby inform future structure-function studies. We plan to pursue this direction in follow-up work.

(3) Investigate spreading depolarization in PRRT2-deficient mice.

We thank the reviewer for this suggestion. Although we have shown that PRRT2 deficiency facilitates spreading depolarization in the cerebellum, whether PRRT2 exerts similar control over spreading depolarization susceptibility in the cerebral cortex remains to be determined. We plan to address this in an independent study and to test how cortical spreading depolarization relates to other PRRT2-associated neurological disorders.

Reviewer #2 (Recommendations for the authors):

This study is, in general, well executed, and the manuscript is well written. However, I do have some questions.

(1) The authors' previous works have shown that PRRT2 regulates both Nav1.2 and Nav1.6, considering the wide expression Nav1.6 in CNS and its role in neuronal activity, what makes the authors not include Nav1.6 in this study?

We thank the reviewer for raising this question. In our previous work, PRRT2 produced broadly similar effects on Nav1.2 and Nav1.6. Therefore, in the initial version of this study, we focused primarily on Nav1.2 as a representative neuronal Nav channel isoform and placed greater emphasis on testing whether PRRT2-dependent regulation of slow inactivation extends across additional Nav isoforms. In response to reviewers’ concern, we have now performed new experiments to directly examine the effect of PRRT2 on Nav1.6 slow inactivation. These results have been incorporated into the revised manuscript. Please refer to Page 8, Lines 211-215; Figures 4E and J.

(2) Please explain why you chose 0 mV rather than -70 mV (closer to membrane potential) in the slow inactivation protocol.

We thank the reviewer for raising this question. Nav channels can enter into slow inactivation from both resting/closed states and activated/open states. In our steady-state slow-inactivation assays, we found that PRRT2 enhances Nav1.2 slow inactivation under both conditions (Figure 1-figure supplement 1A and B). In whole-cell recordings, Nav1.2 channels typically begin to activate at command voltages more depolarized than approximately -60 mV. Accordingly, a conditioning voltage of -70 mV predominantly probes entry into slow inactivation from closed states, whereas 0 mV drives channel activation and more effectively induces slow inactivation. We therefore chose 0 mV as the primary conditioning potential because it is widely used in conventional slow inactivation protocols and induces slow inactivation more robustly than conditioning voltages at -70 mV. We have added this explanation in Methods section of revised manuscript, please refer to Page 20, Lines 569-571.

(3) The authors mentioned that the insertion of V5 markedly reduced the PRRT2 protein level; thus, Prrt2-V5 knock-in mice could be considered as PRRT2 knock-down mice. Is there any noticeable difference in phenotype between Prrt2-V5 knock-in mouse and Prrt2-mutant mouse? In other words, is PRRT2 knockdown sufficient to affect neuronal excitability, or is a complete PRRT2 ablation required?

We thank the reviewer for raising this concern regarding the functional consequences of reduced PRRT2 expression in the Prrt2-V5 knock-in mice. Given that PRRT2 protein levels are markedly reduced in this line, and that cerebellar stimulation-induced dystonia is a characteristic phenotype of PRRT2 deficiency, we tested whether Prrt2-V5 knock-in mice also exhibit this phenotype. We found that electrical stimulation of the cerebellar cortex induced dystonia-like attacks in a subset of Prrt2-V5 knock-in mice. These dystonic behaviors resembled those previously observed in Prrt2-mutant mice, whereas no such behaviors were induced in wild-type mice (Figure 6-figure supplement 1). These findings indicate that a substantial reduction of PRRT2 expression (approximately 80%) is sufficient to impair neuronal function and elicit a disease-relevant phenotype in a subset of animals, supporting the interpretation that the V5 knock-in allele is hypomorphic. We have incorporated these results into the revised manuscript, please refer to Page 9, Lines 265-267; Figure 6-figure supplement 1.

(4) In Discussion (Page 13, lines 358-361), the authors mentioned a putative interaction between PRRT2 and the Nav channel by modeling, while there is no related data. Please either add modeling data or remove those sentences.

We thank the reviewer for this suggestion. To avoid over-interpretation, we have removed the statements regarding the AlphaFold-based interaction model from the revised manuscript. We agree that the interaction interface remains to be demonstrated experimentally, and we now discuss this point in the Limitations section. Please refer to Page 16, Lines 465-468.

(5) Typo: Page 14, line 399, "TMEM232" should be "TMEM233".

We thank the reviewer for pointing out this typo. We have corrected it in the revised manuscript.

Reviewer #3 (Recommendations for the authors):

(1) Mechanistic depth: While the functional data show altered slow-inactivation kinetics, the mechanistic explanation remains superficial. The AlphaFold-based prediction of PRRT2 interaction with DIV-S3 is speculative. The authors should clarify their illustrative rather than evidential intent and avoid over-interpretation.

We thank the reviewer for this comment. To avoid over-interpretation, we have removed the AlphaFold-based interaction prediction from the revised manuscript. We have also expanded the Limitations section to emphasize that direct structural and biochemical mapping of the PRRT2-Nav interface, including targeted mutagenesis, crosslinking, and structural determination, will be necessary to elucidate the molecular basis of this interaction and its effect on channel gating. Please refer to Page 16, Lines 465-468.

(2) Separation of trafficking vs. gating effects: Previous studies showed PRRT2 influences Nav trafficking and surface expression. Here, surface expression changes are not systematically quantified. Such an analysis would strengthen the argument that gating effects are not secondary to altered channel abundance or localization.

We thank the reviewer’s concern regarding the possible contribution of trafficking effects to PRRT2-dependent regulation of Nav channel slow inactivation. We agree that direct analysis of Nav channel surface localization during prolonged depolarization and hyperpolarization would provide stronger evidence to distinguish gating effects from trafficking-dependent mechanisms. However, such experiments are technically challenging in this context: conventional surface biotinylation assays do not provide the temporal resolution required for these rapid protocols, and live-cell imaging approaches to monitor dynamic changes in Nav channel surface expression during slow-inactivation paradigms have not yet been established in our laboratory.

Although PRRT2 has been reported to regulate Nav channel surface expression in heterologous systems, we consider it unlikely that trafficking is the major determinant of the slow-inactivation effects described here. Slow-inactivation develops on a timescale ranging from tens of milliseconds to seconds, whereas detectable changes in Nav channel trafficking and surface abundance generally occur over much longer timescales (minutes to hours) (Freal et al., 2023; Higerd-Rusli et al., 2023). We have expanded the Discussion in a revised manuscript. Please refer to Pages 13, Lines 378-388.

(3) Isoform generalization: Data on other Nav channel subtypes are presented as evidence of a conserved mechanism. However, given tissue-specific expression of PRRT2, these findings may be of limited in vivo relevance. At the very least, additional studies with Nav1.6 should be carried out.

We thank the review for this suggestion. In response, we conducted new experiments to examine the effect of PRRT2 on Nav1.6 slow inactivation. These results show that PRRT2 promotes entry of Nav1.6 channels into slow-inactivated states and delays their recovery, consistent with its effects on the other Nav isoforms examined in this study. We have incorporated these new data into the revised manuscript. Please refer to Page 8, Lines 211-215; Figures 4E and J.

Furthermore, we clarify that the cross-isoform analysis was intended to assess mechanistic generality at the channel level, rather than to imply equivalent physiological relevance across tissues. The functional consequence of PRRT2 depend on the Nav isoform composition and cellular context of each tissue. We also note that the broad isoform activity of the PRRT2 should be considered in any future attempt to manipulate PRRT2 function therapeutically. Pages 14 and 15, Lines 414-416 and 429-430.

(4) In vivo functional link: The EEG after-discharge threshold assay suggests decreased cortical resilience, but causality between slow-inactivation impairment and hyperexcitability remains indirect. Complementary in vivo recordings would strengthen the physiological link.

We thank the reviewer for this helpful suggestion. To further link impaired slow-inactivation to the hyperexcitability, we applied a repetitive stimulation protocol in corpus callosum slices, a white-matter region of brain enriched in both PRRT2 and Nav channels. During high-frequency stimulation (e.g., 20 Hz), the amplitude of the compound action potential progressively decreased over the course of the stimulus train. This phenomenon, often referred to as adaptation, reflects activity-dependent reduction in Nav channel availability (Fleidervish et al., 1996; Mickus et al., 1999; Kim et al., 2012). Compared with wild-type mice, Prrt2-mutant mice exhibited less adaptation during high-frequency stimulation, consistent with impaired slow inactivation during repetitive activity, which may contribute to hyperexcitability (Figure 7-figure supplement 2). We have added these results to the revised manuscript. Please refer to Pages 11, Lines 311-322; Figure 7-figure supplement 2.

(5) Structural interaction: It remains unclear whether PRRT2 binds the α-subunit directly or through accessory proteins. Crosslinking or detergent-solubilization controls of different stringencies could clarify this.

We thank the reviewer for raising this important issue. We agree that our co-immunoprecipitation data do not distinguish whether PRRT2 associates with the Nav channel α-subunit directly or through other components of the protein complex. To avoid over-interpretation, we have revised the relevant text in the manuscript to remove any implication of direct binding and now describe the result as an association between PRRT2 and Nav channels.

We have also expanded the Limitations section to note that additional experiments, such as crosslinking and structural studies, will be required to define the interaction interface between PRRT2 and Nav channels. Please refer to Page 16, Lines 465-468.

(6) Comparisons to other regulators: The paper positions PRRT2 as distinct from FHFs and β-subunits. The data support this, but the discussion could more critically assess whether PRRT2 acts by stabilizing a pore-based inactivated conformation, as suggested for other slow-inactivation modulators.

We thank the reviewer for this insightful suggestion. At present, relatively few modulators have been characterized in detail with respect to their effects on Nav channel slow-inactivation kinetics. Moreover, even for compounds such as lacosamide, which has been proposed to act as a slow-inactivation modulator, the underlying mechanism remains under debate (Errington et al., 2008; Jo and Bean, 2017). Therefore, in the revised manuscript, we discussed the possible mechanism of PRRT2 in the context of current models of Nav channel slow inactivation.

Previous studies suggest that entry into the slow-inactivated state involves at least two coupled processes: conformational changes in the voltage-sensing domains and structural rearrangements in the pore region, including the selectivity filter and intracellular activation gate (Catterall et al., 2024; Silva, 2014). During prolonged depolarization, voltage sensors become stabilized in the up-state, while the pore undergoes progressive rearrangements associated with slow inactivation (Balser et al., 1996; Vilin et al., 1999). Thus, mechanisms that further stabilize voltage sensors in the up-state and/or facilitate pore-based inactivated conformations could enhance slow inactivation.

Within this framework, PRRT2 may enhance slow inactivation by facilitating one or both of these processes, although direct evidence is still lacking. We have incorporated this discussion in relative section of revised manuscript. Please refer to Page 14, Lines 389-404.

Response references:

Jo S, Bean BP. Lacosamide Inhibition of Nav1.7 Voltage-Gated Sodium Channels: Slow Binding to Fast-Inactivated States. Mol Pharmacol. 2017 Apr;91(4):277-286.

Errington AC, Stöhr T, Heers C, Lees G. The investigational anticonvulsant lacosamide selectively enhances slow inactivation of voltage-gated sodium channels. Mol Pharmacol. 2008 Jan;73(1):157-69.

(7) Behavioral/clinical link: Given the strong human genetics background of PRRT2 disorders, a brief analysis or reference to electrophysiological phenotypes in patient neurons would contextualize the cortical findings.

We thank the reviewer for this suggestion. Previous studies showed that iPSC-derived excitatory neurons from a patient carrying a homozygous PRRT2 mutation exhibited increased sodium currents and neuronal hyperexcitability (Fruscione et al., 2018). Given that slow inactivation regulates Nav channel availability and thereby influences neuronal excitability, these electrophysiological abnormalities in patient-derived neurons may, at least in part, reflect impaired PRRT2-dependent regulation of Nav channel slow inactivation. We have added this point to the relative section of the revised manuscript. Please refer to Pages 15, Lines 432-437.

Minor comments

(1) Figures should include statistical sample sizes (n) and ideally overlay data points rather than only means {plus minus} SEM.

We thank the reviewer for this suggestion. In the revised manuscript, we present both individual data points and mean ± SEM in the column graphs. For the line graphs, individual data points were not overlaid because of space and readability constraints, and these panels therefore display mean ± SEM only. Sample sizes for each group are provided in the corresponding figure legends.

(2) The AlphaFold model should be provided as a supplementary figure with confidence scores indicated.

We thank the reviewer for this suggestion. However, because the predicted Nav1.2-PRRT2 interaction interface has not yet been experimentally validated in our study, we chose to remove the AlphaFold-based model from the revised manuscript to avoid over-interpretation.

(3) Clarify whether TTX sensitivity was verified in the axonal bleb preparation.

We thank the reviewer for raising this point. We verified the identity of the sodium currents in the axonal bleb preparation by their sensitivity to TTX, and this information has now been added to Figure 7A in the revised manuscript. Please refer to Page 10, Line 290; Figure 7A.

Version 2

Abstract

During sustained activity, voltage-gated sodium (Nav) channels enter a slow-inactivated state to limit cellular hyperexcitability. Disruption of this regulatory process has been implicated in skeletal, cardiac and neurological disorders. While the kinetics of this process are well characterized, its endogenous modulators remain unclear. Here, we identify Proline-Rich Transmembrane Protein 2 (PRRT2) as a native regulator of Nav channel slow inactivation. We show that PRRT2 facilitates the entry of Nav channels into the slow-inactivated state and delays their recovery, a regulatory effect conserved from zebrafish to humans. PRRT2 forms molecular complexes with Nav channels both in vitro and in vivo. In the mouse cortex, PRRT2 deficiency impairs the slow inactivation of Nav channels in neuronal axons, leading to reduced cortical resilience in response to hyperexcitable challenges. Together, these findings establish PRRT2 as a physiological modulator of Nav channel slow inactivation and reveal a mechanism that supports cortical resilience to pathological perturbations.

Introduction

Voltage-gated sodium (Nav) channels are critical for the initiation and propagation of electrical signals in excitable cells, contributing to cellular excitability in various tissues (Hodgkin & Huxley, 1952; Ulbricht, 1977). In mammals, including humans, Nav channels consist of nine α-subunit isoforms (Nav1.1-Nav1.9), which are differentially expressed and distributed across different tissues to regulate specific physiological functions (Catterall, Goldin, & Waxman, 2005). Mutations in Nav channels that lead to either loss- or gain-of-function of these channels are associated with a wide spectrum of diseases, including neurological disorders (Escayg & Goldin, 2010; Hedrich, Lauxmann, & Lerche, 2019; Meisler, Hill, & Yu, 2021), pain syndromes (Dib-Hajj, Geha, & Waxman, 2017; Waxman, 2013), muscle illnesses (Cannon, 2018; Mantegazza, Cestele, & Catterall, 2021) and cardiac diseases (Remme, 2013). Consequently, the Nav channels are considered promising therapeutic targets for treating these conditions (Noreng, Li, & Payandeh, 2021; Wisedchaisri & Gamal El-Din, 2022).

Voltage-dependent conformational changes enable Nav channels to transition between resting, open and inactivated states (Catterall, Wisedchaisri, & Zheng, 2020). Eukaryotic Nav channels undergo two major forms of inactivation (fast and slow inactivation), which are distinguished by their kinetics of onset and recovery (Goldin, 2003). Fast inactivation is triggered within milliseconds by brief depolarization and is rapidly reversed upon hyperpolarization, typically within tens of milliseconds (Armstrong & Bezanilla, 1977; Bezanilla & Armstrong, 1977). In contrast, slow inactivation develops during prolonged depolarizations and the slow inactivated channels require substantially longer periods to recover, ranging from hundreds of milliseconds to minutes (Rudy, 1978). Previous studies have shown that slow inactivation regulates Nav channel availability during sustained depolarization or high-frequency repetitive activity, thereby modulating the cellular excitability (Goldin, 2003; Silva, 2014; Vilin & Ruben, 2001). Impairment of this regulatory process due to mutations in Nav channels has been associated with disorders such as periodic paralysis (Hayward, Sandoval, & Cannon, 1999), cardiac arrhythmias (Richmond, Featherstone, Hartmann, & Ruben, 1998; Vilin, Makita, George, & Ruben, 1999) and epilepsy (Ghovanloo et al., 2023).

Beyond the intrinsic hierarchical gating mechanisms, auxiliary regulators further modulate Nav channel inactivation dynamics (Goldin, 2003). For example, β subunits of sodium channels (SCN1B) regulate inactivation of Nav channels in the heterologous expression system, although their effects vary across different Nav isoforms and cell types (He & Soderlund, 2014; Vilin et al., 1999; Webb, Wu, & Cannon, 2009). Fibroblast growth factor homologous factors (FHFs) interact with the carboxy terminus of Nav channels (Gade, Marx, & Pitt, 2020; Goetz et al., 2009), promoting intermediated or long-term inactivation, with recovery times spanning hundreds of milliseconds (Dover, Solinas, D’Angelo, & Goldfarb, 2010; Venkatesan, Liu, & Goldfarb, 2014). Although these auxiliary proteins have been identified as regulators of fast or intermediate inactivation, native modulators that specifically target slow inactivation of Nav channels remain largely unknown.

Recent studies on proline-rich transmembrane protein 2 (PRRT2) provide new insights into this issue. The PRRT2, a member of the Dispanins B subfamily (DspB) (Sallman Almen, Bringeland, Fredriksson, & Schioth, 2012), has been identified as a causative gene for paroxysmal kinesigenic dyskinesia(W. J. Chen et al., 2011; Lee et al., 2012; J. L. Wang et al., 2011). In vitro studies have shown that PRRT2 regulates Nav channel surface expression and delays recovery of these channels from inactivated states induced by brief depolarization (Fruscione et al., 2018; Lu et al., 2021; Valente et al., 2023). However, these studies did not clearly distinguish between the effects of PRRT2 on fast versus slow inactivation, nor did they provide direct in vivo evidence for its role in regulating Nav channel inactivation. Our original observations that PRRT2 expression increased the accumulation of inactivated Nav channels during repetitive stimulation (Lu et al., 2021) suggest that PRRT2 may be a previously unrecognized auxiliary regulator of Nav channel slow inactivation.

In this study, we combined whole-cell voltage-clamp recordings, co-immunoprecipitation, and in vivo approaches to investigate the role of PRRT2 in regulating Nav channel slow inactivation, and to examine how deficits in PRRT2-mediated modulation of this process affects neuronal function in awake mice.

Results

PRRT2 regulates the slow inactivation of Nav1.2 channels

Fast and slow inactivation of Nav channels proceed through distinct gating pathways (Goldin, 2003). Although previous studies have implicated PRRT2 in the regulation of Nav channel inactivation (Fruscione et al., 2018; Lu et al., 2021), it remains unclear whether PRRT2 primarily influences fast inactivation, slow inactivation or both. To address this, we expressed mouse PRRT2 in HEK293T cells stably expressing Nav1.2 (Figure 1A) and applied classic electrophysiological paradigms to distinguish between the two inactivation processes.

PRRT2 regulates slow inactivation of Nav1.2 channels

(A and B) Representative immunoblots showing PRRT2 protein expression (A) and schematic of whole-cell recording (B) in Nav1.2-stably expressing HEK293T cells transfected with constructs encoding either EGFP or mouse PRRT2 (mPRRT2). MW, molecular weight. (C) Fast-inactivation decay time constants of sodium currents evoked by 20-ms depolarizing steps to the indicated command voltages (EGFP: n = 13 cells, mPRRT2: n = 13 cells). Insert: representative sodium current traces elicited by a brief depolarization (left) and voltage-clamp protocol used to assess fast inactivation (right). The arrow indicates the decay of the sodium current attributable to fast inactivation. (D) Recovery from fast inactivation of Nav1.2 channels (EGFP: n = 10 cells, mPRRT2: n = 7 cells). Insert: Protocol for examining fast inactivation recovery. (E) Schematic diagram of entry into and recovery from slow inactivation of voltage-gated sodium channels. Slow inactivation modulates the availability of the sodium channels. (F) Entry into slow-inactivated state of Nav1.2 channels induced by progressively longer depolarization (EGFP: n = 6 cells, mPRRT2: n = 6 cells). (G) The effects of PRRT2 on the recovery of Nav1.2 channels from slow-inactivated state induced by 5-s depolarization (EGFP: n = 6 cells, PRRT2: n = 6 cells). Data are presented as mean ± s.e.m. Two-tailed, unpaired Student’s t-test was used in C, and two-way ANOVAs were used in D, F, and G. ****P < 0.0001. n.s., not significant.

Fast inactivation of Nav channels is rapidly triggered by brief depolarization and manifests as a current decay following a transient pore opening (Figure 1B and 1C), consistent with previous reports (Misra, Kahlig, & George, 2008; Thompson et al., 2023). To evaluate the kinetics of fast inactivation, we measured the time constant of current decay across a range of test potentials and found that PRRT2 expression did not significantly alter the rate at which Nav channels entered the fast-inactivated state (Figure 1C). Moreover, the majority of Nav1.2 channels recovered from fast inactivation within ∼10 ms at a hyperpolarized potential of -120 mV, and PRRT2 expression had a minimal effect on this recovery (Figure 1D).

In contrast to fast inactivation, Nav channel slow inactivation develops and recovers over much longer timescales (Figure 1E), typically ranging from hundreds of milliseconds to seconds, or even to minutes (Silva, 2014). Because slow-inactivated channels do not recover during brief hyperpolarization, they can be functionally distinguished from fast-inactivated channels using a classic two-pulse protocol. In this approach, a brief hyperpolarization step (e.g., -120 mV, 10 ms) is inserted between a conditioning pulse and a test pulse to recover fast-in-activated channels. Channels that have entered the slow-inactivated state remain unavailable during the test pulse, resulting in a reduction in peak sodium current (Figure 1E) (Featherstone, Richmond, & Ruben, 1996).

To assess the development of slow inactivation, we applied depolarization pulses of variable duration, followed by a fixed 10-ms hyperpolarization before the test pulse (Figure 1F). We found that a 5-second depolarization at 0 mV was sufficient to drive most Nav1.2 channels into the slow-inactivated state (Figure 1F), consistent with previous studies (Ganguly, Thompson, & George, 2021). PRRT2 significantly promoted the entry of Nav1.2 channels into the slow-inactivated state during sustained depolarization (Figure 1F).

To examine the role of PRRT2 in channel recovery from slow inactivation, we used a protocol in which a fixed 5-second depolarization at 0 mV was followed by varying durations of hyperpolarization at -120 mV to allow channel repriming (Figure 1G). PRRT2 expression markedly slowed the recovery of Nav channels from the slow-inactivated state (Figure 1G).

To further investigate whether the regulation of PRRT2 in slow inactivation is voltage dependent, we performed a steady-state slow inactivation protocol. In this assay, 10-s depolarizing steps from -110 to -10 mV were applied to drive Nav channels into steady state, followed by a fixed hyperpolarization and a test pulse to assess channel availability (Figure 1-figure supplement 1A). We found that PRRT2 promotes the transition of Nav channels into the slow-inactivated state across a broad range of membrane potentials. PRRT2 expression induced a pronounced hyperpolarizing shift in the inactivation curve, with a V0.5 of -76.27 mV compared to -48.05 mV in controls (Figure 1-figure supplement 1B).

In addition to prolonged depolarization, high-frequency repetitive depolarization can also induce Nav channel slow inactivation (Mickus, Jung, & Spruston, 1999). To determine whether PRRT2 regulates slow inactivation under these conditions, we applied a pulse-train protocol with an inter-pulse recovery interval of 10 ms. The use-dependent attenuation of peak sodium current was quantified as the fraction of current remaining across successive depolarizations (Figure 1-figure supplement 1C). Compared with the EGFP control, PRRT2 expression produced a greater decline in peak current during the train, consistent with enhanced slow inactivation during repetitive depolarizations (Figure 1-figure supplement 1C). Together, these findings indicate that PRRT2 is a potent regulator of Nav1.2 channel slow inactivation in vitro.

Functional auxiliary factors in the regulation of Nav channel slow inactivation

Several auxiliary factors of Nav channels have been identified, including the SCN1B and FHF2A (also known as FGF13a), both of which are implicated in modulating Nav channel inactivation (Dover et al., 2010; Isom et al., 1992). In this study, we compared the effects of mouse SCN1B and FGF13a with those of PRRT2 on the development and recovery of Nav channel slow inactivation (Figure 2A).

Functional auxiliary factors in the regulation of Nav channel slow inactivation

(A) Diagram of Nav channel with potent auxiliary factors, encompassing SCN1B, FHF2A (also known as FGF13a) and PRRT2. (B) The effects of auxiliary factors on the entry of Nav1.2 channels into slow inactivation (EGFP: n = 12 cells, mPRRT2: n = 14 cells, SCN1B: n = 9 cells, FGF13a: n = 12 cells). (C) The effects of auxiliary factors on the recovery of Nav1.2 channels from slow-inactivated state induced by 5-s depolarization (EGFP: n = 12 cells, mPRRT2: n = 12 cells, SCN1B: n = 12 cells, FGF13a: n = 12 cells). (D) Schematic representation of the putative topology of mouse PRRT2 and three truncations. Numbers indicate the positions of the amino acid. REL, re-entrant loop. TM, transmembrane domain. N, amino terminus. C, carboxyl terminus. aa, amino acid. (E) Effects of PRRT2 truncations on the entry of Nav1.2 channels into slow inactivation (EGFP: n = 5 cells, mPRRT2(1–266): n = 14 cells, mPRRT2(222–346): n = 12 cells, mPRRT2(256–346): n = 12 cells). (F) Effects of PRRT2 truncations on the recovery of Nav1.2 channels from slow-inactivated state induced by 5-s depolarization (EGFP: n = 5 cells, mPRRT2(1–266): n = 10 cells, mPRRT2(222–346): n = 10 cells, mPRRT2(256–346): n = 12 cells). Data are presented as mean ± s.e.m. The main effect of group was assessed using two-way ANOVAs (B, C, E and F). **P < 0.01, ****P < 0.0001, n.s., not significant.

We found that SCN1B had negligible effects on both the entry into and recovery from the slow-inactivated state of Nav1.2 compared to controls (Figures 2B and C). In contrast, FGF13a markedly accelerated the onset of inactivation (Figure 2B), consistent with previous reports (Dover et al., 2010; Venkatesan et al., 2014). Notably, FGF13a had minimal effect on the recovery of Nav channels from the slow-inactivated state induced by a 5-second depolarization (Figure 2C). These distinct effects of FGF13a on inactivation are consistent with the idea that FGF13a does not primarily modulate classical slow inactivation, but rather promotes a form of rapid-onset and long-term inactivation (Dover et al., 2010; Venkatesan et al., 2014). By comparison, PRRT2 uniquely regulated slow inactivation by both facilitating entry into and delaying recovery from the slow-inactivated state of Nav channels (Figures 2B and C).

The rodent Prrt2 gene encodes a 346-amino-acid protein containing a proline-rich domain in the N-terminus, a re-entrant loop, and a transmembrane domain in the C-terminus (Figure 2D). To identify the functional domain(s) responsible for regulating slow inactivation, we generated a series of protein-truncating variants (Figure 2D).

We first examined PRRT2(222–346), which lacks the N-terminal proline-rich domain (Figure 2D). This variant retained activity similar to the full-length PRRT2, enhancing slow inactivation development and delaying recovery (Figures 2E and F). A shorter construct, PRRT2(256– 346), which retains the intact C-terminus, also preserved regulatory function in modulating slow inactivation compared to the EGFP control (Figures 2E and F). To further test whether the C-terminal region is necessary for the effect of PRRT2 on Nav channel slow inactivation, we examined PRRT2(1–266), a truncated variant lacking the membrane-associated C-terminal domains (Figure 2D). Unlike variants containing the C-terminal region, PRRT2(1–266) showed no detectable effect on either the development or recovery of Nav channel slow inactivation compared to the control (Figures 2E and F). Notably, this lack of effect was not attributable to absent expression of the truncated construct (Figure 2-figure supplement 1A and B). Together, these results indicate that the C-terminal region of PRRT2, including the re-entrant loop and transmembrane domain, is both necessary and sufficient to mediate its regulatory effect on slow inactivation of Nav1.2 channels in vitro.

Evolutionarily conserved effects of PRRT2 on Nav channel slow inactivation

The PRRT2 and PRRT2-like genes emerged in vertebrates and have been conserved throughout evolution (Sallman Almen et al., 2012). The alignment of amino acid sequences from zebrafish, mouse, and human PRRT2 revealed 77.46% identity between human and mouse PRRT2, and 43.60% identity between zebrafish and mouse PRRT2 (Figures 3A and B). Notably, the majority of conserved residues are located within the C-terminal region of the PRRT2 proteins (Figure 3A). Given the essential role of the C-terminus of mouse PRRT2 in regulating Nav channel slow inactivation, we examined whether PRRT2 proteins from different species exhibit similar effects.

Evolutionarily conserved effects of PRRT2 on Nav channel slow inactivation

(A) Sequence alignment of PRRT2 protein in human, mouse and zebrafish. Conserved amino acids are highlighted and the membrane-associated domains in carboxyl terminus of PRRT2 are underlined. REL, re-entrant loop. TM, transmembrane domain. (B) Sequence identity of PRRT2 protein in human (hPRRT2), mouse (mPRRT2) and zebrafish (zfPRRT2). Identity is calculated by using protein BLAST tool from NCBI. (C) The effects of PRRT2 from different species on the entry of Nav1.2 channels into slow inactivation (EGFP: n = 16 cells, mPRRT2: n = 19 cells, hPRRT2: n = 20 cells, zfPRRT2: n = 11 cells). (D) The effects of PRRT2 from different species on the recovery of Nav1.2 channels from slow-inactivated state induced by 5-s depolarization (EGFP: n = 16 cells, mPRRT2: n = 19 cells, hPRRT2: n = 19 cells, zfPRRT2: n = 11 cells). (E) Diagram for the chimeric construct of PRRT2 originated from human and zebrafish. (F) The effect of chimeric PRRT2 on the entry of Nav1.2 channels into slow inactivation (EGFP: n = 11 cells, hPRRT2: n = 6 cells, zfPRRT2: n = 10 cells, Chimera: n = 11 cells). (G) The effect of chimeric PRRT2 on the recovery of Nav1.2 channels from slow-inactivated state induced by 5-s depolarization (EGFP: n = 11 cells, hPRRT2: n = 6 cells, zfPRRT2: n = 10 cells, Chimera: n = 9 cells). Data are presented as mean ± s.e.m. The main effect of group was assessed using two-way ANOVAs (C, D, F and G). **P < 0.01, ****P < 0.0001, n.s., not significant.

By comparing the effects of zebrafish, mouse, and human PRRT2 on Nav channel slow inactivation, we found that all three PRRT2 orthologs similarly promoted the entry of Nav1.2 channels into the slow-inactivated state (Figure 3C). Notably, in recovery assays, zebrafish PRRT2 was more effective at delaying the recovery process compared to its mammalian counterparts (Figure 3D). To identify the domain responsible for this enhanced effect, we generated a chimeric PRRT2 construct by fusing the C-terminal region of zebrafish PRRT2 to the N-terminal region of human PRRT2 (Figure 3E). The chimeric PRRT2 exhibited comparable effects on the rate of slow inactivation development to those observed with both human and zebrafish PRRT2 (Figure 3F). Notably, the chimeric variant displayed enhanced modulation of recovery, resembling zebrafish PRRT2 and significantly differing from human PRRT2 (Figure 3G). These findings indicate that the C-terminal region of zebrafish PRRT2 confers its greater ability to modulate recovery of Nav1.2 from the slow-inactivated state.

We next extended our analysis to other DspB family members, including Trafficking regulator of GLUT4-1 (trarg1) and Transmembrane protein 233 (tmem233), both of which encode proteins with C-terminal regions similar to that of PRRT2 (Figure 3-figure supplement 1A and B). Given the critical role of the PRRT2 C-terminus in modulating slow inactivation, we hypothesized that TRARG1 and TMEM233 might exert similar regulatory effects. Under heterologous expression conditions, both TRARG1 and TMEM233 facilitated the entry of Nav1.2 channels into the slow-inactivated state and delayed their recovery (Figure 3-figure supplement 1C and D). These effects were comparable to those observed with mouse PRRT2, although TMEM233 exhibited slightly weaker modulation than PRRT2 and TRARG1 (Figure 3-figure supplement 1C and D). Together, these results suggest that the role of PRRT2 in regulating Nav channel slow inactivation is evolutionarily conserved.

PRRT2 promotes slow inactivation across multiple human Nav channel isoforms

In humans, the nine Nav channel isoforms (Nav1.1-1.9) share high sequence homology and similar structural features (Goldin, 2001). To determine whether PRRT2 modulates slow inactivation across different Nav channel isoforms, we selected Nav1.1, Nav1.4, Nav1.5, and Nav1.6, which are expressed in distinct tissues and cell types, and assessed the effects of human PRRT2 on their slow inactivation properties (Figures 4A-J).

PRRT2 promotes slow inactivation across multiple human Nav channel isoforms

(A) Protocol for examining the entry of Nav channels into slow-inactivated state. (B-E) The effects of human PRRT2 (hPRRT2) on the entry of Nav1.1, Nav1.4, Nav1.5 and Nav1.6 channels into slow inactivation. In (B), analyses for Nav1.1 (EGFP: n = 10 cells, hPRRT2: n = 12 cells); In (C), analyses for Nav1.4 (EGFP: n = 11 cells, hPRRT2: n = 11 cells); In (D), analyses for Nav1.5 (EGFP: n = 17 cells, hPRRT2: n = 17 cells); In (E), analyses for Nav1.6 (EGFP: n = 5 cells, hPRRT2: n = 4 cells). (F) Protocol for examining the recovery of Nav channels from slow-inactivated state. (G-J) The effects of human PRRT2 on the recovery of Nav1.1, Nav1.4, Nav1.5 and Nav1.6 channels from slow-inactivated state induced by 5-s depolarization. In (G), analyses for Nav1.1 (EGFP: n = 10 cells, hPRRT2: n = 12 cells); In (H), analyses for Nav1.4 (EGFP: n = 11 cells, hPRRT2: n = 11 cells); In (I), analyses for Nav1.5 (EGFP: n = 16 cells, hPRRT2: n = 15 cells); In (J), analyses for Nav1.6 (EGFP: n = 5 cells, hPRRT2: n = 4 cells) Data are presented as mean ± s.e.m. The main effect of group was assessed using two-way ANOVAs (B-E and G-J). **P < 0.01, ***P < 0.001, ****P < 0.0001.

Nav1.1, Nav1.4, and Nav1.6 displayed similar kinetics in the development and recovery of slow inactivation (Figures 4B, C, E, G, H, and J). In contrast, Nav1.5 showed greater resistance to entering the slow-inactivated state (Figures 4D and I), consistent with previous findings (Richmond et al., 1998). Despite these isoform-specific differences, the heterologous expression of PRRT2 consistently promoted entry of Nav channels into the slow-inactivated state (Figures 4A-E) and delayed recovery from it (Figures 4F-J) across all four Nav isoforms tested. These results suggest that PRRT2 modulates slow inactivation through a common mechanism shared by multiple Nav channel isoforms, despite differences in their intrinsic slow-inactivation properties.

Association between PRRT2 and Nav channels in vitro

PRRT2 may regulate slow inactivation through association with Nav channels. To examine the potential physical association between human PRRT2 and Nav1.2, we co-expressed PRRT2-HA and Flag-Nav1.2 in HEK293T cells and performed co-immunoprecipitation (Co-IP) assays (Figures 5A and B). As controls, cells were transfected with one of the two interacting proteins (either PRRT2-HA or Flag-Nav1.2) together with the empty vectors carrying the corresponding tag of the other protein (Figures 5A and B). Immunoprecipitation using an anti-Flag antibody to capture Flag-Nav1.2 successfully co-precipitated PRRT2 in the co-transfection group, but not in the control groups where either protein was expressed (Figure 5A). Conversely, immunoprecipitation with anti-HA antibody showed that Nav1.2 was specifically co-precipitated with PRRT2-HA in the co-transfection group but not in controls (Figure 5B). These results confirm that PRRT2 and Nav1.2 formed a protein complex in vitro.

Association between PRRT2 and Nav channels in vitro

(A and B) Schematic diagram of the co-immunoprecipitation assay used to detect association between human PRRT2 and Nav1.2. Flag-tagged Nav1.2 (+), HA-tagged PRRT2 (+) and empty (-) vectors were transfected in HEK293T cells as indicated. The cell lysates were immunoprecipitated with anti-Flag (A) or anti-HA (B) magnetic beads, the captured proteins were analyzed by SDS-PAGE and immunoblotting. (C) Diagram for HA-tagged truncation of human PRRT2 (PRRT2(1–268)), in which the carboxyl terminus of PRRT2 was deleted. REL, re-entrant loop. TM, transmembrane domain. (D and E) Co-immunoprecipitation assay for potential association between Nav1.2 and PRRT2(1–268). The proteins immunoprecipitated by anti-Flag (D) or anti-HA (E) magnetic beads were analyzed by immunoblotting. Note that HA-tagged PRRT2(1–268) was detected by anti-HA antibody in (D and E). (F) Diagram for HA-tagged truncation of human PRRT2 (PRRT2(250–340)), in which the amino terminus of PRRT2 was deleted. (G and H) Co-immunoprecipitation assay for potential association between Nav1.2 and PRRT2(250–340). The proteins immunoprecipitated by anti-Flag (G) or anti-HA (H) were analyzed by immunoblotting. Note that HA-tagged PRRT2(250–340) was detected by anti-HA antibody in (G and H). (I and J) Co-immunoprecipitation assay for potential association between Nav1.1 and PRRT2. The proteins captured by anti-Flag (I) or anti-HA (J) magnetic beads were analyzed by immunoblotting. In (A, B, D, E, G, H, I and J), blots shown are representative of at least three independent co-immunoprecipitation experiments with similar results. Red rectangles indicate the immunoblotting of bait and potential prey proteins in co-immunoprecipitation. IB, immunoblot. IP, immunoprecipitation. MW, molecular weight.

To identify the Nav1.2-associating region within human PRRT2, we generated truncated constructs encoding either the N-terminus (residues 1–268) or C-terminus (residues 250–340) of PRRT2. Using the same Co-IP approach, we found that PRRT2(1–268), which contains the intracellular proline-rich domain, failed to associate with Nav1.2 (Figures 5C-E). In contrast, the PRRT2(250–340), which includes the re-entrant loop and transmembrane domain, was able to interact with Nav1.2 (Figures 5F-H). These results indicate that the C-terminal region of PRRT2 is necessary and sufficient, in this assay, for association with Nav1.2 channels. Given that PRRT2 modulates slow inactivation across multiple Nav isoforms, we hypothesized that it might also associate with other Nav isoforms, such as Nav1.1. To test this, we co-expressed PRRT2-HA and Flag-Nav1.1 in HEK293T cells and performed Co-IP assays. Under these conditions, PRRT2 was found to associate with Nav1.1 (Figures 5I and J), similar to the association observed with Nav1.2 (Figures 5A and B).

These findings contrast with previous reports, which suggested that PRRT2 preferentially interacted with Nav1.2 over Nav1.1 in vitro (Franchi et al., 2023; Fruscione et al., 2018; Valente et al., 2023). This discrepancy may stem from differences in the experimental conditions used to assess protein-protein interactions. For example, in our study, Nav1.1 and PRRT2 were co-expressed in HEK293T cells, and cells were solubilized using the detergent n-Dodecyl β-D-maltoside (DDM). Variations in experimental conditions could influence the stability of the PRRT2-Nav channel complex and the efficiency of the Co-IP assay.

Collectively, our findings suggest that PRRT2 associates with Nav channels through its membrane-associated C-terminal domains, forming molecular complexes in vitro.

PRRT2 forms a molecular complex with Nav1.2 in the mouse brain

To extend our findings from the heterologous expression system to an in vivo context, we first investigated whether PRRT2 forms a protein complex with Nav channels in brain tissue. Due to the limited efficacy of available antibodies for immunoprecipitating native PRRT2 or Nav1.2 from brain lysates, we employed a tag-based capture strategy, which relies on the appropriate insertion of a commonly used epitope tag into the target protein.

We generated Prrt2-V5 knock-in mice via CRISPR/Cas9 technology (Yang, Wang, & Jaenisch, 2014), in which a V5 epitope tag was inserted at the C-terminus of PRRT2 (Figures 6A and B). Notably, compared with wild-type mice, PRRT2 protein levels in tissue lysates were markedly reduced in Prrt2-V5 knock-in mice (Figure 6C-D). These mice displayed susceptibility to dystonia attacks, resembling the phenotype previously observed in Prrt2-mutant mice (Wu et al., 2025), but with lower penetrance (Figure 6-figure supplement 1A-C).

PRRT2 forms a molecular complex with Nav1.2 in the mouse brain

(A) Schematic showing the generation of Prrt2-V5 knock-in (V5-KI) mice using Cas9 technology. V5-tag sequence was inserted precisely at the end of Prrt2 gene. PAM, protospacer-adjacent motif (NGG). ssDNA, single-strand DNA. (B) Sanger sequencing around sgRNA targeting site in Prrt2-V5 knock-in mouse. The V5-tag, stop codon and homologous arms were denoted. (C) Co-immunoprecipitation assay for potential interaction between Nav1.2 and PRRT2-V5 in the brain tissue. The proteins immunoprecipitated by anti-V5 nanobody were analyzed by immunoblotting. Blots shown are representative of four independent co-immunoprecipitation experiments. IB, immunoblot. IP, immunoprecipitation. MW, molecular weight. WT, wild-type. KI, knock-in. (D-F) Quantification of the density of protein bands for PRRT2 (D) and Nav1.2 (E) in lysate, and for Nav1.2 and ATP1B2 in co-immunoprecipitation experiments (F) (n = 4 mice for each group). Data are presented as mean ± s.e.m. In (D-F), two-tailed, paired Student’s t-test was used for determining statistical significance. **P < 0.01, ****P < 0.0001, n.s., not significant.

Despite the reduced PRRT2 expression, Nav1.2 expression remained unchanged (Figure 6E), and co-immunoprecipitation assays remained effective in the Prrt2-V5 knock-in mice. Using anti-V5 nanobody-conjugated magnetic beads, PRRT2 was effectively immunoprecipitated from membrane lysates of brain tissue (Figure 6C). A notable amount of Nav1.2 was co-immunoprecipitated in the Prrt2-V5 knock-in group, whereas only trace levels were detected in wild-type controls (Figures 6C and F). In contrast, ATP1B2, another membrane protein, was not co-immunoprecipitated in the Prrt2-V5 knock-in group (Figures 6C and F), supporting the specificity of the association between PRRT2 and Nav1.2. Together, although the V5 knock-in allele is hypomorphic, these results provide in vivo evidence that PRRT2 forms a molecular complex with Nav1.2 channels in the mouse brain.

PRRT2 deficiency impairs the regulation of Nav channel slow inactivation in cortical neuronal axons

Given the dense distribution of both PRRT2 and Nav channels in the cerebral and cerebellar cortices (Yamano, Miyazaki, & Nukina, 2022), we next investigated whether PRRT2 regulates Nav channel slow inactivation in cortical neurons. Cortical slices were prepared from wild-type and Prrt2-mutant mice, the latter lacking PRRT2 expression (Tan et al., 2018). Unlike the relatively simple morphology of HEK293T cells, pyramidal neurons have complex arborizations, which present space-clamp limitations during whole-cell voltage-clamp recordings (Bar-Yehuda & Korngreen, 2008). To mitigate these issues, we followed established protocols to isolate axonal blebs from cortical pyramidal neurons (Hu & Shu, 2012). These spherical, membrane-bound blebs form at the cutting surface of cortical slices and are readily visualized (Figure 7A). Sodium currents recorded from axon blebs were voltage-dependent and sensitive to tetrodotoxin (TTX) (Figure 7A), confirming that they were primarily mediated by Nav channels (Hu & Shu, 2012).

PRRT2 deficiency impairs the regulation of Nav channel slow inactivation and neuronal resilience in mice

(A) Schematic showing the isolated axonal bleb recording in cortical neuron (left) and the representative brightfield image for the axonal bleb in brain slice (middle). The arrow indicates an axonal bleb. Scale bar, 20 μm. The representative sodium current traces recorded in axonal blebs (right). Scale bar, 0.5 nA, 5 ms. TTX, tetrodotoxin. (B) Sodium current density recorded in isolated axonal blebs from wild-type (WT) and Prrt2- mutant mice (WT: n = 18 blebs from 7 mice, Prrt2-mutant: n = 20 blebs from 8 mice). Sodium currents were measured with a brief depolarization to 0 mV preceded by a 200-ms hyperpolarizing prepulse to -110 mV. (C) Protocol for assessing Nav channel slow-inactivation (upper) and the representative traces of the sodium currents evoked by conditioning depolarization pulse and test pulses (bottom). Scale bar, 0.5 nA, 5 ms. (D) Fraction of available Nav channels after 5-s depolarization (WT: n = 18 blebs from 7 mice, Prrt2-mutant: n = 19 blebs from 8 mice). (E) Schematic illustration of the cortical electrostimulation and EEG recording in mice. Ref, reference electrode. Gnd, ground electrode. Stim, Stimulation. (F) Illustration showing the experimental schedule and the stimulation parameters. Stimulation was delivered once daily with stepwise increases in current intensity. (G) Representative traces and power of EEG signals in wild-type and Prrt2-mutant mice before and after 2-s cortical stimulation. After-discharges were indicated by the arrow and red line. Scale bar: 0.2 mV. Electrical stimulation (120 µA) was applied during 60–62 s period, where the stimulus-induced artifacts were removed prior to analysis. (H) Threshold of electrical stimulation to induce after-discharges in WT and Prrt2-mutant mice (WT: n = 16 mice, Prrt2-mutant: n = 8 mice). (I) Percentage of after-discharge occurrence in WT and Prrt2-mutant mice after electrical stimulation (WT: n = 16, Prrt2-mutant: n = 8). Data are presented as mean ± s.e.m. Two-tailed, unpaired Student’s t-test was used in (B and D) and two-tailed, unpaired Mann-Whitney test was used in (H). In (I), the main effect of group was assessed using two-way ANOVAs. ****P < 0.0001, n.s., not significant.

PRRT2 deficiency did not affect Nav1.2 protein expression in forebrain tissues (Figure 7-figure supplement 1A and B), nor did it alter the sodium current density in cortical axonal blebs when measured with a brief depolarization to 0 mV preceded by a 200-ms hyperpolarizing prepulse to -110 mV, a protocol designed to relieve inactivation before current elicitation (Figure 7B). Because prolonged voltage-clamp recordings from isolated axonal blebs are technically challenging, we adopted a simplified protocol to assess Nav channel slow inactivation, using a holding potential of -70 mV to approximate the resting membrane potential of cortical neurons. In this protocol, a single 5-s depolarization to 0 mV was applied, and the fraction of available Nav channels was determined after a 10-ms hyperpolarization (Figure 7C).

Approximately 20% of available sodium current remained in wild-type blebs, compared to ∼40% in Prrt2-mutant blebs (Figures 7C and D), indicating that during sustained depolarization, fewer Nav channels in the Prrt2-mutant group entered the slow-inactivated state. Notably, because the proportion of channels entering slow inactivation differed substantially between wild-type and Prrt2-mutant groups, direct comparison of recovery kinetics was not feasible and was therefore not performed. Nevertheless, these findings provide direct evidence that endogenous PRRT2 is required for effective regulation of Nav channel slow inactivation in cortical neurons.

PRRT2 deficiency compromises cortical resilience in awake mice

Excitatory neurons exhibit intrinsic adaptation, whereby their responses progressively decline during sustained or repetitive activation (Fleidervish, Friedman, & Gutnick, 1996; Kim, Owen, Holmes, & Grover, 2012). We therefore hypothesized that PRRT2 may contribute to this process by promoting Nav channel slow inactivation.

To test this idea, we applied a repetitive stimulation protocol in the corpus callosum, a whitematter region of the brain enriched in both PRRT2 and Nav channels. Compound action potentials, recorded with a glass electrode positioned 150 μm from the stimulation site, were used as a functional readout related to Nav channel availability during repetitive activity (Figure 7-figure supplement 2A). Compared with wild-type mice, Prrt2-mutant mice exhibited a smaller decline in compound action potential amplitude during high-frequency stimulation (20 Hz), but not during low-frequency stimulation (1 Hz), indicating reduced adaptation (Figure 7- figure supplement 2B and C). This pattern is consistent with impaired slow inactivation during repetitive activity.

Neuronal adaptation is important for limiting excitability during episodes of aberrant network hyperactivity (Fleidervish et al., 1996; Zang, Marder, & Marom, 2023). To further determine whether PRRT2-dependent impairment of neuronal adaptation contributes to hyperexcitability in vivo, we next assessed cortical resilience in awake mice, a measure of the resistance of the cerebral cortex to pathological perturbations. We used a stimulation-response paradigm in which stimulating electrodes were implanted into the left cerebral cortex, targeting both the sensory and visual cortices, while EEG recording electrodes were placed in the contralateral hemisphere, which receives projections from the stimulated region (Figure 7E). Pulsed electrical stimulation (1-ms pulse width at 60 Hz for 2 s) was delivered to the cortex of mice once daily with stepwise increases in current intensity, and the minimal current required to evoke after-discharges (ADs) in the contralateral cortex was determined in each animal (Figure 7F). The threshold current required to induce after-discharges was used as a functional indicator of cortical resilience to aberrant excitatory inputs (Blume, Jones, & Pathak, 2004).

Prrt2-mutant mice exhibited significantly lower stimulation thresholds for evoking after-discharges compared to wild-type controls (Figures 7G and H). Specifically, stimulation at an average intensity of 85 µA reliably induced after-discharges in Prrt2-mutant mice, whereas a significantly higher intensity (190 μA) was required to elicit similar responses in wild-type mice (Figures 7H and I), indicating that PRRT2 contributes importantly to maintaining cortical resilience in response to excitatory perturbations.

Discussion

In this study, we identified PRRT2 as a physiological auxiliary factor of Nav channel slow inactivation. We demonstrate that PRRT2 exerts a dual action by both facilitating the transition of Nav channels into a slow-inactivated state and delaying their recovery. Importantly, this regulatory action is evolutionarily conserved, and requires the integrity of the PRRT2 C-terminal region. Using both Prrt2-V5 knock-in and Prrt2-mutant mouse models, we provide compelling in vivo evidence that PRRT2 forms a molecular complex with Nav channels and modulates their slow inactivation, thereby contributing to neuronal adaptation and cortical resilience to perturbations.

In eukaryotic Nav channels, ion permeability is regulated through at least four distinct and co-ordinated gating mechanisms: voltage-gated activation and deactivation, fast inactivation, and slow inactivation (Catterall, 2023). Slow inactivation, an evolutionarily conserved intrinsic property of Nav channels, is mediated by conformational rearrangements and has distinct molecular determinants from those governing fast inactivation (Payandeh, 2018; Silva, 2014; Vilin & Ruben, 2001).

The role of PRRT2 in the regulation of Nav channel slow inactivation appears to be unique compared to known regulators such as SCN1B and FHFs. The modulatory effects of the SCN1B on Nav channel inactivation were initially observed in Xenopus oocytes (Wallner, Weigl, Meera, & Lotan, 1993), where it was found to influence sodium currents. Subsequent studies proposed that the SCN1B impedes slow inactivation, a hypothesis later confirmed in mammalian cells co-expressing Nav1.4 and SCN1B (Webb et al., 2009). However, its regulatory effects on slow inactivation varied across Nav isoforms (He & Soderlund, 2014; Makita, Bennett, & George, 1996; Vilin et al., 1999; Xu et al., 2007). In this study, we observed little effect of SCN1B on the onset and recovery of slow inactivation in Nav1.2, consistent with previous findings (Xu et al., 2007). In contrast to SCN1B, PRRT2 exerts a more consistent effect across Nav isoforms, enhancing slow inactivation regardless of the specific Nav subtype.

FHF proteins (FHF1-4, also known as FGF11-14) constitute a subfamily of FGFs that lack a secretion signal and function distinctly from traditional FGFs (Olsen et al., 2003). Previous studies demonstrated that FHFs mediate a rapid-onset, long-term inactivation, which differs mechanically from traditional slow inactivation (Dover et al., 2010). Unlike the FHF2A-mediated long-term inactivation, which can be triggered by a brief 2-ms depolarization, PRRT2 requires a longer depolarization to produce a detectable effect on the onset of slow inactivation of Nav channels. Additionally, PRRT2 significantly delays recovery of Nav channels from the slow-inactivated state induced by 5-second depolarization, whereas FHF2A has little effect on this process. These findings suggest that PRRT2 and FHF2A modulate Nav channel inactivation through different mechanisms.

Previous studies in heterologous overexpression systems have shown that PRRT2 can influence Nav channel trafficking and surface expression (Fruscione et al., 2018; Valente et al., 2023), raising the possibility that the observed effects on slow inactivation are secondary to altered channel abundance or localization. However, slow inactivation develops on a timescale of tens of milliseconds to seconds, whereas detectable changes in Nav channel trafficking and surface abundance generally occur over much longer intervals (minutes to hours) (Freal et al., 2023; Higerd-Rusli et al., 2023). These distinct temporal profiles argue against trafficking as the primary basis for the effects of PRRT2 on Nav channel slow inactivation described here, although direct quantification of dynamic changes in Nav channel surface expression will be required to fully exclude such a contribution (Liu, Wang, Pitt, & Liu, 2022; Tyagi, Higerd-Rusli, Akin, Waxman, & Dib-Hajj, 2025).

Our truncation analysis showed that the C-terminal region of PRRT2, which contains the reentrant loop and transmembrane domain, is sufficient to reproduce the enhancement of slow inactivation. This finding suggests that PRRT2 modulates slow inactivation through a membrane-embedded or membrane-proximal mechanism, rather than through its intracellular proline-rich N-terminus. Current models of Nav channel slow inactivation propose that entry into the slow-inactivated state involves coupled conformational changes in both the voltage-sensing domains and the pore region (Catterall, Gamal El-Din, & Wisedchaisri, 2024; Silva, 2014), including contributions from the external pore mouth (Balser et al., 1996; Xiong et al., 2006), the selectivity filter (Chatterjee et al., 2018; Todt, Dudley, Kyle, French, & Fozzard, 1999), the S6 segments (Chancey, Shockett, & O’Reilly, 2007; Y. Chen, Yu, Surmeier, Scheuer, & Catterall, 2006; S. Y. Wang & Wang, 1997), and regions encompassing S4, S4-S5 linker, and S5 (Bendahhou, Cummins, Kula, Fu, & Ptacek, 2002; Cummins & Sigworth, 1996; Hayward et al., 1999; Mitrovic, George, & Horn, 2000; D. W. Wang, Makita, Kitabatake, Balser, & George, 2000). Within this framework, the membrane-associated C-terminal region of PRRT2 may enhance slow inactivation by stabilizing one or both of these conformational rearrangements, although the precise structural basis of this effect remains to be determined.

In mammals, Nav channel isoforms (Nav1.1–1.9) are distributed across the nervous system, skeletal muscle, heart and secretory glands (Cusdin, Clare, & Jackson, 2008). Although we have shown that PRRT2 exerts a consistent effect on slow inactivation across various Nav channel isoforms in vitro, the regulation of Nav channel slow inactivation is likely to be more complex under physiological conditions. PRRT2 is predominantly expressed in the nervous system(W. J. Chen et al., 2011), with enrichment in excitatory neurons (Tan et al., 2018), making it unlikely to directly modulate Nav1.4 and Nav1.5 channels, which are primarily expressed in skeletal muscle and cardiac tissue, respectively (Cusdin et al., 2008). A similar rationale applies to Nav1.1, which is largely confined to inhibitory interneurons (Yu et al., 2006), where PRRT2 expression is minimal or absent. These observations suggest that the functional consequence of PRRT2 depend on the Nav isoform composition and cellular context of each tissue.

The partial mismatch between PRRT2 and Nav channel expression in vivo raises an important question about how Nav channel slow inactivation is regulated in cells that do not express PRRT2. One possibility is the expression of alternative regulators, such as TRARG1 or TMEM233, both of which belong to the DspB family and are expressed in tissues distinct from PRRT2 (Ehrlich, Lacey, & Ehrlich, 2020; Oort, Warden, Baumann, Knotts, & Adams, 2007; Santana-Varela et al., 2021; Shibata et al., 2007). These proteins may serve analogous functions in modulating Nav channel slow inactivation. Another possibility is that certain PRRT2-negative cells have a reduced requirement for Nav channel slow inactivation modulation due to their intrinsic rhythmic firing properties. For instance, in the cerebellar cortex, the PRRT2 is absent in Purkinje cells, which exhibit repetitive firing (Lu et al., 2021). In such rhythmic firing cells, strong enhancement of Nav channel slow inactivation may be undesirable, as excessive accumulation of inactivated Nav channels could disrupt their regular firing patterns. Thus, the broad isoform activity of the PRRT2 should be considered in any future attempt to manipulate PRRT2 function therapeutically.

Given the critical role of PRRT2 in regulating Nav channel slow inactivation, its deficiency would be expected to alter neuronal excitability and neuronal responses to prolonged or repetitive activity, thereby contributing to central nervous system dysfunction. Evidence from cultured neurons, animal models, and human genetics is broadly consistent with this view. Studies using human induced pluripotent stem cells (iPSC)-derived excitatory neurons have shown that homozygous PRRT2 mutation increases sodium currents and neuronal excitability (Fruscione et al., 2018). In mouse models, loss-of-function mutation or knockout of PRRT2 in cerebellar granule cells increases intrinsic excitability and facilitates spreading depolarization in the cerebellar cortex (Binda, Valente, Marte, Baldelli, & Benfenati, 2021; Lu et al., 2021), whereas in the forebrain, PRRT2 deficiency impairs Nav channel slow inactivation in cortical neurons and reduces resistance to pathological perturbation in awake mice. In addition, PRRT2 mutations in humans are associated with infantile convulsions (Lee et al., 2012), paroxysmal dyskinesia(W. J. Chen et al., 2011; J. L. Wang et al., 2011), and hemiplegic migraine (Riant et al., 2022), all disorders linked to abnormal neuronal network excitability.

Although the present study supports a functional link between PRRT2-dependent regulation of Nav channel slow inactivation and cortical resilience, PRRT2 may influence cortical excitability and stability through additional Nav-dependent mechanisms, including reduced current density and shifts in the voltage dependence of channel inactivation (Fruscione et al., 2018; Lu et al., 2021; Valente et al., 2023). Notably, because PRRT2 facilitates Nav channel entry into slow-inactivated states both from closed states and from open states during prolonged depolarization, some of these previously reported effects may partly reflect enhanced slow inactivation and the resulting reduction in Nav channel availability. In addition to these Nav-dependent mechanisms, previous studies have shown that PRRT2 also regulates synaptic vesicle cycling (Coleman et al., 2018; Tan et al., 2018; Valente et al., 2016) and presynaptic surface expression of Cav2.1 channels (Ferrante et al., 2021). These effects are also expected to influence neurotransmitter release and, consequently, neuronal and network excitability, and may therefore contribute to cortical resilience.

Together, the identification of PRRT2 as an auxiliary factor regulating Nav channel slow inactivation provides new insight into the fine-tuning of Nav channel function and offers a mechanistic framework for understanding how PRRT2 deficiency may contribute to episodic neurological disorders associated with excitability dysregulation.

Limitations of the study

While our study identifies the functional domain of PRRT2 and provides biological evidence that PRRT2 associates with Nav1.2, the precise binding site on the Nav channel that mediates PRRT2-dependent regulation of slow inactivation remains undefined. Direct structural and biochemical mapping of the PRRT2-Nav interface, including targeted mutagenesis, crosslinking, and structural determination, will be necessary to elucidate the molecular basis of this interaction and its effect on channel gating. Cryo-electron microscopy (cryo-EM) represents a promising approach for resolving the structural details of the Nav-PRRT2 complex, particularly in light of recent advancements in determining high-resolution structures of various Nav channel isoforms and their complexes with auxiliary factors (Jiang et al., 2020; Pan et al., 2019; Pan et al., 2021).

Materials and Methods

Molecular cloning

Mammalian expression constructs were generated using the pCAGIG backbone (Addgene, 11159), which contains an internal ribosome entry site (IRES) followed by EGFP (Matsuda & Cepko, 2004). Coding sequences for full-length or truncated proteins of interest (POIs) were inserted in the pCAGIG vector using a seamless cloning method (Beyotime, D7010). The POIs included human PRRT2 (hPRRT2), hPRRT2(1–268), hPRRT2(250–340), mouse PRRT2 (mPRRT2), mPRRT2(222–346), mPRRT2(256–346), mPRRT2(1–266), zebrafish PRRT2 (zfPRRT2), mouse SCN1B, mouse FHF2A, mouse TRARG1 and mouse TMEM233. A PRRT2 chimeric construct was generated by fusing the N-terminal fragment of human PRRT2 (hPRRT2(1–222)) with the C-terminal region of zebrafish PRRT2 (zPRRT2(110– 226)), and cloning the resulting sequence into pCAGIG vector. Plasmids encoding human PRRT2(W. J. Chen et al., 2011) were kindly provided by Dr. Zhi-Ying Wu’s Laboratory (Zhejiang University). Plasmids of pCAG-Flag-Nav1.2 (Pan et al., 2019) and pCAG-Flag- Nav1.1 (Pan et al., 2021) were provided by Huai-Zong Shen’s Laboratory (Westlake University). All constructs generated in this study were verified by Sanger sequencing (BioSune, Boshang Biology Technology, Shanghai).

Cell culture and heterologous expression

HEK293T cell lines stably expressing human Nav1.2 or Nav1.6 and CHO cell lines stably expressing human Nav1.1, Nav1.4 or Nav1.5 were obtained from the laboratory of ICE Bioscience Inc. Wild-type HEK293T cells (SCSP-502) used for transient transfection were obtained from Cell Bank/Stem Cell Bank of Chinese Academy of Sciences. All cells were cultured in Dulbecco’s modified Eagle medium (DMEM; Gibco, 11995065) supplemented with 10% fetal bovine serum (FBS; Gibco,10099141) and penicillin (50 U/mL)-streptomycin (50 μg/mL) (Gibco,15140163), and maintained at 37 °C in a humidified incubator with 5% CO2. The HEK293T cell line used in this study is of female origin and tested negative for mycoplasma contamination.

For heterologous protein expression in whole-cell voltage-clamp recording, 2 μg of plasmid DNA was transfected into HEK293T or CHO cells using Lipofectamine 2000 (ThermoFisher, 11668-019) according to the manufacturer’s instructions. For co-immunoprecipitation assays, 2–3 μg of total plasmid DNA (encoding two proteins of interest) was used for co-transfection. Four hours after the transfection, the medium was replaced with fresh growth medium, and cells were maintained for an additional 20 hours before subsequent experiments.

Animals

Prrt2-mutant mice, carrying a premature stop codon in exon 2 of the Prrt2 gene (Tan et al., 2018), were used in electrophysiological recording experiments.

Prrt2-V5 knock-in mice, used for co-immunoprecipitation (Co-IP) experiments, were generated in this study via CRISPR/Cas9 technology (Yang et al., 2014). Briefly, single-guide RNA (sgRNA: 5’-TCTCCCACAGTGTATAAGTG-3’) targeting the location nearby the stop codon of Prrt2 gene allele, together with a single-stranded DNA donor template encoding a V5 tag flanked by 53-bp homology arms (ssDNA: 5’-CTTCCCTGTC TGTCTTTCCC TCTCCTCTCC CACAGTGTAT AAGGGCAAGC CCATCCCCAA CCCCCTGCTG GGCCTGGACA GCACCTGAGG GGCTCTGCCC TGCATTCCAA GACTTTTCTT CCTGTTGG-3’) and Cas9 mRNA, were microinjected into zygotes from fertilized C57BL/6J mice. The injected zygotes were implanted into pseudopregnant female mice. Offspring were screened for correct insertion of the V5 tag immediately upstream of the Prrt2 stop codon by PCR, and the insertion was subsequently verified by Sanger sequencing. Positive founder (F0) mice were crossed with C57BL/6J mice to establish the Prrt2-V5 knock-in line.

Both Prrt2-mutant and Prrt2-V5 lines were backcrossed onto the C57BL/6J background for more than 10 generations. C57BL/6J mice used for breeding were obtained from the Shanghai Laboratory Animal Center (SLAC), Chinese Academy of Science.

Mice were housed in a 12-h light-dark cycle (light on at 7:00 a.m.) with ad libitum access to food and water. All procedures involving animals were approved by the Animal Care and Use Committee of the Center for Excellence in Brain Science and Intelligence Technology, Chinese Academy of Sciences (Approval number, NA-009-2022). Both male and female mice were used in experiments.

Whole-cell voltage clamp recording for sodium currents in cell lines

Sodium currents were recorded from HEK293T or CHO cell lines stably expressing individual human Nav channel isoforms (ICE Bioscience Inc.). Cells were transfected with plasmids as described in the main text. Twenty-four hours after transfection, cells were trypsinized and replated at lower density onto poly-D-lysine-coated coverslips. EGFP-positive cells were selected for recording 24 hours later.

Whole-cell voltage-clamp recordings were conducted at room temperature (23-24 °C). The bath solution contained (in mM): 50 NaCl, 90 choline chloride, 3.5 KCl, 1 MgCl2, 2 CaCl2, 10 D-Glucose, 1.25 NaH2PO4, and 10 HEPES (pH adjusted to 7.4 with NaOH). The internal pipette solution contained (in mM): 50 CsCl, 60 CsF, 10 NaCl, 20 EGTA, 10 HEPES (pH adjusted to 7.2 with CsOH). The osmolarity of the external and internal solutions was adjusted to 300 and 290 mOsmol/kg, respectively.

Pipette electrodes were pulled from borosilicate glass capillary (Sutter Instruments, BF150- 86-10) using a horizontal puller (P-97, Sutter Instruments). The pipette resistance was approximately 2 MΩ when filled with internal pipette solution. After whole-cell configuration was achieved, cells were allowed to equilibrate for 5 minutes prior to recording. Series resistance was compensated by 80% to minimize voltage errors. Currents were recorded with an amplifier (HEKA, EPC-10), sampled at 10 kHz, and acquired using Patchmaster (HEKA, v2x92).

Fast inactivation assay for Nav channels

To assess the fast-inactivation kinetics of Nav channels, sodium currents were evoked by 20-ms depolarizing steps to test potentials between -20 and +30 mV in 10-mV increments, from a holding potential of -120 mV. The decay time constant (tau) was obtained by mono-exponential fitting of the current trace.

Recovery from fast inactivation was measured using a two-pulse protocol, in which a 10-ms depolarizing pulse to 0 mV (P1) was followed by a variable-duration recovery period at -120 mV and a second 10-ms test pulse at 0 mV (P2). Recovery was quantified as the normalized peak current of P2 relative to P1.

Slow inactivation assay for Nav channels

To evaluate the development of slow inactivation of Nav channels, a series of conditioning prepulses of increasing duration at 0 mV was applied, followed by a brief 10-ms recovery at -120 mV and a 10-ms test pulse at 0 mV. Inter-sweep intervals were set at 15 s. Available fractions of Nav channels were assessed by normalizing the peak currents in the test pulse to that of the conditioning pulse.

For slow inactivation recovery, a 5-s conditioning pulse at 0 mV was followed by variable recovery durations at -120 mV, then a 10-ms test pulse at 0 mV. The inter-sweep interval was 15 s.

In experiments assessing the effect of zebrafish PRRT2 on slow inactivation recovery, a modified protocol was used to avoid cumulative inactivation. Within a single sweep, a 5-s depolarization was followed by a series of 10-ms test pulses to 0 mV, each separated by progressively longer recovery periods at -120 mV (Webb et al., 2009).

In this study, we chose 0 mV as the primary conditioning potential because it is widely used in conventional slow inactivation protocols and induces slow inactivation more robustly than more negative conditioning voltages, such as -70 mV.

Steady-state slow inactivation assay for Nav channels

To measure steady-state slow inactivation, 10-s conditioning pulses ranging from -110 to -10 mV (in 10 mV increments) were applied, followed by a 10-ms hyperpolarizing step to -120 mV and subsequently a 5-ms test pulse to 0 mV. The peak test current was normalized to the maximum sodium current recorded across all conditions in the protocol and reported as I/Imax.

Slice preparation

Mice (postnatal day 14-20) were anesthetized with 4% isoflurane. The brain tissues were rapidly extracted and immersed in ice-cold sucrose-based cutting solution containing (in mM): 2.5 KCl, 1.25 NaH2PO4, 26 NaHCO3, 10 MgSO4, 0.5 CaCl2, 10 D-Glucose, 205 sucrose, 1 sodium pyruvate, 1.3 sodium L-ascorbate. Coronal cortical slices (250 μm) were prepared using a vibratome (Leica, VT1200S).

Slices were then transferred immediately to an incubation beaker containing aerated artificial cerebrospinal fluid (ACSF) composed of (in mM): 126 NaCl, 2.5 KCl, 2 MgSO4, 2 CaCl2, 26 NaHCO3, 1.25 NaH2PO4 and 25 dextrose (pH 7.4, 315 mOsm). Slices were incubated at 35 °C for 30 min, then maintained in the same solution at room temperature until use. The oxygenated ACSF was equilibrated with 95% O2 and 5% CO2.

Electrical stimulation and field recording in the corpus callosum

Coronal slices were transferred to a recording chamber. A stimulation electrode (125 µm in diameter, concentric bipolar electrode, FHC, USA) was placed on the corpus callosum. The Ag-AgCl recording electrode (2-3 MΩ) was located at a site 150 µm away from the stimulation electrode along the corpus callosum. Stimulating pulses (0.1 ms pulse width, 100 µA) were delivered at 1 Hz for 100 seconds or at 20 Hz for 5 seconds. Compound action potentials were recorded as a functional readout related to Nav channel availability.

Electrophysiological recording in isolated axonal blebs

Slices were transferred to a recording chamber and stabilized with a platinum anchor grid during recording. Resealed axonal ends (“blebs”) in the cerebral cortex were visualized using an upright infrared differential interference contrast (IR-DIC) microscope. Individual blebs were mechanically isolated by sweeping a cutting pipette beneath it, and subsequently patched with glass electrodes (resistance ∼7 MΩ).

For sodium current recordings, pipettes were filled with a potassium gluconate-based internal solution containing (in mM): 130 K-gluconate, 1 MgCl2, 11 EGTA, 1 CaCl2, 1 KCl, 2 Mg-ATP, 0.3 Na-GTP, 10 HEPES (pH adjusted to 7.2, ∼290 mOsm), supplemented with 2 mM TEA-Cl. The external ACSF solution was supplemented with 5 mM 4-Aminopydine (4-AP) and 200 μM CdCl2 to block K+ and Ca2+ currents, respectively. 1 μM tetrodotoxin (TTX) was used to block TTX-sensitive Nav channels.

Whole-cell voltage-clamp recordings were performed using a MultiClamp amplifier (Molecular Devices, 700B). Currents were filtered at 4 kHz (low-pass) and digitized at 20 kHz via a Digidata interface (Molecular Devices, 1440A). Series resistance was compensated by 80% to minimize voltage errors. Leak currents were subtracted by using an online P/4 procedure. The holding potential was set to -70 mV, approximating the resting membrane potential of cortical neurons. To assess sodium current density, a hyperpolarization prepulse to -110 mV (200 ms) was used to remove inactivation, followed by a 20-ms depolarization to 0 mV. Sodium current density (nA/pF) was calculated by normalizing peak current to bleb capacitance. To evaluate Nav channel slow inactivation, a single 5-s depolarizing pulse (0 mV) was applied to induce slow inactivation followed by test pulses (0 mV, 20 ms) separated by 10-ms recovery intervals at -70 mV. The fraction of available Nav channels was calculated by normalizing the peak current amplitude of each test pulse to the peak current measured at the onset of the 5-s depolarization.

Co-immunoprecipitation in transfected HEK293T cells

In co-immunoprecipitation experiments, one protein-of-interest was tagged with an HA epitope and the other with a Flag epitope. The pCAGIG vector bearing either a HA or Flag tag alone was used as a control in paired constructs. HEK293T cells transfected with a pair of constructs were harvested 24 hours post-transfection for protein expression. Cells were lysed in ice-cold lysis buffer (20 mM Tris-HCl, 150 mM NaCl, 0.5% n-Dodecyl β-D-maltoside (DDM, Anatrace, D310), and protease inhibitor cocktail (Selleckchem, B14002)) with gentle agitation at 4 °C for 2 hours. Lysates were centrifuged at 20,000 x g for 20 minutes at 4 °C to remove insoluble debris. The supernatant was transferred to a fresh tube and kept on ice.

Protein concentration of supernatant was determined using a colorimetric bicinchoninic acid (BCA) assay (Beyotime, P0012S). For each immunoprecipitation (IP), 400 μg of proteins was incubated with 30 μL of anti-HA nanobody magnetic beads (AlfaLifeBio, KTSM1335) or antiFlag magnetic beads (Bimake, B26102) at 4 °C with rotation for 1 hour. Beads were then separated with DynaMag spin and washed three times with wash buffer (20 mM Tris-HCl, 150 mM NaCl, 0.05% DDM). The captured proteins were eluted with 25 μL SDS-PAGE loading buffer containing 100 mM DTT, followed by heating at 45 °C for 10 min. Samples were stored at -20 °C prior to western blot analysis.

Co-immunoprecipitation in mouse brain tissues

Brain tissues (∼100 mg per sample) were freshly dissected from wild-type or Prrt2-V5 knockin mice, and homogenized on ice using glass homogenizers in 1 mL of Tris-buffered saline (TBS, 20 mM Tris-HCl, 150 mM NaCl, protease inhibitor cocktail). The homogenates were centrifuged at 800 x g for 10 min at 4 °C to remove nuclei and cell debris. Supernatants were subjected to centrifugation at 40,000 x g for 45 minutes at 4 °C to isolate membrane fractions. The resulting membrane-enriched pellets were resuspended in TBS and solubilized in lysis buffer containing 0.3% DDM for 3 hours at 4 °C. Lysates were then cleared by centrifugation at 20,000 x g for 20 minutes at 4 °C. Protein concentration was measured via BCA assay. For each IP, ∼200 μg of proteins was incubated with 30 μL of V5-Trap magnetic beads (Chromotek, v5tma-20) at 4 °C with rotation for 1.5 hours. Beads were washed three times with wash buffer (20 mM Tris-HCl, 150 mM NaCl, 0.15% DDM), and bound proteins were eluted with 25 μL SDS-PAGE loading buffer containing 100 mM DTT, followed by heating at 45 °C for 10 min. Eluates were kept at -20 °C until use in western blotting.

Western blotting

Protein samples were separated using 4-12% PAGE Gel (Nanjing ACE Biotechnology, ET15412L) in MES-SDS running buffer at 140 V for 60 minutes. Proteins were transferred onto polyvinylidene difluoride (PVDF) membranes at 300 mA for 2 hours at 4 °C. The PVDF membranes were blocked in 5% nonfat milk in TBST (TBS with 0.1% Tween-20) for 1 hour at room temperature, followed by incubation with indicated primary antibodies overnight at 4 °C with gentle agitation.

After three washes in TBST, membranes were incubated with HRP-conjugated secondary antibodies for 2 hours at room temperature. Following another three washes, chemiluminescent substrate (TIANGEN Biotech, PA112) was applied, and signals were detected using a blot imaging system (Analytikjena, ChemStudio 815). Densitometric analysis of protein bands was performed using QuantityOne software (Biorad, Quantity One 1-D analysis, v4.6.2). The primary antibodies we used in western blotting were as follows: Anti-human PRRT2 (Atlas antibodies, HPA014447), Anti-Nav1.2 (Alomone Labs, ASC-002), Anti-Nav1.1 (Alomone Labs, ASC-001), Anti-HA-Tag (Cell signaling technology, 3724S), Anti-GAPDH-HRP (KangChen Bio-tech,KC-5G5), Anti-ATP1B2 (Abcam, ab185210), Anti-mouse PRRT2 (Wiiget Biotech, Rp3246) and Anti-β Actin-HRP (Cell signaling technology, 5125S).

Animal Surgery

Mice were anesthetized with isoflurane (RWD life science, R510-22-10), using 4% for induction and 2% for maintenance. Body temperature was maintained throughout the procedure with a heating pad. Meloxicam (5 mg/kg, subcutaneous) was administered preoperatively for analgesia.

After securing the mice in a stereotaxic frame (RWD life science, Model 68528), 0.25% bupivacaine was locally injected at the incision sites for perioperative analgesia, and ophthalmic ointment was applied to protect the cornea. A midline incision was made to expose the skull, which was cleaned with 2% hydrogen peroxide using sterile cotton swabs to remove connective tissue. The skull surface was leveled to a stereotaxic plane, and the bregma was identified and marked. Once dry, the exposed skull was coated with a light-curing self-etch adhesive (3M ESPE Single Bond Universal).

For cerebellar electrical stimulation experiments, two small holes separated by 1.5 mm were drilled in the skull above the fourth/fifth cerebellar lobules [anterior-posterior (AP), -6.25 mm; medial-lateral (ML), ±0.75 mm; dorsal-ventral (DV), -1.0 mm from the dura]. Two metal electrodes (100 μm in diameter; California Fine Wire Company, M426550) were implanted and secured with dental cement.

For cortical electrical stimulation and EEG recording, two small craniotomies (∼0.6 mm in diameter) were made over the left sensory cortex (AP -0.8 mm; ML 1.5 mm) and visual cortex (AP -3.5 mm, ML 1.5 mm) for the insertion of bipolar stimulation electrodes (180 μm in diameter), which were implanted to a depth of -0.6 mm from the dura. Stainless steel screws were used as EEG electrodes: one was implanted into the right visual cortex (AP -2.25 mm, ML - 1.5 mm), and the second into the right cerebellar vermis to serve as the reference. An additional screw was placed in the left cerebellar hemisphere as the ground. The electrodes and a custom-made head plate were affixed to the skull using dental cement.

Post-operatively, meloxicam (5 mg/kg, subcutaneously) was administered once daily for 3 consecutive days. Mice were allowed to recover for one week prior to electrical stimulation.

Cerebellar electrical stimulation and behavior assessment

Before cerebellar electrical stimulation, mice were habituated to the head-fixed platform for 30 minutes per day over three consecutive days. Electrical stimulation (300 μA, 60 Hz, 1-ms pulse width, 2 s duration) was delivered to the cerebellar cortex through a stimulus isolator (AMPI, ISO Flex) driven by a Master-8 stimulator. Mouse behavior was recorded with a video camera at a frame rate of 30 frames per second for 10 minutes.

Dystonia in mice were defined by the presence of one or more of the following abnormal postures or movements: prolonged forelimb elevation; persistent backward extension of the ears, tightly apposed to the head; an elongated trunk posture with the abdomen close to the ground and prolonged elevation or excessive extension of hindlimbs.

Electrostimulation and EEG recording in the cerebral cortex

One week after surgery, mice were habituated to a custom-built head plate holder for 30 minutes per day over three consecutive days prior to testing.

For stimulation experiments, the pre-implanted electrodes were connected to a stimulus isolator (AMPI, ISO Flex). Electrical pulses (1 ms pulse width, 60 Hz frequency, 2 s duration) were delivered through the isolator and controlled via TTL signals from a pulse generator (RWD life science, R820). EEG signals (bandpass: 1-300 Hz) were amplified 1000-fold using an AC/DC differential amplifier (A-M Systems, Model 3000), digitized at 1 kHz with a Digidata (Molecular Devices, 1332A), and acquired using Clampex software (Molecular Devices, v10.7).

To assess cortical resilience in Prrt2-mutant versus wild-type mice, a stimulus-response protocol was employed using progressively increasing current intensities (0, 20, 40, 80, 120, 160, 200, 240, 280, 320 µA). Each stimulation was separated by a 24-hour interval. The stimulation threshold was defined as the lowest current intensity that induced after-discharges (lasting ≥ 2 seconds) in the EEG trace. Each EEG recording lasted for 5 minutes, with electrical stimulation delivered immediately following a 1-minute baseline period.

Statistics

All grouped data are presented as mean ± standard error of the mean (s.e.m) unless otherwise stated. Statistical analyses were performed using Graphpad Prism 8 (GraphPad Software Inc, La Jolla, US). No statistical methods were used to pre-determine the sample sizes. The D’Agostino-Pearson omnibus (K2) test was used to assess normality.

Two-tailed, unpaired Student’s t-tests were applied in Figures 7B and 7D. A two-tailed, paired Student’s t-test was used in Figures 6D-F and Figure 7-figure supplement 1B. Two-tailed, unpaired Mann-Whitney test was applied in Figure 7H. Two-way ANOVAs were conducted for comparisons in Figures 1C-D, 1F-G, 2B-C, 2E-F, 3C-D, 3F-G, 4B-E, 4G-J, 7I, Figure 1-figure supplement 1B-C, Figure 3-figure supplement 1C-D, and Figure 7-figure supplement 2B-C. If necessary, the Geisser-Greenhouse correction was applied to adjust for violations of sphericity.

Statistical tests and sample sizes (including number of cells, slices, or animals) are stated in the figure legends. No data were excluded from analyses. Asterisks in figures denote significance levels: *P < 0.05, **P < 0.01, ***P < 0.001, ****P < 0.0001, n.s., not significant.

Co-immunoprecipitation assays were independently repeated at least three times in both in vitro and in vivo tests. All attempts at replicating experiments presented in the manuscript have obtained consistent conclusions.

Electrophysiological recordings were conducted with the experimenter blinded to group identity (transfected condition or animal genotype).

eLife Assessment

This important study identifies PRRT2 as an auxiliary regulator of Nav channel slow inactivation in vitro and in vivo, proposing that PRRT2 facilitates entry into, and delays recovery from, the slow-inactivated state. The revised manuscript has been substantially strengthened, providing compelling evidence that PRRT2 is relevant to normal brain physiology and disease pathophysiology, providing a mechanistic link between PRRT2 mutations and episodic neurological phenotypes. Overall, this study will be of interest to ion channel biophysicists and neurophysiologists, particularly those studying channelopathies.

Reviewer #1 (Public review):

Summary:

The manuscript by Lu and colleagues demonstrate convincingly that PRRT2 interacts with brain voltage-gated sodium channels to enhance slow inactivation in vitro and in vivo. The work is interesting and rigorously conducted. The relevance to normal physiology and disease pathophysiology (e.g., PRRT2-related genetic neurodevelopmental disorders) seems high. Some simple additional experiments could elevate the impact and make the study more complete.

Strengths:

Experiments are conducted rigorously including experimenter blinding and appropriate controls. Data presentation is excellent and logical. The paper is well written for a general scientific audience.

Comments on revised version.

The manuscript by Lu and colleagues has been revised sufficiently to address all my prior concerns.

Experiments are conducted rigorously including experimenter blinding and appropriate controls. Data presentation is excellent and logical. The paper is well written for a general scientific audience.

Reviewer #2 (Public review):

Summary:

As a member of DspB subfamily, PRRT2 is predominantly expressed in CNS and has been associated with various paroxysmal neurological disorders. Previous studies have shown that PRRT2 interacts with Nav and Cav channels, modulating channel properties and neuronal excitability.

In this manuscript, Lu et al. demonstrate that PRRT2 is a potent regulator of Nav channel slow inactivation, promoting the development of Nav slow inactivation and impeding the recovery from slow inactivation. This effect is highly conserved in PRRT2s across species as well as among DspB family members (TRARG1 and TMEM233). The authors further confirmed the interaction between Nav channels and PRRT2 in heterologous expression systems as well as in Prrt2-V5 knock-in mice. Prrt2-mutant mice, which lack PRRT2 expression, require lower stimulation thresholds for evoking after-discharges when compared with WT mice.

Overall, this is a well-executed and methodologically comprehensive study. This work offers valuable insight into the physiological functions of PRRT2 and reveals a potential pathogenic mechanism underlying PRRT2-associated neurological disorders.

The revised manuscript has addressed most of the concerns raised by the reviewers and has been substantially strengthened, although I still have several concerns regarding the discussion section.

Strengths:

(1) Overall, this is a well-executed and methodologically comprehensive study. The electrophysiological data strongly support the conclusion that PRRT2 is a potent regulator of Nav channel slow inactivation. The observation that this regulation is conserved in PRRT2 across species and among DspB family members raises the possibility that altered regulation of Nav channels may also contribute to the pathogenesis of TRARG1- or TMEM233-associated disorders.

(2) Co-immunoprecipitation assay performed using brain tissue from genetically modified Prrt2-V5 knock-in mice provides convincing in vivo evidence for the interaction between PRRT2 and Nav1.2 channels.

(3) Prrt2-V5 KI mice show markedly reduced PRRT2 protein expression and display phenotypes similar to those observed in Prrt2-mutant mice, supporting an important role of PRRT2 in regulating neuronal and network excitability.

Weaknesses:

(1) Nav1.6 is also highly expressed in cortical neurons and is widely regarded as a major contributor to action potential initiation and sustained high-frequency firing. Given that PRRT2 similarly regulates the fast and slow inactivation of Nav1.6 and Nav1.2 channels, the potential contribution of Nav1.6 regulation to neuronal and network excitability should be discussed.

(2) Slow inactivation is generally considered to develop over timescales ranging from hundreds of milliseconds to seconds or longer. Therefore, the statement in Discussion (Page 13, line 381-382) that "slow inactivation develops on a timescale of tens of milliseconds to seconds" may not accurately reflect the conventional kinetic definition of slow inactivation and should be clarified.

(3) Page 14, line 417-430: "question about how Nav channel slow inactivation is regulated in cells that do not express PRRT2".

PRRT2 is unlikely to be the sole regulator of Nav channel slow inactivation. Other molecules and signaling pathways may regulate Nav channel and contribute to neuronal excitability. In addition, neuronal excitability can also be regulated through modulating other Nav properties, such as long-term inactivation or slow recovery from inactivation, as well as through modulating the activity of other ion channels, for example, Kv7.2 and Kv7.3 channels. Therefore, PRRT2-negative cells may utilize alternative mechanisms to fine-tune neuronal excitability. In its current form, this paragraph somewhat overstates the role of PRRT2 and would benefit from a more balanced discussion.

(4) Page 50, Figure 7-figure supplement 2: It would be helpful to include representative traces of the 1st and the last (20th) compound APs in panels B and C.

Reviewer #3 (Public review):

This paper reveals that the neuronal protein PRRT2, previously known for its association with paroxysmal dyskinesia and infantile seizures, modulates the slow inactivation of voltage-gated sodium ion (Nav) channels, a gating process that limits excitability during prolonged activity. Using electrophysiology, molecular biology, and mouse models, the authors show that PRRT2 accelerates entry of Nav channels into the slow-inactivated state and slows their recovery, effectively dampening excessive excitability. The effect seems evolutionarily conserved, requires the C-terminal region of PRRT2, and is recapitulated in cortical neurons, where PRRT2 deficiency leads to hyper-responsiveness and reduced cortical resilience in vivo. These findings extend the functional repertoire of PRRT2, identifying it as a physiological brake on neuronal excitability. The work provides a mechanistic link between PRRT2 mutations and episodic neurological phenotypes.

Comments:

(1) The precise structural interface and the molecular basis of gating modulation remain inferred rather than demonstrated.

(2) The in vivo phenotype reflects a complex circuit outcome and does not isolate slow-inactivation defects per se.

(3) Expression of PRRT2 in muscle or heart is low, so the cross-isoform claims are likely of limited physiological significance.

(4) The mechanistic separation between trafficking of PRRT2 and its gating effects is not clearly resolved.

(5) Additional studies with Nav1.6 should be carried out.

Comments on revised version.

These comments have been addressed in the revised version.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #2 (Public review):

Summary:

As a member of DspB subfamily, PRRT2 is predominantly expressed in CNS and has been associated with various paroxysmal neurological disorders. Previous studies have shown that PRRT2 interacts with Nav and Cav channels, modulating channel properties and neuronal excitability.

In this manuscript, Lu et al. demonstrate that PRRT2 is a potent regulator of Nav channel slow inactivation, promoting the development of Nav slow inactivation and impeding the recovery from slow inactivation. This effect is highly conserved in PRRT2s across species as well as among DspB family members (TRARG1 and TMEM233). The authors further confirmed the interaction between Nav channels and PRRT2 in heterologous expression systems as well as in Prrt2-V5 knock-in mice. Prrt2-mutant mice, which lack PRRT2 expression, require lower stimulation thresholds for evoking after-discharges when compared with WT mice.

Overall, this is a well-executed and methodologically comprehensive study. This work offers valuable insight into the physiological functions of PRRT2 and reveals a potential pathogenic mechanism underlying PRRT2-associated neurological disorders.

The revised manuscript has addressed most of the concerns raised by the reviewers and has been substantially strengthened, although I still have several concerns regarding the discussion section.

Strengths:

(1) Overall, this is a well-executed and methodologically comprehensive study. The electrophysiological data strongly support the conclusion that PRRT2 is a potent regulator of Nav channel slow inactivation. The observation that this regulation is conserved in PRRT2 across species and among DspB family members raises the possibility that altered regulation of Nav channels may also contribute to the pathogenesis of TRARG1- or TMEM233-associated disorders.

(2) Co-immunoprecipitation assay performed using brain tissue from genetically modified Prrt2-V5 knock-in mice provides convincing in vivo evidence for the interaction between PRRT2 and Nav1.2 channels.

(3) Prrt2-V5 KI mice show markedly reduced PRRT2 protein expression and display phenotypes similar to those observed in Prrt2-mutant mice, supporting an important role of PRRT2 in regulating neuronal and network excitability.

We sincerely thank the reviewer for the meticulous evaluation of our revised manuscript and for the constructive and insightful comments.

Weaknesses:

(1) Nav1.6 is also highly expressed in cortical neurons and is widely regarded as a major contributor to action potential initiation and sustained high-frequency firing. Given that PRRT2 similarly regulates the fast and slow inactivation of Nav1.6 and Nav1.2 channels, the potential contribution of Nav1.6 regulation to neuronal and network excitability should be discussed.

We appreciate the reviewer’s suggestion. In the revised manuscript, we have clarified that PRRT2-mediated regulation of Nav1.2, together with its regulation of Nav1.6, may contribute to neuronal and network excitability in the cortex. Please refer to Page 15, Line 439-440.

(2) Slow inactivation is generally considered to develop over timescales ranging from hundreds of milliseconds to seconds or longer. Therefore, the statement in Discussion (Page 13, line 381-382) that "slow inactivation develops on a timescale of tens of milliseconds to seconds" may not accurately reflect the conventional kinetic definition of slow inactivation and should be clarified.

We thank the reviewer for this comment. We have corrected the timescale description in the Discussion accordingly. Please refer to Page 13, Line 382.

(3) Page 14, line 417-430: "question about how Nav channel slow inactivation is regulated in cells that do not express PRRT2".

PRRT2 is unlikely to be the sole regulator of Nav channel slow inactivation. Other molecules and signaling pathways may regulate Nav channel and contribute to neuronal excitability. In addition, neuronal excitability can also be regulated through modulating other Nav properties, such as long-term inactivation or slow recovery from inactivation, as well as through modulating the activity of other ion channels, for example, Kv7.2 and Kv7.3 channels. Therefore, PRRT2-negative cells may utilize alternative mechanisms to fine-tune neuronal excitability. In its current form, this paragraph somewhat overstates the role of PRRT2 and would benefit from a more balanced discussion.

We appreciate the reviewer’s constructive suggestion. We agree that PRRT2 is unlikely to be the sole regulator of neuronal excitability or Nav channel slow inactivation. In the revised Discussion, we have clarified that PRRT2-dependent regulation of Nav channel slow inactivation represents one mechanism among several that fine-tune neuronal excitability. Other mechanisms, including regulation of potassium channels such as Kv7.2/Kv7.3, and other ion channel- or signaling- dependent pathways, may also contribute to excitability control in both PRRT2-positive and PRRT2-negative neurons. We have revised relevant paragraph to provide a more balanced discussion of alternative mechanisms. Please refer to Page 15, Line 430-438.

(4) Page 50, Figure 7-figure supplement 2: It would be helpful to include representative traces of the 1st and the last compound APs in panels B and C.

We appreciate the reviewer’s valuable suggestion. We have now added representative traces of the first and last compound action potentials, corresponding to the 1st and the 100th responses, respectively, to panels B and C of Figure 7-figure supplement 2. Please refer to Page 51, Figure 7-figure supplement 2.

Recommendations for the authors:

Reviewer #2 (Recommendations for the authors):

(1) Page 19, line 546: a sampling rate of 10kHz was used for recording Nav currents. Because the Nav channel isoforms examined in this manuscript (Nav1.1, Nav1.2, Nav1.4, Nav1.5, and Nav1.6) exhibit extremely rapid activation and inactivation kinetics, the temporal resolution provided by a 10 kHz sampling rate may not be optimal for detailed kinetic analysis. Although I am not requesting additional experiments, the authors may wish to choose a higher sampling rate (e.g., 50 kHz) in future Nav channel studies.

We thank the reviewer for this helpful suggestion regarding the sampling rate for Nav current recordings. We will adopt higher sampling rates for rapid kinetic analyses of sodium currents in our future work.

(2) Typo: Page 24, line 712-713: should "a Digidata (Molecular Devices, 1332A)" be "... 1322A"?

We thank the reviewer for pointing out this typo. We have corrected “Digidata 1332A” to “Digidata 1322A” in the relevant Method section of revised manuscript. Please refer to Page 25, Line 722.

Version 3

Abstract

During sustained activity, voltage-gated sodium (Nav) channels enter a slow-inactivated state to limit cellular hyperexcitability. Disruption of this regulatory process has been implicated in skeletal, cardiac and neurological disorders. While the kinetics of this process are well characterized, its endogenous modulators remain unclear. Here, we identify Proline-Rich Transmembrane Protein 2 (PRRT2) as a native regulator of Nav channel slow inactivation. We show that PRRT2 facilitates the entry of Nav channels into the slow-inactivated state and delays their recovery, a regulatory effect conserved from zebrafish to humans. PRRT2 forms molecular complexes with Nav channels both in vitro and in vivo. In the mouse cortex, PRRT2 deficiency impairs the slow inactivation of Nav channels in neuronal axons, leading to reduced cortical resilience in response to hyperexcitable challenges. Together, these findings establish PRRT2 as a physiological modulator of Nav channel slow inactivation and reveal a mechanism that supports cortical resilience to pathological perturbations.

Introduction

Voltage-gated sodium (Nav) channels are critical for the initiation and propagation of electrical signals in excitable cells, contributing to cellular excitability in various tissues(Hodgkin & Huxley, 1952; Ulbricht, 1977). In mammals, including humans, Nav channels consist of nine α-subunit isoforms (Nav1.1-Nav1.9), which are differentially expressed and distributed across different tissues to regulate specific physiological functions(Catterall, Goldin, & Waxman, 2005). Mutations in Nav channels that lead to either loss- or gain-of-function of these channels are associated with a wide spectrum of diseases, including neurological disorders(Es-cayg & Goldin, 2010; Hedrich, Lauxmann, & Lerche, 2019; Meisler, Hill, & Yu, 2021), pain syndromes(Dib-Hajj, Geha, & Waxman, 2017; Waxman, 2013), muscle illnesses(Cannon, 2018; Mantegazza, Cestele, & Catterall, 2021) and cardiac diseases(Remme, 2013). Consequently, the Nav channels are considered promising therapeutic targets for treating these con-ditions(Noreng, Li, & Payandeh, 2021; Wisedchaisri & Gamal El-Din, 2022).

Voltage-dependent conformational changes enable Nav channels to transition between resting, open and inactivated states(Catterall, Wisedchaisri, & Zheng, 2020). Eukaryotic Nav channels undergo two major forms of inactivation (fast and slow inactivation), which are distinguished by their kinetics of onset and recovery(Goldin, 2003). Fast inactivation is triggered within milliseconds by brief depolarization and is rapidly reversed upon hyperpolarization, typically within tens of milliseconds (Armstrong & Bezanilla, 1977; Bezanilla & Armstrong, 1977). In contrast, slow inactivation develops during prolonged depolarizations and the slow inactivated channels require substantially longer periods to recover, ranging from hundreds of milliseconds to minutes(Rudy, 1978). Previous studies have shown that slow inactivation regulates Nav channel availability during sustained depolarization or high-frequency repetitive activity, thereby modulating the cellular excitability(Goldin, 2003; Silva, 2014; Vilin & Ruben, 2001). Impairment of this regulatory process due to mutations in Nav channels has been associated with disorders such as periodic paralysis(Hayward, Sandoval, & Cannon, 1999), cardiac arrhythmias(Richmond, Featherstone, Hartmann, & Ruben, 1998; Vilin, Makita, George, & Ruben, 1999) and epilepsy(Ghovanloo et al., 2023).

Beyond the intrinsic hierarchical gating mechanisms, auxiliary regulators further modulate Nav channel inactivation dynamics(Goldin, 2003). For example, β subunits of sodium channels (SCN1B) regulate inactivation of Nav channels in the heterologous expression system, although their effects vary across different Nav isoforms and cell types(He & Soderlund, 2014; Vilin et al., 1999; Webb, Wu, & Cannon, 2009). Fibroblast growth factor homologous factors (FHFs) interact with the carboxy terminus of Nav channels(Gade, Marx, & Pitt, 2020; Goetz et al., 2009), promoting intermediated or long-term inactivation, with recovery times spanning hundreds of milliseconds(Dover, Solinas, D’Angelo, & Goldfarb, 2010; Venkatesan, Liu, & Goldfarb, 2014). Although these auxiliary proteins have been identified as regulators of fast or intermediate inactivation, native modulators that specifically target slow inactivation of Nav channels remain largely unknown.

Recent studies on proline-rich transmembrane protein 2 (PRRT2) provide new insights into this issue. The PRRT2, a member of the Dispanins B subfamily (DspB)(Sallman Almen, Bringeland, Fredriksson, & Schioth, 2012), has been identified as a causative gene for paroxysmal kinesigenic dyskinesia(W. J. Chen et al., 2011; Lee et al., 2012; J. L. Wang et al., 2011). In vitro studies have shown that PRRT2 regulates Nav channel surface expression and delays recovery of these channels from inactivated states induced by brief depolarization(Fruscione et al., 2018; Lu et al., 2021; Valente et al., 2023). However, these studies did not clearly distinguish between the effects of PRRT2 on fast versus slow inactivation, nor did they provide direct in vivo evidence for its role in regulating Nav channel inactivation. Our original observations that PRRT2 expression increased the accumulation of inactivated Nav channels during repetitive stimulation(Lu et al., 2021) suggest that PRRT2 may be a previously unrecognized auxiliary regulator of Nav channel slow inactivation.

In this study, we combined whole-cell voltage-clamp recordings, co-immunoprecipitation, and in vivo approaches to investigate the role of PRRT2 in regulating Nav channel slow inactivation, and to examine how deficits in PRRT2-mediated modulation of this process affects neuronal function in awake mice.

Results

PRRT2 regulates the slow inactivation of Nav1.2 channels

Fast and slow inactivation of Nav channels proceed through distinct gating pathways(Goldin, 2003). Although previous studies have implicated PRRT2 in the regulation of Nav channel inactivation(Fruscione et al., 2018; Lu et al., 2021), it remains unclear whether PRRT2 primarily influences fast inactivation, slow inactivation or both. To address this, we expressed mouse PRRT2 in HEK293T cells stably expressing Nav1.2 (Figure 1A) and applied classic electrophysiological paradigms to distinguish between the two inactivation processes.

PRRT2 regulates slow inactivation of Nav1.2 channels

(A and B) Representative immunoblots showing PRRT2 protein expression (A) and schematic of whole-cell recording (B) in Nav1.2-stably expressing HEK293T cells transfected with constructs encoding either EGFP or mouse PRRT2 (mPRRT2). MW, molecular weight. (C) Fast-inactivation decay time constants of sodium currents evoked by 20-ms depolarizing steps to the indicated command voltages (EGFP: n = 13 cells, mPRRT2: n = 13 cells). Insert: representative sodium current traces elicited by a brief depolarization (left) and voltage-clamp protocol used to assess fast inactivation (right). The arrow indicates the decay of the sodium current attributable to fast inactivation. (D) Recovery from fast inactivation of Nav1.2 channels (EGFP: n = 10 cells, mPRRT2: n = 7 cells). Insert: Protocol for examining fast inactivation recovery. (E) Schematic diagram of entry into and recovery from slow inactivation of voltage-gated sodium channels. Slow inactivation modulates the availability of the sodium channels. (F) Entry into slow-inactivated state of Nav1.2 channels induced by progressively longer depolarization (EGFP: n = 6 cells, mPRRT2: n = 6 cells). (G) The effects of PRRT2 on the recovery of Nav1.2 channels from slow-inactivated state induced by 5-s depolarization (EGFP: n = 6 cells, PRRT2: n = 6 cells). Data are presented as mean ± s.e.m. Two-tailed, unpaired Student’s t-test was used in C, and two-way ANOVAs were used in D, F, and G. ****P < 0.0001. n.s., not significant.

Fast inactivation of Nav channels is rapidly triggered by brief depolarization and manifests as a current decay following a transient pore opening (Figure 1B and 1C), consistent with previous reports(Misra, Kahlig, & George, 2008; Thompson et al., 2023). To evaluate the kinetics of fast inactivation, we measured the time constant of current decay across a range of test potentials and found that PRRT2 expression did not significantly alter the rate at which Nav channels entered the fast-inactivated state (Figure 1C). Moreover, the majority of Nav1.2 channels recovered from fast inactivation within ∼10 ms at a hyperpolarized potential of -120 mV, and PRRT2 expression had a minimal effect on this recovery (Figure 1D).

In contrast to fast inactivation, Nav channel slow inactivation develops and recovers over much longer timescales (Figure 1E), typically ranging from hundreds of milliseconds to seconds, or even to minutes(Silva, 2014). Because slow-inactivated channels do not recover during brief hyperpolarization, they can be functionally distinguished from fast-inactivated channels using a classic two-pulse protocol. In this approach, a brief hyperpolarization step (e.g., -120 mV, 10 ms) is inserted between a conditioning pulse and a test pulse to recover fast-in-activated channels. Channels that have entered the slow-inactivated state remain unavailable during the test pulse, resulting in a reduction in peak sodium current (Figure 1E)(Feather-stone, Richmond, & Ruben, 1996).

To assess the development of slow inactivation, we applied depolarization pulses of variable duration, followed by a fixed 10-ms hyperpolarization before the test pulse (Figure 1F). We found that a 5-second depolarization at 0 mV was sufficient to drive most Nav1.2 channels into the slow-inactivated state (Figure 1F), consistent with previous studies(Ganguly, Thompson, & George, 2021). PRRT2 significantly promoted the entry of Nav1.2 channels into the slow-inactivated state during sustained depolarization (Figure 1F).

To examine the role of PRRT2 in channel recovery from slow inactivation, we used a protocol in which a fixed 5-second depolarization at 0 mV was followed by varying durations of hyperpolarization at -120 mV to allow channel repriming (Figure 1G). PRRT2 expression markedly slowed the recovery of Nav channels from the slow-inactivated state (Figure 1G).

To further investigate whether the regulation of PRRT2 in slow inactivation is voltage dependent, we performed a steady-state slow inactivation protocol. In this assay, 10-s depolarizing steps from -110 to -10 mV were applied to drive Nav channels into steady state, followed by a fixed hyperpolarization and a test pulse to assess channel availability (Figure 1-figure supplement 1A). We found that PRRT2 promotes the transition of Nav channels into the slow-inacti-vated state across a broad range of membrane potentials. PRRT2 expression induced a pronounced hyperpolarizing shift in the inactivation curve, with a V0.5 of -76.27 mV compared to - 48.05 mV in controls (Figure 1-figure supplement 1B).

In addition to prolonged depolarization, high-frequency repetitive depolarization can also induce Nav channel slow inactivation(Mickus, Jung, & Spruston, 1999). To determine whether PRRT2 regulates slow inactivation under these conditions, we applied a pulse-train protocol with an inter-pulse recovery interval of 10 ms. The use-dependent attenuation of peak sodium current was quantified as the fraction of current remaining across successive depolarizations (Figure 1-figure supplement 1C). Compared with the EGFP control, PRRT2 expression produced a greater decline in peak current during the train, consistent with enhanced slow inactivation during repetitive depolarizations (Figure 1-figure supplement 1C). Together, these findings indicate that PRRT2 is a potent regulator of Nav1.2 channel slow inactivation in vitro.

Functional auxiliary factors in the regulation of Nav channel slow inactivation

Several auxiliary factors of Nav channels have been identified, including the SCN1B and FHF2A (also known as FGF13a), both of which are implicated in modulating Nav channel in-activation(Dover et al., 2010; Isom et al., 1992). In this study, we compared the effects of mouse SCN1B and FGF13a with those of PRRT2 on the development and recovery of Nav channel slow inactivation (Figure 2A).

Functional auxiliary factors in the regulation of Nav channel slow inactivation

(A) Diagram of Nav channel with potent auxiliary factors, encompassing SCN1B, FHF2A (also known as FGF13a) and PRRT2. (B) The effects of auxiliary factors on the entry of Nav1.2 channels into slow inactivation (EGFP: n = 12 cells, mPRRT2: n = 14 cells, SCN1B: n = 9 cells, FGF13a: n = 12 cells). (C) The effects of auxiliary factors on the recovery of Nav1.2 channels from slow-inactivated state induced by 5-s depolarization (EGFP: n = 12 cells, mPRRT2: n = 12 cells, SCN1B: n = 12 cells, FGF13a: n = 12 cells). (D) Schematic representation of the putative topology of mouse PRRT2 and three truncations. Numbers indicate the positions of the amino acid. REL, re-entrant loop. TM, transmembrane domain. N, amino terminus. C, carboxyl terminus. aa, amino acid. (E) Effects of PRRT2 truncations on the entry of Nav1.2 channels into slow inactivation (EGFP: n = 5 cells, mPRRT2(1–266): n = 14 cells, mPRRT2(222–346): n = 12 cells, mPRRT2(256–346): n = 12 cells). (F) Effects of PRRT2 truncations on the recovery of Nav1.2 channels from slow-inactivated state induced by 5-s depolarization (EGFP: n = 5 cells, mPRRT2(1–266): n = 10 cells, mPRRT2(222–346): n = 10 cells, mPRRT2(256–346): n = 12 cells). Data are presented as mean ± s.e.m. The main effect of group was assessed using two-way ANOVAs (B, C, E and F). **P < 0.01, ****P < 0.0001, n.s., not significant.

We found that SCN1B had negligible effects on both the entry into and recovery from the slow-inactivated state of Nav1.2 compared to controls (Figures 2B and C). In contrast, FGF13a markedly accelerated the onset of inactivation (Figure 2B), consistent with previous reports(Dover et al., 2010; Venkatesan et al., 2014). Notably, FGF13a had minimal effect on the recovery of Nav channels from the slow-inactivated state induced by a 5-second depolarization (Figure 2C). These distinct effects of FGF13a on inactivation are consistent with the idea that FGF13a does not primarily modulate classical slow inactivation, but rather promotes a form of rapid-onset and long-term inactivation(Dover et al., 2010; Venkatesan et al., 2014). By comparison, PRRT2 uniquely regulated slow inactivation by both facilitating entry into and delaying recovery from the slow-inactivated state of Nav channels (Figures 2B and C). The rodent Prrt2 gene encodes a 346-amino-acid protein containing a proline-rich domain in the N-terminus, a re-entrant loop, and a transmembrane domain in the C-terminus (Figure 2D). To identify the functional domain(s) responsible for regulating slow inactivation, we generated a series of protein-truncating variants (Figure 2D).

We first examined PRRT2(222–346), which lacks the N-terminal proline-rich domain (Figure 2D). This variant retained activity similar to the full-length PRRT2, enhancing slow inactivation development and delaying recovery (Figures 2E and F). A shorter construct, PRRT2(256–346), which retains the intact C-terminus, also preserved regulatory function in modulating slow inactivation compared to the EGFP control (Figures 2E and F). To further test whether the C-terminal region is necessary for the effect of PRRT2 on Nav channel slow inactivation, we examined PRRT2(1–266), a truncated variant lacking the membrane-associated C-termi-nal domains (Figure 2D). Unlike variants containing the C-terminal region, PRRT2(1–266) showed no detectable effect on either the development or recovery of Nav channel slow inactivation compared to the control (Figures 2E and F). Notably, this lack of effect was not attributable to absent expression of the truncated construct (Figure 2-figure supplement 1A and B). Together, these results indicate that the C-terminal region of PRRT2, including the re-en-trant loop and transmembrane domain, is both necessary and sufficient to mediate its regulatory effect on slow inactivation of Nav1.2 channels in vitro.

Evolutionarily conserved effects of PRRT2 on Nav channel slow inactivation

The PRRT2 and PRRT2-like genes emerged in vertebrates and have been conserved throughout evolution(Sallman Almen et al., 2012). The alignment of amino acid sequences from zebrafish, mouse, and human PRRT2 revealed 77.46% identity between human and mouse PRRT2, and 43.60% identity between zebrafish and mouse PRRT2 (Figures 3A and B). Notably, the majority of conserved residues are located within the C-terminal region of the PRRT2 proteins (Figure 3A). Given the essential role of the C-terminus of mouse PRRT2 in regulating Nav channel slow inactivation, we examined whether PRRT2 proteins from different species exhibit similar effects.

Evolutionarily conserved effects of PRRT2 on Nav channel slow inactivation

(A) Sequence alignment of PRRT2 protein in human, mouse and zebrafish. Conserved amino acids are highlighted and the membrane-associated domains in carboxyl terminus of PRRT2 are underlined. REL, re-entrant loop. TM, transmembrane domain. (B) Sequence identity of PRRT2 protein in human (hPRRT2), mouse (mPRRT2) and zebrafish (zfPRRT2). Identity is calculated by using protein BLAST tool from NCBI. (C) The effects of PRRT2 from different species on the entry of Nav1.2 channels into slow inactivation (EGFP: n = 16 cells, mPRRT2: n = 19 cells, hPRRT2: n = 20 cells, zfPRRT2: n = 11 cells). (D) The effects of PRRT2 from different species on the recovery of Nav1.2 channels from slow-inactivated state induced by 5-s depolarization (EGFP: n = 16 cells, mPRRT2: n = 19 cells, hPRRT2: n = 19 cells, zfPRRT2: n = 11 cells). (E) Diagram for the chimeric construct of PRRT2 originated from human and zebrafish. (F) The effect of chimeric PRRT2 on the entry of Nav1.2 channels into slow inactivation (EGFP: n = 11 cells, hPRRT2: n = 6 cells, zfPRRT2: n = 10 cells, Chimera: n = 11 cells). (G) The effect of chimeric PRRT2 on the recovery of Nav1.2 channels from slow-inactivated state induced by 5-s depolarization (EGFP: n = 11 cells, hPRRT2: n = 6 cells, zfPRRT2: n = 10 cells, Chimera: n = 9 cells). Data are presented as mean ± s.e.m. The main effect of group was assessed using two-way ANOVAs (C, D, F and G). **P < 0.01, ****P < 0.0001, n.s., not significant.

By comparing the effects of zebrafish, mouse, and human PRRT2 on Nav channel slow inactivation, we found that all three PRRT2 orthologs similarly promoted the entry of Nav1.2 channels into the slow-inactivated state (Figure 3C). Notably, in recovery assays, zebrafish PRRT2 was more effective at delaying the recovery process compared to its mammalian counterparts (Figure 3D). To identify the domain responsible for this enhanced effect, we generated a chimeric PRRT2 construct by fusing the C-terminal region of zebrafish PRRT2 to the N-terminal region of human PRRT2 (Figure 3E). The chimeric PRRT2 exhibited comparable effects on the rate of slow inactivation development to those observed with both human and zebrafish PRRT2 (Figure 3F). Notably, the chimeric variant displayed enhanced modulation of recovery, resembling zebrafish PRRT2 and significantly differing from human PRRT2 (Figure 3G). These findings indicate that the C-terminal region of zebrafish PRRT2 confers its greater ability to modulate recovery of Nav1.2 from the slow-inactivated state.

We next extended our analysis to other DspB family members, including Trafficking regulator of GLUT4-1 (trarg1) and Transmembrane protein 233 (tmem233), both of which encode proteins with C-terminal regions similar to that of PRRT2 (Figure 3-figure supplement 1A and B). Given the critical role of the PRRT2 C-terminus in modulating slow inactivation, we hypothesized that TRARG1 and TMEM233 might exert similar regulatory effects. Under heterologous expression conditions, both TRARG1 and TMEM233 facilitated the entry of Nav1.2 channels into the slow-inactivated state and delayed their recovery (Figure 3-figure supplement 1C and D). These effects were comparable to those observed with mouse PRRT2, although TMEM233 exhibited slightly weaker modulation than PRRT2 and TRARG1 (Figure 3-figure supplement 1C and D). Together, these results suggest that the role of PRRT2 in regulating Nav channel slow inactivation is evolutionarily conserved.

PRRT2 promotes slow inactivation across multiple human Nav channel isoforms

In humans, the nine Nav channel isoforms (Nav1.1-1.9) share high sequence homology and similar structural features(Goldin, 2001). To determine whether PRRT2 modulates slow inactivation across different Nav channel isoforms, we selected Nav1.1, Nav1.4, Nav1.5, and Nav1.6, which are expressed in distinct tissues and cell types, and assessed the effects of human PRRT2 on their slow inactivation properties (Figures 4A-J).

PRRT2 promotes slow inactivation across multiple human Nav channel isoforms

(A) Protocol for examining the entry of Nav channels into slow-inactivated state. (B-E) The effects of human PRRT2 (hPRRT2) on the entry of Nav1.1, Nav1.4, Nav1.5 and Nav1.6 channels into slow inactivation. In (B), analyses for Nav1.1 (EGFP: n = 10 cells, hPRRT2: n = 12 cells); In (C), analyses for Nav1.4 (EGFP: n = 11 cells, hPRRT2: n = 11 cells); In (D), analyses for Nav1.5 (EGFP: n = 17 cells, hPRRT2: n = 17 cells); In (E), analyses for Nav1.6 (EGFP: n = 5 cells, hPRRT2: n = 4 cells). (F) Protocol for examining the recovery of Nav channels from slow-inactivated state. (G-J) The effects of human PRRT2 on the recovery of Nav1.1, Nav1.4, Nav1.5 and Nav1.6 channels from slow-inactivated state induced by 5-s depolarization. In (G), analyses for Nav1.1 (EGFP: n = 10 cells, hPRRT2: n = 12 cells); In (H), analyses for Nav1.4 (EGFP: n = 11 cells, hPRRT2: n = 11 cells); In (I), analyses for Nav1.5 (EGFP: n = 16 cells, hPRRT2: n = 15 cells); In (J), analyses for Nav1.6 (EGFP: n = 5 cells, hPRRT2: n = 4 cells) Data are presented as mean ± s.e.m. The main effect of group was assessed using two-way ANOVAs (B-E and G-J). **P < 0.01, ***P < 0.001, ****P < 0.0001.

Nav1.1, Nav1.4, and Nav1.6 displayed similar kinetics in the development and recovery of slow inactivation (Figures 4B, C, E, G, H, and J). In contrast, Nav1.5 showed greater resistance to entering the slow-inactivated state (Figures 4D and I), consistent with previous findings(Richmond et al., 1998). Despite these isoform-specific differences, the heterologous expression of PRRT2 consistently promoted entry of Nav channels into the slow-inactivated state (Figures 4A-E) and delayed recovery from it (Figures 4F-J) across all four Nav isoforms tested. These results suggest that PRRT2 modulates slow inactivation through a common mechanism shared by multiple Nav channel isoforms, despite differences in their intrinsic slow-inactivation properties.

Association between PRRT2 and Nav channels in vitro

PRRT2 may regulate slow inactivation through association with Nav channels. To examine the potential physical association between human PRRT2 and Nav1.2, we co-expressed PRRT2-HA and Flag-Nav1.2 in HEK293T cells and performed co-immunoprecipitation (Co-IP) assays (Figures 5A and B). As controls, cells were transfected with one of the two interacting proteins (either PRRT2-HA or Flag-Nav1.2) together with the empty vectors carrying the corresponding tag of the other protein (Figures 5A and B). Immunoprecipitation using an anti-Flag antibody to capture Flag-Nav1.2 successfully co-precipitated PRRT2 in the co-transfec-tion group, but not in the control groups where either protein was expressed (Figure 5A). Conversely, immunoprecipitation with anti-HA antibody showed that Nav1.2 was specifically co-precipitated with PRRT2-HA in the co-transfection group but not in controls (Figure 5B). These results confirm that PRRT2 and Nav1.2 formed a protein complex in vitro.

Association between PRRT2 and Nav channels in vitro

(A and B) Schematic diagram of the co-immunoprecipitation assay used to detect association between human PRRT2 and Nav1.2. Flag-tagged Nav1.2 (+), HA-tagged PRRT2 (+) and empty (-) vectors were transfected in HEK293T cells as indicated. The cell lysates were immunoprecipitated with anti-Flag (A) or anti-HA (B) magnetic beads, the captured proteins were analyzed by SDS-PAGE and immunoblotting. (C) Diagram for HA-tagged truncation of human PRRT2 (PRRT2(1–268)), in which the carboxyl terminus of PRRT2 was deleted. REL, re-entrant loop. TM, transmembrane domain. (D and E) Co-immunoprecipitation assay for potential association between Nav1.2 and PRRT2(1–268). The proteins immunoprecipitated by anti-Flag (D) or anti-HA (E) magnetic beads were analyzed by immunoblotting. Note that HA-tagged PRRT2(1–268) was detected by anti-HA antibody in (D and E). (F) Diagram for HA-tagged truncation of human PRRT2 (PRRT2(250–340)), in which the amino terminus of PRRT2 was deleted. (G and H) Co-immunoprecipitation assay for potential association between Nav1.2 and PRRT2(250–340). The proteins immunoprecipitated by anti-Flag (G) or anti-HA (H) were analyzed by immunoblotting. Note that HA-tagged PRRT2(250–340) was detected by anti-HA antibody in (G and H). (I and J) Co-immunoprecipitation assay for potential association between Nav1.1 and PRRT2. The proteins captured by anti-Flag (I) or anti-HA (J) magnetic beads were analyzed by immunoblotting. In (A, B, D, E, G, H, I and J), blots shown are representative of at least three independent co-immunoprecipitation experiments with similar results. Red rectangles indicate the immunoblotting of bait and potential prey proteins in co-immunoprecipitation. IB, immunoblot. IP, immunoprecipitation. MW, molecular weight.

To identify the Nav1.2-associating region within human PRRT2, we generated truncated constructs encoding either the N-terminus (residues 1–268) or C-terminus (residues 250–340) of PRRT2. Using the same Co-IP approach, we found that PRRT2(1–268), which contains the intracellular proline-rich domain, failed to associate with Nav1.2 (Figures 5C-E). In contrast, the PRRT2(250–340), which includes the re-entrant loop and transmembrane domain, was able to interact with Nav1.2 (Figures 5F-H). These results indicate that the C-terminal region of PRRT2 is necessary and sufficient, in this assay, for association with Nav1.2 channels. Given that PRRT2 modulates slow inactivation across multiple Nav isoforms, we hypothesized that it might also associate with other Nav isoforms, such as Nav1.1. To test this, we co-expressed PRRT2-HA and Flag-Nav1.1 in HEK293T cells and performed Co-IP assays. Under these conditions, PRRT2 was found to associate with Nav1.1 (Figures 5I and J), similar to the association observed with Nav1.2 (Figures 5A and B).

These findings contrast with previous reports, which suggested that PRRT2 preferentially interacted with Nav1.2 over Nav1.1 in vitro(Franchi et al., 2023; Fruscione et al., 2018; Valente et al., 2023). This discrepancy may stem from differences in the experimental conditions used to assess protein-protein interactions. For example, in our study, Nav1.1 and PRRT2 were co-expressed in HEK293T cells, and cells were solubilized using the detergent n-Dodecyl β-D-maltoside (DDM). Variations in experimental conditions could influence the stability of the PRRT2-Nav channel complex and the efficiency of the Co-IP assay.

Collectively, our findings suggest that PRRT2 associates with Nav channels through its membrane-associated C-terminal domains, forming molecular complexes in vitro.

PRRT2 forms a molecular complex with Nav1.2 in the mouse brain

To extend our findings from the heterologous expression system to an in vivo context, we first investigated whether PRRT2 forms a protein complex with Nav channels in brain tissue. Due to the limited efficacy of available antibodies for immunoprecipitating native PRRT2 or Nav1.2 from brain lysates, we employed a tag-based capture strategy, which relies on the appropriate insertion of a commonly used epitope tag into the target protein.

We generated Prrt2-V5 knock-in mice via CRISPR/Cas9 technology(Yang, Wang, & Jaenisch, 2014), in which a V5 epitope tag was inserted at the C-terminus of PRRT2 (Figures 6A and B). Notably, compared with wild-type mice, PRRT2 protein levels in tissue lysates were markedly reduced in Prrt2-V5 knock-in mice (Figure 6C-D). These mice displayed susceptibility to dystonia attacks, resembling the phenotype previously observed in Prrt2-mutant mice(Wu et al., 2025), but with lower penetrance (Figure 6-figure supplement 1A-C). Despite the reduced PRRT2 expression, Nav1.2 expression remained unchanged (Figure 6E), and co-immunoprecipitation assays remained effective in the Prrt2-V5 knock-in mice. Using anti-V5 nanobody-conjugated magnetic beads, PRRT2 was effectively immunoprecipitated from membrane lysates of brain tissue (Figure 6C). A notable amount of Nav1.2 was co-immunoprecipitated in the Prrt2-V5 knock-in group, whereas only trace levels were detected in wild-type controls (Figures 6C and F). In contrast, ATP1B2, another membrane protein, was not co-immunoprecipitated in the Prrt2-V5 knock-in group (Figures 6C and F), supporting the specificity of the association between PRRT2 and Nav1.2. Together, although the V5 knock-in allele is hypomorphic, these results provide in vivo evidence that PRRT2 forms a molecular complex with Nav1.2 channels in the mouse brain.

PRRT2 forms a molecular complex with Nav1.2 in the mouse brain

(A) Schematic showing the generation of Prrt2-V5 knock-in (V5-KI) mice using Cas9 technology. V5-tag sequence was inserted precisely at the end of Prrt2 gene. PAM, protospacer-ad-jacent motif (NGG). ssDNA, single-strand DNA. (B) Sanger sequencing around sgRNA targeting site in Prrt2-V5 knock-in mouse. The V5-tag, stop codon and homologous arms were denoted. (C) Co-immunoprecipitation assay for potential interaction between Nav1.2 and PRRT2-V5 in the brain tissue. The proteins immunoprecipitated by anti-V5 nanobody were analyzed by immunoblotting. Blots shown are representative of four independent co-immunoprecipitation experiments. IB, immunoblot. IP, immunoprecipitation. MW, molecular weight. WT, wild-type. KI, knock-in. (D-F) Quantification of the density of protein bands for PRRT2 (D) and Nav1.2 (E) in lysate, and for Nav1.2 and ATP1B2 in co-immunoprecipitation experiments (F) (n = 4 mice for each group). Data are presented as mean ± s.e.m. In (D-F), two-tailed, paired Student’s t-test was used for determining statistical significance. **P < 0.01, ****P < 0.0001, n.s., not significant.

PRRT2 deficiency impairs the regulation of Nav channel slow inactivation in cortical neuronal axons

Given the dense distribution of both PRRT2 and Nav channels in the cerebral and cerebellar cortices(Yamano, Miyazaki, & Nukina, 2022), we next investigated whether PRRT2 regulates Nav channel slow inactivation in cortical neurons. Cortical slices were prepared from wild-type and Prrt2-mutant mice, the latter lacking PRRT2 expression(Tan et al., 2018). Unlike the relatively simple morphology of HEK293T cells, pyramidal neurons have complex arborizations, which present space-clamp limitations during whole-cell voltage-clamp recordings(Bar-Ye-huda & Korngreen, 2008). To mitigate these issues, we followed established protocols to isolate axonal blebs from cortical pyramidal neurons(Hu & Shu, 2012). These spherical, membrane-bound blebs form at the cutting surface of cortical slices and are readily visualized (Figure 7A). Sodium currents recorded from axon blebs were voltage-dependent and sensitive to tetrodotoxin (TTX) (Figure 7A), confirming that they were primarily mediated by Nav chan-nels(Hu & Shu, 2012).

PRRT2 deficiency impairs the regulation of Nav channel slow inactivation and neuronal resilience in mice

(A) Schematic showing the isolated axonal bleb recording in cortical neuron (left) and the representative brightfield image for the axonal bleb in brain slice (middle). The arrow indicates an axonal bleb. Scale bar, 20 μm. The representative sodium current traces recorded in axonal blebs (right). Scale bar, 0.5 nA, 5 ms. TTX, tetrodotoxin. (B) Sodium current density recorded in isolated axonal blebs from wild-type (WT) and Prrt2-mutant mice (WT: n = 18 blebs from 7 mice, Prrt2-mutant: n = 20 blebs from 8 mice). Sodium currents were measured with a brief depolarization to 0 mV preceded by a 200-ms hyperpolarizing prepulse to -110 mV. (C) Protocol for assessing Nav channel slow-inactivation (upper) and the representative traces of the sodium currents evoked by conditioning depolarization pulse and test pulses (bottom). Scale bar, 0.5 nA, 5 ms. (D) Fraction of available Nav channels after 5-s depolarization (WT: n = 18 blebs from 7 mice, Prrt2-mutant: n = 19 blebs from 8 mice). (E) Schematic illustration of the cortical electrostimulation and EEG recording in mice. Ref, reference electrode. Gnd, ground electrode. Stim, Stimulation. (F) Illustration showing the experimental schedule and the stimulation parameters. Stimulation was delivered once daily with stepwise increases in current intensity. (G) Representative traces and power of EEG signals in wild-type and Prrt2-mutant mice before and after 2-s cortical stimulation. After-discharges were indicated by the arrow and red line. Scale bar: 0.2 mV. Electrical stimulation (120 µA) was applied during 60–62 s period, where the stimulus-induced artifacts were removed prior to analysis. (H) Threshold of electrical stimulation to induce after-discharges in WT and Prrt2-mutant mice (WT: n = 16 mice, Prrt2-mutant: n = 8 mice). (I) Percentage of after-discharge occurrence in WT and Prrt2-mutant mice after electrical stimulation (WT: n = 16, Prrt2-mutant: n = 8). Data are presented as mean ± s.e.m. Two-tailed, unpaired Student’s t-test was used in (B and D) and two-tailed, unpaired Mann-Whit-ney test was used in (H). In (I), the main effect of group was assessed using two-way ANO-VAs. ****P < 0.0001, n.s., not significant.

PRRT2 deficiency did not affect Nav1.2 protein expression in forebrain tissues (Figure 7-figure supplement 1A and B), nor did it alter the sodium current density in cortical axonal blebs when measured with a brief depolarization to 0 mV preceded by a 200-ms hyperpolarizing prepulse to -110 mV, a protocol designed to relieve inactivation before current elicitation (Figure 7B). Because prolonged voltage-clamp recordings from isolated axonal blebs are technically challenging, we adopted a simplified protocol to assess Nav channel slow inactivation, using a holding potential of -70 mV to approximate the resting membrane potential of cortical neurons. In this protocol, a single 5-s depolarization to 0 mV was applied, and the fraction of available Nav channels was determined after a 10-ms hyperpolarization (Figure 7C).

Approximately 20% of available sodium current remained in wild-type blebs, compared to ∼40% in Prrt2-mutant blebs (Figures 7C and D), indicating that during sustained depolarization, fewer Nav channels in the Prrt2-mutant group entered the slow-inactivated state. Notably, because the proportion of channels entering slow inactivation differed substantially between wild-type and Prrt2-mutant groups, direct comparison of recovery kinetics was not feasible and was therefore not performed. Nevertheless, these findings provide direct evidence that endogenous PRRT2 is required for effective regulation of Nav channel slow inactivation in cortical neurons.

PRRT2 deficiency compromises cortical resilience in awake mice

Excitatory neurons exhibit intrinsic adaptation, whereby their responses progressively decline during sustained or repetitive activation(Fleidervish, Friedman, & Gutnick, 1996; Kim, Owen, Holmes, & Grover, 2012). We therefore hypothesized that PRRT2 may contribute to this process by promoting Nav channel slow inactivation.

To test this idea, we applied a repetitive stimulation protocol in the corpus callosum, a white-matter region of the brain enriched in both PRRT2 and Nav channels. Compound action potentials, recorded with a glass electrode positioned 150 μm from the stimulation site, were used as a functional readout related to Nav channel availability during repetitive activity (Figure 7-figure supplement 2A). Compared with wild-type mice, Prrt2-mutant mice exhibited a smaller decline in compound action potential amplitude during high-frequency stimulation (20 Hz), but not during low-frequency stimulation (1 Hz), indicating reduced adaptation (Figure 7-figure supplement 2B and C). This pattern is consistent with impaired slow inactivation during repetitive activity.

Neuronal adaptation is important for limiting excitability during episodes of aberrant network hyperactivity (Fleidervish et al., 1996; Zang, Marder, & Marom, 2023). To further determine whether PRRT2-dependent impairment of neuronal adaptation contributes to hyperexcitability in vivo, we next assessed cortical resilience in awake mice, a measure of the resistance of the cerebral cortex to pathological perturbations. We used a stimulation-response paradigm in which stimulating electrodes were implanted into the left cerebral cortex, targeting both the sensory and visual cortices, while EEG recording electrodes were placed in the contralateral hemisphere, which receives projections from the stimulated region (Figure 7E). Pulsed electrical stimulation (1-ms pulse width at 60 Hz for 2 s) was delivered to the cortex of mice once daily with stepwise increases in current intensity, and the minimal current required to evoke after-discharges (ADs) in the contralateral cortex was determined in each animal (Figure 7F). The threshold current required to induce after-discharges was used as a functional indicator of cortical resilience to aberrant excitatory inputs(Blume, Jones, & Pathak, 2004).

Prrt2-mutant mice exhibited significantly lower stimulation thresholds for evoking after-dis-charges compared to wild-type controls (Figures 7G and H). Specifically, stimulation at an average intensity of 85 µA reliably induced after-discharges in Prrt2-mutant mice, whereas a significantly higher intensity (190 μA) was required to elicit similar responses in wild-type mice (Figures 7H and I), indicating that PRRT2 contributes importantly to maintaining cortical resilience in response to excitatory perturbations.

Discussion

In this study, we identified PRRT2 as a physiological auxiliary factor of Nav channel slow inactivation. We demonstrate that PRRT2 exerts a dual action by both facilitating the transition of Nav channels into a slow-inactivated state and delaying their recovery. Importantly, this regulatory action is evolutionarily conserved, and requires the integrity of the PRRT2 C-terminal region. Using both Prrt2-V5 knock-in and Prrt2-mutant mouse models, we provide compelling in vivo evidence that PRRT2 forms a molecular complex with Nav channels and modulates their slow inactivation, thereby contributing to neuronal adaptation and cortical resilience to perturbations.

In eukaryotic Nav channels, ion permeability is regulated through at least four distinct and coordinated gating mechanisms: voltage-gated activation and deactivation, fast inactivation, and slow inactivation(Catterall, 2023). Slow inactivation, an evolutionarily conserved intrinsic property of Nav channels, is mediated by conformational rearrangements and has distinct molecular determinants from those governing fast inactivation(Payandeh, 2018; Silva, 2014; Vilin & Ruben, 2001).

The role of PRRT2 in the regulation of Nav channel slow inactivation appears to be unique compared to known regulators such as SCN1B and FHFs. The modulatory effects of the SCN1B on Nav channel inactivation were initially observed in Xenopus oocytes(Wallner, Weigl, Meera, & Lotan, 1993), where it was found to influence sodium currents. Subsequent studies proposed that the SCN1B impedes slow inactivation, a hypothesis later confirmed in mammalian cells co-expressing Nav1.4 and SCN1B(Webb et al., 2009). However, its regulatory effects on slow inactivation varied across Nav isoforms(He & Soderlund, 2014; Makita, Bennett, & George, 1996; Vilin et al., 1999; Xu et al., 2007). In this study, we observed little effect of SCN1B on the onset and recovery of slow inactivation in Nav1.2, consistent with previous findings(Xu et al., 2007). In contrast to SCN1B, PRRT2 exerts a more consistent effect across Nav isoforms, enhancing slow inactivation regardless of the specific Nav subtype. FHF proteins (FHF1-4, also known as FGF11-14) constitute a subfamily of FGFs that lack a secretion signal and function distinctly from traditional FGFs(Olsen et al., 2003). Previous studies demonstrated that FHFs mediate a rapid-onset, long-term inactivation, which differs mechanically from traditional slow inactivation(Dover et al., 2010). Unlike the FHF2A-medi-ated long-term inactivation, which can be triggered by a brief 2-ms depolarization, PRRT2 requires a longer depolarization to produce a detectable effect on the onset of slow inactivation of Nav channels. Additionally, PRRT2 significantly delays recovery of Nav channels from the slow-inactivated state induced by 5-second depolarization, whereas FHF2A has little effect on this process. These findings suggest that PRRT2 and FHF2A modulate Nav channel inactivation through different mechanisms.

Previous studies in heterologous overexpression systems have shown that PRRT2 can influence Nav channel trafficking and surface expression(Fruscione et al., 2018; Valente et al., 2023), raising the possibility that the observed effects on slow inactivation are secondary to altered channel abundance or localization. However, slow inactivation develops on a timescale of hundreds of milliseconds to seconds, whereas detectable changes in Nav channel trafficking and surface abundance generally occur over much longer intervals (minutes to hours) (Freal et al., 2023; Higerd-Rusli et al., 2023). These distinct temporal profiles argue against trafficking as the primary basis for the effects of PRRT2 on Nav channel slow inactivation described here, although direct quantification of dynamic changes in Nav channel surface expression will be required to fully exclude such a contribution(Liu, Wang, Pitt, & Liu, 2022; Tyagi, Higerd-Rusli, Akin, Waxman, & Dib-Hajj, 2025).

Our truncation analysis showed that the C-terminal region of PRRT2, which contains the re-entrant loop and transmembrane domain, is sufficient to reproduce the enhancement of slow inactivation. This finding suggests that PRRT2 modulates slow inactivation through a membrane-embedded or membrane-proximal mechanism, rather than through its intracellular pro-line-rich N-terminus. Current models of Nav channel slow inactivation propose that entry into the slow-inactivated state involves coupled conformational changes in both the voltage-sens-ing domains and the pore region(Catterall, Gamal El-Din, & Wisedchaisri, 2024; Silva, 2014), including contributions from the external pore mouth(Balser et al., 1996; Xiong et al., 2006), the selectivity filter(Chatterjee et al., 2018; Todt, Dudley, Kyle, French, & Fozzard, 1999), the S6 segments(Chancey, Shockett, & O’Reilly, 2007; Y. Chen, Yu, Surmeier, Scheuer, & Catterall, 2006; S. Y. Wang & Wang, 1997), and regions encompassing S4, S4-S5 linker, and S5 (Bendahhou, Cummins, Kula, Fu, & Ptacek, 2002; Cummins & Sigworth, 1996; Hayward et al., 1999; Mitrovic, George, & Horn, 2000; D. W. Wang, Makita, Kitabatake, Balser, & George, 2000). Within this framework, the membrane-associated C-terminal region of PRRT2 may enhance slow inactivation by stabilizing one or both of these conformational rearrangements, although the precise structural basis of this effect remains to be determined.

In mammals, Nav channel isoforms (Nav1.1–1.9) are distributed across the nervous system, skeletal muscle, heart and secretory glands(Cusdin, Clare, & Jackson, 2008). Although we have shown that PRRT2 exerts a consistent effect on slow inactivation across various Nav channel isoforms in vitro, the regulation of Nav channel slow inactivation is likely to be more complex under physiological conditions. PRRT2 is predominantly expressed in the nervous system(W. J. Chen et al., 2011), with enrichment in excitatory neurons(Tan et al., 2018), making it unlikely to directly modulate Nav1.4 and Nav1.5 channels, which are primarily expressed in skeletal muscle and cardiac tissue, respectively(Cusdin et al., 2008). A similar rationale applies to Nav1.1, which is largely confined to inhibitory interneurons(Yu et al., 2006), where PRRT2 expression is minimal or absent. These observations suggest that the functional consequence of PRRT2 depend on the Nav isoform composition and cellular context of each tissue.

The partial mismatch between PRRT2 and Nav channel expression in vivo raises an important question about how Nav channel slow inactivation is regulated in cells that do not express PRRT2. One possibility is the expression of alternative regulators, such as TRARG1 or TMEM233, both of which belong to the DspB family and are expressed in tissues distinct from PRRT2(Ehrlich, Lacey, & Ehrlich, 2020; Oort, Warden, Baumann, Knotts, & Adams, 2007; Santana-Varela et al., 2021; Shibata et al., 2007). These proteins may serve analogous functions in modulating Nav channel slow inactivation. Another possibility is that certain PRRT2-negative cells have a reduced requirement for Nav channel slow inactivation modulation due to their intrinsic rhythmic firing properties. For instance, in the cerebellar cortex, the PRRT2 is absent in Purkinje cells, which exhibit repetitive firing(Lu et al., 2021). In such rhythmic firing cells, strong enhancement of Nav channel slow inactivation may be undesirable, as excessive accumulation of inactivated Nav channels could disrupt their regular firing patterns. Thus, the broad isoform activity of the PRRT2 should be considered in any future attempt to manipulate PRRT2 function therapeutically. Notably, Nav channel slow inactivation represents only one of several mechanisms that regulate neuronal excitability. In both PRRT2-positive and PRRT2-negative neurons, excitability may also be fine-tuned through alternative mechanisms, including regulation of potassium channels such as Kv7.2/Kv7.3 and additional ion channel- or signaling-dependent pathways(Debanne, Inglebert, & Russier, 2019; Jones, Gamper, & Gao, 2021). These distinct mechanisms likely operate in a coordinated manner to maintain neuronal excitability, with PRRT2-dependent regulation of Nav channel slow inactivation contributing serving as one component of this broader homeostatic framework(Marom & Marder, 2023).

Given the critical role of PRRT2 in regulating the slow inactivation of Nav1.2 and Nav1.6 channels, both of which are highly expressed in cortical neurons, PRRT2 deficiency may alter neuronal excitability and neuronal responses to prolonged or repetitive activity, thereby contributing to central nervous system dysfunction. Evidence from cultured neurons, animal models, and human genetics is broadly consistent with this view. Studies using human induced pluripotent stem cells (iPSC)-derived excitatory neurons have shown that homozygous PRRT2 mutation increases sodium currents and neuronal excitability(Fruscione et al., 2018). In mouse models, loss-of-function mutation or knockout of PRRT2 in cerebellar granule cells increases intrinsic excitability and facilitates spreading depolarization in the cerebellar cortex(Binda, Valente, Marte, Baldelli, & Benfenati, 2021; Lu et al., 2021), whereas in the forebrain, PRRT2 deficiency impairs Nav channel slow inactivation in cortical neurons and reduces resistance to pathological perturbation in awake mice. In addition, PRRT2 mutations in humans are associated with infantile convulsions(Lee et al., 2012), paroxysmal dyskinesia(W. J. Chen et al., 2011; J. L. Wang et al., 2011), and hemiplegic migraine(Riant et al., 2022), all disorders linked to abnormal neuronal network excitability.

Although the present study supports a functional link between PRRT2-dependent regulation of Nav channel slow inactivation and cortical resilience, PRRT2 may influence cortical excitability and stability through additional Nav-dependent mechanisms, including reduced current density and shifts in the voltage dependence of channel inactivation(Fruscione et al., 2018; Lu et al., 2021; Valente et al., 2023). Notably, because PRRT2 facilitates Nav channel entry into slow-inactivated states both from closed states and from open states during prolonged depolarization, some of these previously reported effects may partly reflect enhanced slow inactivation and the resulting reduction in Nav channel availability. In addition to these Nav-depend-ent mechanisms, previous studies have shown that PRRT2 also regulates synaptic vesicle cycling(Coleman et al., 2018; Tan et al., 2018; Valente et al., 2016) and presynaptic surface expression of Cav2.1 channels(Ferrante et al., 2021). These effects are also expected to influence neurotransmitter release and, consequently, neuronal and network excitability, and may therefore contribute to cortical resilience.

Together, the identification of PRRT2 as an auxiliary factor regulating Nav channel slow inactivation provides new insight into the fine-tuning of Nav channel function and offers a mechanistic framework for understanding how PRRT2 deficiency may contribute to episodic neurological disorders associated with excitability dysregulation.

Limitations of the study

While our study identifies the functional domain of PRRT2 and provides biological evidence that PRRT2 associates with Nav1.2, the precise binding site on the Nav channel that mediates PRRT2-dependent regulation of slow inactivation remains undefined. Direct structural and biochemical mapping of the PRRT2-Nav interface, including targeted mutagenesis, crosslinking, and structural determination, will be necessary to elucidate the molecular basis of this interaction and its effect on channel gating. Cryo-electron microscopy (cryo-EM) represents a promising approach for resolving the structural details of the Nav-PRRT2 complex, particularly in light of recent advancements in determining high-resolution structures of various Nav channel isoforms and their complexes with auxiliary factors(Jiang et al., 2020; Pan et al., 2019; Pan et al., 2021).

Materials and Methods

Molecular cloning

Mammalian expression constructs were generated using the pCAGIG backbone (Addgene, 11159), which contains an internal ribosome entry site (IRES) followed by EGFP(Matsuda & Cepko, 2004). Coding sequences for full-length or truncated proteins of interest (POIs) were inserted in the pCAGIG vector using a seamless cloning method (Beyotime, D7010). The POIs included human PRRT2 (hPRRT2), hPRRT2(1–268), hPRRT2(250–340), mouse PRRT2 (mPRRT2), mPRRT2(222–346), mPRRT2(256–346), mPRRT2(1–266), zebrafish PRRT2 (zfPRRT2), mouse SCN1B, mouse FHF2A, mouse TRARG1 and mouse TMEM233. A PRRT2 chimeric construct was generated by fusing the N-terminal fragment of human PRRT2 (hPRRT2(1–222)) with the C-terminal region of zebrafish PRRT2 (zPRRT2(110–226)), and cloning the resulting sequence into pCAGIG vector. Plasmids encoding human PRRT2(W. J. Chen et al., 2011) were kindly provided by Dr. Zhi-Ying Wu’s Laboratory (Zhejiang University). Plasmids of pCAG-Flag-Nav1.2(Pan et al., 2019) and pCAG-Flag-Nav1.1(Pan et al., 2021) were provided by Huai-Zong Shen’s Laboratory (Westlake University). All constructs generated in this study were verified by Sanger sequencing (BioSune, Boshang Biology Technology, Shanghai).

Cell culture and heterologous expression

HEK293T cell lines stably expressing human Nav1.2 or Nav1.6 and CHO cell lines stably expressing human Nav1.1, Nav1.4 or Nav1.5 were obtained from the laboratory of ICE Bioscience Inc. Wild-type HEK293T cells (SCSP-502) used for transient transfection were obtained from Cell Bank/Stem Cell Bank of Chinese Academy of Sciences. All cells were cultured in Dulbecco’s modified Eagle medium (DMEM; Gibco, 11995065) supplemented with 10% fetal bovine serum (FBS; Gibco,10099141) and penicillin (50 U/mL)-streptomycin (50 μg/mL) (Gibco,15140163), and maintained at 37 °C in a humidified incubator with 5% CO2. The HEK293T cell line used in this study is of female origin and tested negative for mycoplasma contamination.

For heterologous protein expression in whole-cell voltage-clamp recording, 2 μg of plasmid DNA was transfected into HEK293T or CHO cells using Lipofectamine 2000 (ThermoFisher, 11668-019) according to the manufacturer’s instructions. For co-immunoprecipitation assays, 2–3 μg of total plasmid DNA (encoding two proteins of interest) was used for co-transfection. Four hours after the transfection, the medium was replaced with fresh growth medium, and cells were maintained for an additional 20 hours before subsequent experiments.

Animals

Prrt2-mutant mice, carrying a premature stop codon in exon 2 of the Prrt2 gene(Tan et al., 2018), were used in electrophysiological recording experiments.

Prrt2-V5 knock-in mice, used for co-immunoprecipitation (Co-IP) experiments, were generated in this study via CRISPR/Cas9 technology(Yang et al., 2014). Briefly, single-guide RNA (sgRNA: 5’-TCTCCCACAGTGTATAAGTG-3’) targeting the location nearby the stop codon of Prrt2 gene allele, together with a single-stranded DNA donor template encoding a V5 tag flanked by 53-bp homology arms (ssDNA: 5’-CTTCCCTGTC TGTCTTTCCC TCTCCTCTCC CACAGTGTAT AAGGGCAAGC CCATCCCCAA CCCCCTGCTG GGCCTGGACA GCAC-CTGAGG GGCTCTGCCC TGCATTCCAA GACTTTTCTT CCTGTTGG-3’) and Cas9 mRNA, were microinjected into zygotes from fertilized C57BL/6J mice. The injected zygotes were implanted into pseudopregnant female mice. Offspring were screened for correct insertion of the V5 tag immediately upstream of the Prrt2 stop codon by PCR, and the insertion was subsequently verified by Sanger sequencing. Positive founder (F0) mice were crossed with C57BL/6J mice to establish the Prrt2-V5 knock-in line.

Both Prrt2-mutant and Prrt2-V5 lines were backcrossed onto the C57BL/6J background for more than 10 generations. C57BL/6J mice used for breeding were obtained from the Shanghai Laboratory Animal Center (SLAC), Chinese Academy of Science.

Mice were housed in a 12-h light-dark cycle (light on at 7:00 a.m.) with ad libitum access to food and water. All procedures involving animals were approved by the Animal Care and Use Committee of the Center for Excellence in Brain Science and Intelligence Technology, Chinese Academy of Sciences (Approval number, NA-009-2022). Both male and female mice were used in experiments.

Whole-cell voltage clamp recording for sodium currents in cell lines

Sodium currents were recorded from HEK293T or CHO cell lines stably expressing individual human Nav channel isoforms (ICE Bioscience Inc.). Cells were transfected with plasmids as described in the main text. Twenty-four hours after transfection, cells were trypsinized and re-plated at lower density onto poly-D-lysine-coated coverslips. EGFP-positive cells were selected for recording 24 hours later.

Whole-cell voltage-clamp recordings were conducted at room temperature (23-24 °C). The bath solution contained (in mM): 50 NaCl, 90 choline chloride, 3.5 KCl, 1 MgCl2, 2 CaCl2, 10 D-Glucose, 1.25 NaH2PO4, and 10 HEPES (pH adjusted to 7.4 with NaOH). The internal pipette solution contained (in mM): 50 CsCl, 60 CsF, 10 NaCl, 20 EGTA, 10 HEPES (pH adjusted to 7.2 with CsOH). The osmolarity of the external and internal solutions was adjusted to 300 and 290 mOsmol/kg, respectively.

Pipette electrodes were pulled from borosilicate glass capillary (Sutter Instruments, BF150-86-10) using a horizontal puller (P-97, Sutter Instruments). The pipette resistance was approximately 2 MΩ when filled with internal pipette solution. After whole-cell configuration was achieved, cells were allowed to equilibrate for 5 minutes prior to recording. Series resistance was compensated by 80% to minimize voltage errors. Currents were recorded with an amplifier (HEKA, EPC-10), sampled at 10 kHz, and acquired using Patchmaster (HEKA, v2x92).

Fast inactivation assay for Nav channels

To assess the fast-inactivation kinetics of Nav channels, sodium currents were evoked by 20-ms depolarizing steps to test potentials between -20 and +30 mV in 10-mV increments, from a holding potential of -120 mV. The decay time constant (tau) was obtained by mono-expo-nential fitting of the current trace.

Recovery from fast inactivation was measured using a two-pulse protocol, in which a 10-ms depolarizing pulse to 0 mV (P1) was followed by a variable-duration recovery period at -120 mV and a second 10-ms test pulse at 0 mV (P2). Recovery was quantified as the normalized peak current of P2 relative to P1.

Slow inactivation assay for Nav channels

To evaluate the development of slow inactivation of Nav channels, a series of conditioning prepulses of increasing duration at 0 mV was applied, followed by a brief 10-ms recovery at -120 mV and a 10-ms test pulse at 0 mV. Inter-sweep intervals were set at 15 s. Available fractions of Nav channels were assessed by normalizing the peak currents in the test pulse to that of the conditioning pulse.

For slow inactivation recovery, a 5-s conditioning pulse at 0 mV was followed by variable recovery durations at -120 mV, then a 10-ms test pulse at 0 mV. The inter-sweep interval was 15 s.

In experiments assessing the effect of zebrafish PRRT2 on slow inactivation recovery, a modified protocol was used to avoid cumulative inactivation. Within a single sweep, a 5-s depolarization was followed by a series of 10-ms test pulses to 0 mV, each separated by progressively longer recovery periods at -120 mV(Webb et al., 2009).

In this study, we chose 0 mV as the primary conditioning potential because it is widely used in conventional slow inactivation protocols and induces slow inactivation more robustly than more negative conditioning voltages, such as -70 mV.

Steady-state slow inactivation assay for Nav channels

To measure steady-state slow inactivation, 10-s conditioning pulses ranging from -110 to -10 mV (in 10 mV increments) were applied, followed by a 10-ms hyperpolarizing step to -120 mV and subsequently a 5-ms test pulse to 0 mV. The peak test current was normalized to the maximum sodium current recorded across all conditions in the protocol and reported as I/Imax.

Slice preparation

Mice (postnatal day 14-20) were anesthetized with 4% isoflurane. The brain tissues were rapidly extracted and immersed in ice-cold sucrose-based cutting solution containing (in mM): 2.5 KCl, 1.25 NaH2PO4, 26 NaHCO3, 10 MgSO4, 0.5 CaCl2, 10 D-Glucose, 205 sucrose, 1 sodium pyruvate, 1.3 sodium L-ascorbate. Coronal cortical slices (250 μm) were prepared using a vibratome (Leica, VT1200S).

Slices were then transferred immediately to an incubation beaker containing aerated artificial cerebrospinal fluid (ACSF) composed of (in mM): 126 NaCl, 2.5 KCl, 2 MgSO4, 2 CaCl2, 26 NaHCO3, 1.25 NaH2PO4 and 25 dextrose (pH 7.4, 315 mOsm). Slices were incubated at 35 °C for 30 min, then maintained in the same solution at room temperature until use. The ox-ygenated ACSF was equilibrated with 95% O2 and 5% CO2.

Electrical stimulation and field recording in the corpus callosum

Coronal slices were transferred to a recording chamber. A stimulation electrode (125 µm in diameter, concentric bipolar electrode, FHC, USA) was placed on the corpus callosum. The Ag-AgCl recording electrode (2-3 MΩ) was located at a site 150 µm away from the stimulation electrode along the corpus callosum. Stimulating pulses (0.1 ms pulse width, 100 µA) were delivered at 1 Hz for 100 seconds or at 20 Hz for 5 seconds. Compound action potentials were recorded as a functional readout related to Nav channel availability.

Electrophysiological recording in isolated axonal blebs

Slices were transferred to a recording chamber and stabilized with a platinum anchor grid during recording. Resealed axonal ends (“blebs”) in the cerebral cortex were visualized using an upright infrared differential interference contrast (IR-DIC) microscope. Individual blebs were mechanically isolated by sweeping a cutting pipette beneath it, and subsequently patched with glass electrodes (resistance ∼7 MΩ).

For sodium current recordings, pipettes were filled with a potassium gluconate-based internal solution containing (in mM): 130 K-gluconate, 1 MgCl2, 11 EGTA, 1 CaCl2, 1 KCl, 2 Mg-ATP, 0.3 Na-GTP, 10 HEPES (pH adjusted to 7.2, ∼290 mOsm), supplemented with 2 mM TEA-Cl. The external ACSF solution was supplemented with 5 mM 4-Aminopydine (4-AP) and 200 μM CdCl2 to block K+ and Ca2+ currents, respectively. 1 μM tetrodotoxin (TTX) was used to block TTX-sensitive Nav channels.

Whole-cell voltage-clamp recordings were performed using a MultiClamp amplifier (Molecular Devices, 700B). Currents were filtered at 4 kHz (low-pass) and digitized at 20 kHz via a Digidata interface (Molecular Devices, 1440A). Series resistance was compensated by 80% to minimize voltage errors. Leak currents were subtracted by using an online P/4 procedure. The holding potential was set to -70 mV, approximating the resting membrane potential of cortical neurons. To assess sodium current density, a hyperpolarization prepulse to -110 mV (200 ms) was used to remove inactivation, followed by a 20-ms depolarization to 0 mV. Sodium current density (nA/pF) was calculated by normalizing peak current to bleb capacitance. To evaluate Nav channel slow inactivation, a single 5-s depolarizing pulse (0 mV) was applied to induce slow inactivation followed by test pulses (0 mV, 20 ms) separated by 10-ms recovery intervals at -70 mV. The fraction of available Nav channels was calculated by normalizing the peak current amplitude of each test pulse to the peak current measured at the onset of the 5-s depolarization.

Co-immunoprecipitation in transfected HEK293T cells

In co-immunoprecipitation experiments, one protein-of-interest was tagged with an HA epitope and the other with a Flag epitope. The pCAGIG vector bearing either a HA or Flag tag alone was used as a control in paired constructs. HEK293T cells transfected with a pair of constructs were harvested 24 hours post-transfection for protein expression. Cells were lysed in ice-cold lysis buffer (20 mM Tris-HCl, 150 mM NaCl, 0.5% n-Dodecyl β-D-maltoside (DDM, Anatrace, D310), and protease inhibitor cocktail (Selleckchem, B14002)) with gentle agitation at 4 °C for 2 hours. Lysates were centrifuged at 20,000 x g for 20 minutes at 4 °C to remove insoluble debris. The supernatant was transferred to a fresh tube and kept on ice.

Protein concentration of supernatant was determined using a colorimetric bicinchoninic acid (BCA) assay (Beyotime, P0012S). For each immunoprecipitation (IP), 400 μg of proteins was incubated with 30 μL of anti-HA nanobody magnetic beads (AlfaLifeBio, KTSM1335) or anti-Flag magnetic beads (Bimake, B26102) at 4 °C with rotation for 1 hour. Beads were then separated with DynaMag spin and washed three times with wash buffer (20 mM Tris-HCl, 150 mM NaCl, 0.05% DDM). The captured proteins were eluted with 25 μL SDS-PAGE loading buffer containing 100 mM DTT, followed by heating at 45 °C for 10 min. Samples were stored at -20 °C prior to western blot analysis.

Co-immunoprecipitation in mouse brain tissues

Brain tissues (∼100 mg per sample) were freshly dissected from wild-type or Prrt2-V5 knockin mice, and homogenized on ice using glass homogenizers in 1 mL of Tris-buffered saline (TBS, 20 mM Tris-HCl, 150 mM NaCl, protease inhibitor cocktail). The homogenates were centrifuged at 800 x g for 10 min at 4 °C to remove nuclei and cell debris. Supernatants were subjected to centrifugation at 40,000 x g for 45 minutes at 4 °C to isolate membrane fractions. The resulting membrane-enriched pellets were resuspended in TBS and solubilized in lysis buffer containing 0.3% DDM for 3 hours at 4 °C. Lysates were then cleared by centrifugation at 20,000 x g for 20 minutes at 4 °C. Protein concentration was measured via BCA assay. For each IP, ∼200 μg of proteins was incubated with 30 μL of V5-Trap magnetic beads (Chro-motek, v5tma-20) at 4 °C with rotation for 1.5 hours. Beads were washed three times with wash buffer (20 mM Tris-HCl, 150 mM NaCl, 0.15% DDM), and bound proteins were eluted with 25 μL SDS-PAGE loading buffer containing 100 mM DTT, followed by heating at 45 °C for 10 min. Eluates were kept at -20 °C until use in western blotting.

Western blotting

Protein samples were separated using 4-12% PAGE Gel (Nanjing ACE Biotechnology, ET15412L) in MES-SDS running buffer at 140 V for 60 minutes. Proteins were transferred onto polyvinylidene difluoride (PVDF) membranes at 300 mA for 2 hours at 4 °C. The PVDF membranes were blocked in 5% nonfat milk in TBST (TBS with 0.1% Tween-20) for 1 hour at room temperature, followed by incubation with indicated primary antibodies overnight at 4 °C with gentle agitation.

After three washes in TBST, membranes were incubated with HRP-conjugated secondary antibodies for 2 hours at room temperature. Following another three washes, chemiluminescent substrate (TIANGEN Biotech, PA112) was applied, and signals were detected using a blot imaging system (Analytikjena, ChemStudio 815). Densitometric analysis of protein bands was performed using QuantityOne software (Biorad, Quantity One 1-D analysis, v4.6.2). The primary antibodies we used in western blotting were as follows: Anti-human PRRT2 (Atlas antibodies, HPA014447), Anti-Nav1.2 (Alomone Labs, ASC-002), Anti-Nav1.1 (Alomone Labs, ASC-001), Anti-HA-Tag (Cell signaling technology, 3724S), Anti-GAPDH-HRP (KangChen Bio-tech,KC-5G5), Anti-ATP1B2 (Abcam, ab185210), Anti-mouse PRRT2 (Wiiget Biotech, Rp3246) and Anti-β Actin-HRP (Cell signaling technology, 5125S).

Animal Surgery

Mice were anesthetized with isoflurane (RWD life science, R510-22-10), using 4% for induction and 2% for maintenance. Body temperature was maintained throughout the procedure with a heating pad. Meloxicam (5 mg/kg, subcutaneous) was administered preoperatively for analgesia.

After securing the mice in a stereotaxic frame (RWD life science, Model 68528), 0.25% bupi-vacaine was locally injected at the incision sites for perioperative analgesia, and ophthalmic ointment was applied to protect the cornea. A midline incision was made to expose the skull, which was cleaned with 2% hydrogen peroxide using sterile cotton swabs to remove connective tissue. The skull surface was leveled to a stereotaxic plane, and the bregma was identified and marked. Once dry, the exposed skull was coated with a light-curing self-etch adhesive (3M ESPE Single Bond Universal).

For cerebellar electrical stimulation experiments, two small holes separated by 1.5 mm were drilled in the skull above the fourth/fifth cerebellar lobules [anterior-posterior (AP), -6.25 mm; medial-lateral (ML), ±0.75 mm; dorsal-ventral (DV), -1.0 mm from the dura]. Two metal electrodes (100 μm in diameter; California Fine Wire Company, M426550) were implanted and secured with dental cement.

For cortical electrical stimulation and EEG recording, two small craniotomies (∼0.6 mm in diameter) were made over the left sensory cortex (AP -0.8 mm; ML 1.5 mm) and visual cortex (AP -3.5 mm, ML 1.5 mm) for the insertion of bipolar stimulation electrodes (180 μm in diameter), which were implanted to a depth of -0.6 mm from the dura. Stainless steel screws were used as EEG electrodes: one was implanted into the right visual cortex (AP -2.25 mm, ML -1.5 mm), and the second into the right cerebellar vermis to serve as the reference. An additional screw was placed in the left cerebellar hemisphere as the ground. The electrodes and a custom-made head plate were affixed to the skull using dental cement.

Post-operatively, meloxicam (5 mg/kg, subcutaneously) was administered once daily for 3 consecutive days. Mice were allowed to recover for one week prior to electrical stimulation

Cerebellar electrical stimulation and behavior assessment

Before cerebellar electrical stimulation, mice were habituated to the head-fixed platform for 30 minutes per day over three consecutive days. Electrical stimulation (300 μA, 60 Hz, 1-ms pulse width, 2 s duration) was delivered to the cerebellar cortex through a stimulus isolator (AMPI, ISO Flex) driven by a Master-8 stimulator. Mouse behavior was recorded with a video camera at a frame rate of 30 frames per second for 10 minutes.

Dystonia in mice were defined by the presence of one or more of the following abnormal postures or movements: prolonged forelimb elevation; persistent backward extension of the ears, tightly apposed to the head; an elongated trunk posture with the abdomen close to the ground and prolonged elevation or excessive extension of hindlimbs.

Electrostimulation and EEG recording in the cerebral cortex

One week after surgery, mice were habituated to a custom-built head plate holder for 30 minutes per day over three consecutive days prior to testing.

For stimulation experiments, the pre-implanted electrodes were connected to a stimulus isolator (AMPI, ISO Flex). Electrical pulses (1 ms pulse width, 60 Hz frequency, 2 s duration) were delivered through the isolator and controlled via TTL signals from a pulse generator (RWD life science, R820). EEG signals (bandpass: 1-300 Hz) were amplified 1000-fold using an AC/DC differential amplifier (A-M Systems, Model 3000), digitized at 1 kHz with a Digidata (Molecular Devices, 1322A), and acquired using Clampex software (Molecular Devices, v10.7).

To assess cortical resilience in Prrt2-mutant versus wild-type mice, a stimulus-response protocol was employed using progressively increasing current intensities (0, 20, 40, 80, 120, 160, 200, 240, 280, 320 µA). Each stimulation was separated by a 24-hour interval. The stimulation threshold was defined as the lowest current intensity that induced after-discharges (lasting ≥ 2 seconds) in the EEG trace. Each EEG recording lasted for 5 minutes, with electrical stimulation delivered immediately following a 1-minute baseline period.

Statistics

All grouped data are presented as mean ± standard error of the mean (s.e.m) unless otherwise stated. Statistical analyses were performed using Graphpad Prism 8 (GraphPad Software Inc, La Jolla, US). No statistical methods were used to pre-determine the sample sizes. The D’Agostino-Pearson omnibus (K2) test was used to assess normality.

Two-tailed, unpaired Student’s t-tests were applied in Figures 7B and 7D. A two-tailed, paired Student’s t-test was used in Figures 6D-F and Figure 7-figure supplement 1B. Two-tailed, unpaired Mann-Whitney test was applied in Figure 7H. Two-way ANOVAs were conducted for comparisons in Figures 1C-D, 1F-G, 2B-C, 2E-F, 3C-D, 3F-G, 4B-E, 4G-J, 7I, Figure 1-figure supplement 1B-C, Figure 3-figure supplement 1C-D, and Figure 7-figure supplement 2B-C. If necessary, the Geisser-Greenhouse correction was applied to adjust for violations of sphericity.

Statistical tests and sample sizes (including number of cells, slices, or animals) are stated in the figure legends. No data were excluded from analyses. Asterisks in figures denote significance levels: *P < 0.05, **P < 0.01, ***P < 0.001, ****P < 0.0001, n.s., not significant.

Co-immunoprecipitation assays were independently repeated at least three times in both in vitro and in vivo tests. All attempts at replicating experiments presented in the manuscript have obtained consistent conclusions.

Electrophysiological recordings were conducted with the experimenter blinded to group identity (transfected condition or animal genotype).

Materials availability

All plasmids and mice generated in this study are available from the lead contact with a completed materials transfer agreement.